### 2023-07-25 report/markdown/report.Rmd
Initial commit, report files

---
title: "Prognostic and biologic relevance of collagen biosynthesis pathway in prostate cancer"
subtitle: "Manuscript file"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

author: 
  - Isabel Heidegger:
      institute: [uro]
  - Maria Frantzi:
      institute: [mos]
  - Piotr Tymoszuk:
      institute: [daas]
  - Georgios Fotakis:
      institute: [bioinf]
  - Christian Langer:
      institute: [radio]
  - Friedrich Aigner:
      institute: [radio]
  - Zlatko Trajanoski:
      institute: [bioinf]
  - Harald Mischak:
      institute: [mos]
  - Dominik Wolf:
      institute: [onco]
  - Andreas Pircher: 
      institute: [onco]
      email: andreas.pircher@i-med.ac.at
      correspondence: true

institute:
  - uro: Department of Urology, Medical University of Innsbruck, Innsbruck, Austria
  - daas: Data Analytics As a Service Tirol, Innsbruck, Austria
  - mos: Mosaiques-diagnostics GmbH, Hannover, Germany
  - bioinf: Biocenter, Institute of Bioinformatics, Medical University of Innsbruck, Austria
  - radio: Department of Radiology, Medical University of Innsbruck, Innsbruck, Austria
  - onco: Department of Internal Medicine V, Hematology and Oncology, Medical University of Innsbruck, Innsbruck, Austria

keywords: "prostate cancer, collagen, cancer biology, clustering, machine learning"

output: 
  bookdown::word_document2:   
    reference_docx: ms_template.docx

bibliography: coll_biblio.bib

csl: frontiers-in-immunology.csl

header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

```{r, setup, include = FALSE}

library(bookdown)
library(flextable)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      dev = "png", dpi = 600, 
                      tab.cap.pre = 'Supplementary Table ', tab.cap.sep = " ", 
                      tab.cap.style = 'Table Heading')

set_flextable_defaults(font.family = 'Cambria', font.size = 10)

```

\newpage

# Introbuction

-- To be completed -----

\newpage

# Methods

## Multiparametric resonance tomography

-- To be completed -----

## Transcriptome data analysis

A detailed description of the transcriptome analysis procedures is provided in __Supplementary Methods__.

### Software

Transcriptome data was analyzed with R version 4.2.0. 
Basic data transformation tasks and result visualization were accomplished with the _tidyverse_, _rlang_, _ggplot2_ and _cowplot_ packages [@Wickham2019; @Henry2022; @Wickham2016; @Wilke2019]. 
Additional result visualization tools included the 
_ggvenn_ (Venn plots) [@Yan2021], 
_microViz_ (https://github.com/PiotrTymoszuk/microViz, volcano and Forest plots), 
_ExDA_ (https://github.com/PiotrTymoszuk/ExDA, violin and ribbon plots), 
_survminer_ [@Kassambara2016] and _coxExtensions_ (https://github.com/PiotrTymoszuk/coxExtensions) (Kaplan - Meier plots), 
_factoextra_ [@Kassambara2020] and _clustTools_ (https://github.com/PiotrTymoszuk/clustTools) packages (cluster diagnostic graphs and heat maps).

### Study cohorts and genes of interest

The following cohorts were re-analyzed: 
(1) The Cancer Gene Atlas (TCGA) prostate cancer data set [@Abeshouse2015] (n = `r study_data$tcga$expression %>% filter(tissue == 'tumor') %>% nrow` cancer samples, n = `r study_data$tcga$expression %>% filter(tissue == 'benign') %>% nrow` matched normal prostate tissue samples), 
(2) GSE16560 (n = `r study_data$GSE16560$expression %>% nrow` cancer samples) [@Sboner2010], 
(3) GSE40272 (n = `r study_data$GSE40272$expression %>% filter(tissue == 'tumor') %>% nrow` cancer samples and n = `r study_data$GSE40272$expression %>% filter(tissue == 'benign') %>% nrow` matched normal prostate tissue samples) [@Gulzar2013], 
(4) GSE70768 (n = `r study_data$GSE70768$expression %>% filter(tissue == 'tumor') %>% nrow` cancer samples and n = `r study_data$GSE70768$expression %>% filter(tissue == 'benign') %>% nrow` matched normal prostate tissue samples) [@Ross-Adams2015] and 
(5) GSE70769 (n = `r study_data$GSE70769$expression %>% nrow` cancer samples) [@Ross-Adams2015].

The clinical and normalized level 3 RNAseq data for the TCGA cohort were downloaded with the TCGA Assembler 2 script (https://github.com/compgenome365/TCGA-Assembler-2/blob/master/TCGA-Assembler/). 
The clinical and author-provided normalized microarray expression data for the remaining collectives were fetched from the Gene Expression Omnibus (GEO) with an in-house developed script available within the study analysis pipeline.
The available clinical information included age, blood PSA concentration at diagnosis,
Gleason scoring and pathological tumor, node and metastasis staging, overall (GSE16560) and biochemical relapse-free survival (remaining cohorts) (__Table \@ref(tab:tab-cohorts)__). 
The gene expression values were presented as log~2~(Expression) or log~2~(TranscriptCount + 1). 
Expression data annotation was accomplished with the packages _AnnotationDbi_ [@Pages2022] and _org.Hs.eg.db_ [@Carlson2022].
Non-malignant cell content in tumor samples were estimated with the MCP Counter [@Becht2016] and xCell algorithms [@Aran2017] (package _immunodeconv_) [@Sturm2019].

The collagen pathway genes included the published Collagen Signature [@Kocher2021] and the Reactome Collagen
formation pathway (R-HSA-1474290) and was constrained to `r nrow(globals$genes_interest)` genes detected in all cohorts (__Table \@ref(tab:tab-genes)__).

### Hypothesis testing, correlation and multiple testing correction

Differences in gene expression between the matched prostate cancer and normal prostate tissue samples were investigated with two-tailed paired T test. 
Differences in infiltration levels and Gleason scores between tumor sample clusters were assessed by Kruskal-Wallis test with $\eta^2$ effect size statistic. 
Differences in the collagen score between clusters were analyzed with one-way ANOVA with $\eta^2$ effect size statistic. 
Pair-wise correlation of the collagen pathway genes was assessed by Pearson's correlation test. 
Correlation of the collagen scores and Gleason scores was investigated by Spearman's correlation test. 
If not indicate otherwise, p values within each analysis task were corrected for multiple testing with Benjamini-Hochberg method [@Benjamini1995]. 
Base R functions (_t.test()_, _cor.test()_, _aov()_) and tools provided by the packages _rstatix_ [@Kassambara2021], _microViz_ and _ExDA_ were used for statistical hypothesis testing and correlation analysis.

### Semi-supervised clustering

Clustering of the TCGA cohort tumor samples in respect to normalized (Z-score) collagen pathway gene expression was accomplished by the PAM (partition around medoids) algorithm with the Manhattan distance measure (packages _cluster_, _philentropy_ and _ClustTools_) [@Drost2018; @Schubert2019]. 
This procedure was characterized by a good explanatory performance measured by 'explained' clustering variance and the superior rate of correct cluster assignment in 10-fold cross-validation (package _clustTools_) [@Lange2004; @Leng2014] (__Supplementary Figure S6A__). 
The number of clusters, n = 3, was chosen based on the bend of the within-cluster sum of square curve and the peak of the mean silhouette statistic (__Supplementary Figure S6BC__). 
The clusters were named according to the increasing expression of the collagen subunit genes as: Collagen^low^, Collagen^intermediate^ and Collagen^high^. 
Assignment of tumor samples from the test collectives (GSE16560, GSE40272, GSE70768 and GSE70769) to the collagen clusters was done by the k-NN classifier (k-nearest neighbors, k = 7, package _clustTools_) [@Sahanic2021; @Leng2014]. 
The quality of semi-supervised clustering was assessed by comparing the 'explained' clustering variance measure between the training and test cohorts (__Supplementary Figure S6D__).

### Differential gene expression, GO enrichment and signaling modulation between tumor sample clusters

Genes differentially expressed between the collagen clusters were identified by one-way ANOVA and linear modeling with Benjamini-Hochberg correction for multiple testing (FDR: false discovery rate, package _microViz_) [@Benjamini1995]. 
Transcripts significantly regulated in the Collagen^intermediate^ versus Collagen^low^ clusters (__Supplementary Table S1__) and transcripts significantly regulated in the Collagen^high^ versus Collagen^low^ clusters (__Supplementary Table S2__) were identified by the following criteria: 
(1) significance in ANOVA (pFDR~ANOVA~ < 0.05) and 
(2) significant regulation versus the Collagen^low^ cluster in linear modeling (pFDR($\beta$) < 0.05).

Biological process gene ontology (GO) enrichment in the genes significantly regulated between the collagen clusters was analyzed with the _goana()_ function from the _limma_ package [@Young2010]. 
Modulation of KEGG-listed signaling pathways based on fold-regulation estimates of the significantly regulated genes was done with the _SPIA_ package [@Tarca2009]. 
GO enrichment and signaling modulation p values were corrected for multiple testing with Benjamini-Hochberg method [@Benjamini1995].

Pooled estimates of regulation of gene expression versus the Collagen^low^ cluster in the study cohorts were computed with the inverse variance method (package _meta_) [@Balduzzi2019]. 
Based on pooled regulation estimates, modulation of particular components of signaling pathways was visualized with the _pathview_ package [@Luo2013].

### Prediction of differences in biochemical reaction modulation

Gene - biochemical reaction assignment rules were were retrieved from the Recon2 human metabolism model (BiGG database and R package _BiGGR_) [@King2016; @Gavai2015]. 
Estimates of biochemical reaction fold-regulation were computed based on the estimates of whole-genome gene expression differences between the clusters and their standard errors obtained by linear modeling as described for the differential gene expression analysis. 
Such analysis followed the gene - protein - reaction (GPR principle) as described by Gavai et al. [@Gavai2015]. 
Biochemical reaction regulation estimate errors, 95% confidence intervals and significance were computed with a Monte Carlo simulation with n = 1000 random normal distribution samplings of the gene regulation estimates. 
The analysis was done with the package _BiGGR_ [@Gavai2015] and the development package _biggrExtra_ (https://github.com/PiotrTymoszuk/biggrExtra). 
Pooled estimates of fold-regulation of reaction regulation versus the Collagen^low^ cluster were computed with the inverse variance method implemented by the package _meta_ [@Balduzzi2019] and visualized in Forest plots.

### Collagen score development and survival modeling

The collagen score was developed based on normalized (Z-score) expression of the collagen pathway genes in the TCGA training cohort with the Elastic Net Cox modeling and biochemical relapse-free survival as a response variable (package _glmnet_) [@Friedman2010; @Simon2011; @Zou2005]. 
The shrinkage parameter $\lambda$ was found by 200-repeat 10-fold cross-validation (function _cv.glmnet()_) and corresponded to the lowest model deviance [@Friedman2010]. 
The collagen score was defined as the linear predictor score of the Elastic Net model fitted with the optimal $\lambda$ parameter (function _glmnet()_) and was computed for the training and the test collectives (GSE16560, GSE40272, GSE70768 and GSE70769) with the _predict.glmnet()_ method. 
The non-zero Elastic Net model coefficients fitted in the training cohort were extracted with the _coef.glmnet()_ method and exponentiated to obtain hazard ratios (HR).

Survival modeling for the collagen score was done with univariable Cox regression or by multivariable Cox regression with age and Gleason score confounders (packages _survival_ and _coxExtensions_) [@Therneau2000]. 
Calibration of the univariable collagen score stratified by tertiles (low, intermediate and high) was tested with the D'Agostino - Nam $\chi^2$ statistic (package _coxExtensions_) [@DAgostino2003].

## Proteomics

-- To be completed ------

## Data and code availability

Publicly available data sets were analyzed. 
The transcriptome R analysis pipeline is available from GitHub (https://github.com/PiotrTymoszuk/collagen_pca).

\newpage

# Results

## Prostate cancer is identified as a stiff, bona-fide collagen rich lesion by mpMRI

mpMRI is considered a standard diagnostic procedure prior to prostate biopsy. 
We examined mpMRI data for presence of fibrosis in n = ... patients with suspected prostate malignancy due to elevated PSA levels. 
As shown in __Figure XY__ (or Table?), cancer areas defined as ... harbored significantly more fibrotic regions than adjacent benign tissue. 
This suggests collagen accumulation and/or enhanced collagen turnover as a characteristic of prostate cancer.

## Collagen synthesis genes are upregulated in prostate carcinoma

To validate that at the molecular biology level, we compared mRNA expression levels of `r nrow(globals$genes_interest)` genes related to collagen synthesis, post-translational modification and degradation (__Table \@ref(tab:tab-genes)__) between prostate cancer and benign prostate tissue samples in three publicly available data sets (TCGA, GSE40272 and GSE70768) [@Abeshouse2015; @Gulzar2013; @Ross-Adams2015]. 
Two genes involved in synthesis of proline, the limiting aminoacid for collagen production, _ALDH18A1_ and _PYCR1_, along with two genes involved in isomerization and cross-linking of collagen chains, _P4HB_ and _PPIB_, were expressed at significantly higher levels in the cancer tissue than in normal prostate in all analyzed collectives. 
This may suggest a faster rate of collagen synthesis in prostate cancer as compared with the normal gland. 
In turn, genes coding for the collagen IV subunit (_COL4A2_) and a laminin protein (_LAMB3_) were significantly downregulated in prostate carcinoma in all investigated collectives (__Figure \@ref(fig:fig-norm-tumor)__, __Supplementary Figures S1__ and __S2__).

## Biologically distinct Collagen^low^, Collagen^intermediate^ and Collagen^high^ subsets of prostate cancers

To get an in-depth insight of biological consequences of collagen pathway regulation in prostate carcinoma, we resorted to a transcriptome analysis of five publicly available data sets: TCGA [@Abeshouse2015], GSE16560 [@Sboner2010], GSE40272 [@Gulzar2013], GSE70768 [@Ross-Adams2015] and GSE70769 [@Ross-Adams2015].

Analysis of co-expression of the collagen pathway genes in cancer tissue samples revealed a consistent, positive co-regulation pattern of the collagen I, IV, V and IV subunit genes which was evident in all five collectives. 
Such transcript tended to be regulated together with genes linked to proteolytic modification and secretion of pro-collagens (_PCOLCE_) and collagen lysine oxidation (_LOX_, _LOXL1_, _LOXL2_) (__Supplementary Figures S3 - S5__).

By means of semi-supervised PAM clustering of tumor samples in respect to `r nrow(globals$genes_interest)` collagen pathway genes of interest [@Schubert2019], three subsets of prostate carcinoma could be identified (__Supplementary Figure S6A - C__). 
Of note, this clustering structure developed in the training TCGA cohort was highly reproducible in two test cohorts (GSE70768 and GSE70769) and moderately conserved in the remaining two test collectives (GSE16560 and GSE40272) as measured by the explained clustering variances and observed in visual analysis of the collagen gene expression heat maps (__Supplementary Figures S6D__ and __S7__). 
The most conspicuous differences between the newly identified cancer sample clusters concerned expression of collagen subunit genes such as _COL1A1_, _COL1A2_, _COL3A1_, _COL4A1_ or _COL6A3_ along with the differing levels of transcripts of collagen processing proteins _LOXL1_ and _PCOLCE_. 
Differences in expression of genes involved in proline synthesis and recycling (_ALDH18A1_, _PYCR1_ and _PEPD_) between the clusters were minor. 
Based on these differing levels of collagen subunit and modification protein gene levels, we named the cancer subsets identified by the clustering analysis Collagen^low^, Collagen^intermediate^ and Collagen^high^ clusters. 
Of note, 47 - 71% of cancer samples were classified as Collagen^intermediate^, followed by the Collagen^low^ (19 - 31%) and Collagen^high^ clusters (10 - 22% of specimens) (__Figure \@ref(fig:fig-clusters)__, __Supplementary Figure S7__).

In three collectives (TCGA, GSE16560, GSE70769), the collagen cluster assignment was significantly associated with Gleason score, with the highest values observed in the Collagen^high^ tumors (__Supplementary Figure S8__). 
This may indicate, that increased collagen production may be a hallmark of prostate cancer progression.

Concerning cellular composition of the tumor microenvironment (TME), Collagen^high^ and Collagen^intermediate^ cancers were characterized by an elevated infiltration by cancer-associated fibroblasts (CAF) and endothelial cells predicted by the MCP counter deceonvolution procedure [@Becht2016] in all investigated cohorts (__Figure \@ref(fig:fig-infil)__). 
Levels of lymphocyte and myeloid leukocytes in the TME predicted by the MCP counter algorithm were in turn comparable between the collagen clusters (_Supplementary Figure S9__). 
Importantly, the significantly altered levels of CAF and endothelial cells in the clusters were corroborated in an analysis with another deconvolution algorithm, xCell [@Aran2017], in the majority of study cohorts (CAF: TCGA, GSE16560 and GSE70768, endothelial cells: TCGA, GSE40272, GSE70768 and GSE70769, __Supplementary Figure S10__). 
Collectively, this suggests a strong interplay between collagen accumulation and CAF levels in the TME of prostate cancer along with an increased pace of angiogenesis in collagen-rich malignancies.

Next, we delved into alteration of transcriptome in the collagen clusters. 
Genes significantly regulated between the clusters were identified by multiple testing-corrected one-way ANOVA and linear modeling with the Collagen^low^ clusters serving as a reference level [@Benjamini1995]. 
Among the genes most significantly upregulated in Collagen^high^ or Collagen^intermediate^ tumors, transcripts of collagen subunit (_COL1A2_, _COL3A1_, _COL4A1_, _COL4A_, _COL5A1_), proteins involved in collagen processing (_PCOLCE_, _TIMP2_), extracellular matrix (ECM) proteins (_SPARC_), a cytoskeleton protein (_MFAP4_) and a pro-angiogenic signaling receptor (_PDGFRB_) were identified in at least two cohorts (__Supplementary Figures S11 - S13__, __Supplementary Tables S1 - S2__).
The subsequent GO term enrichment analysis within the sets of genes significantly regulated in the Collagen^intermediate^ and the Collagen^high^ cancers pertained to organ and tissue developmental processes, ECM organization, cell differentiation and motility and vasculature development as leading biological processes significantly modulated in collagen-rich prostate cancer (__Figure \@ref(fig:fig-go)__, __Supplementary Figures S14 - S15__). 
As found with the SPIA algorithm [@Tarca2009], three signaling pathways were predicted to be activated in the Collagen^intermediate^ tumors as compared with the Collagen^low^ prostate cancers: focal adhesion, ECM-receptor interaction and regulation of actin cytoskeleton in all investigated cohorts. 
The predicted magnitude of their activation was even higher in the Collagen^high^ cluster. 
In addition, the Collagen^high^ tumors displayed a consistent and strong (14 - 28 fold) activation of the KEGG's small cell lung cancer signaling pathway as compared with the Collagen^low^ cluster (__Figure \@ref(fig:fig-spia)__). 
At the levels of single components of the focal adhesion, ECM-receptor interaction and actin cytoskeleton pathways, an enhanced integrin and growth factor signaling in the Collagen^intermediate^ and Collagen^high^ tumors is predicted to trigger the PI3K/AKT/PAK cascade resulting in cell proliferation and apoptosis inhibition. 
Furthermore, the ROCK/MLC circuit activated in collagen-rich tumors is expected to stimulate cell motility (__Supplementary Figure S16 - S18__). 
Modulation of the small cell lung cancer pathway genes was in turn predicted to activate the NF-$\kappa$B cascade and the downstream anti-apoptotic, growth and angiogenic signaling (__Supplementary Figure S19__). 
Another signaling pathways predicted to be activated in collagen-rich cancers in at least two study collectives included 'pathways in cancer', 'leukocyte transendothelial migration' and inflammatory 'Staphylococcus aureus infection' (__Supplementary Figures S20 - S22__).

Differences in metabolism between the Collagen clusters concerned primarily energy metabolism, fatty acid synthesis and fatty acid oxidation, katabolism of amininoacid amine group and turnover of glycan moieties of extracellular matrix components. 
Furthermore, Collagen^high^ tumors were predicted to have a lower activity of fatty acid synthesis and transport reactions (__Supplementary Figures S23 - S24__, __Supplementary Tables S3 - S4__).

In more details, the rates of multiple reactions involved in fatty acid oxidation, tri-carboxylic acid cycle (TCA) and oxidative phosphorylation were predicted to be significantly lower in the Collagen^intermediate^ cluster than in the Collagen^low^ subset. 
These differences were even larger in Collagen^high^ cancers as compared with Collagen^low^ tumors (__Supplementary Figures S25 - S27__).

Collectively, as indicated by the results of infiltration, transcriptome, GO enrichment, signaling and biochemical reaction analysis, the Collagen^low^, Collagen^intermediate^ and Collagen^high^ clusters represent biologically distinct sub-entities of prostate cancers. 
The increasing expression of collagen and ECM component genes may be paralleled by malignancy progression, increasing CAF infiltration and may drive tumor cell proliferation, resistance to apoptosis, tumor angiogenesis as well as suppression of oxidative energy metabolism and fatty acid oxidation.

## Collagen gene expression predicts prostate cancer survival

We asked how alterations of the collagen pathway gene expression impact on prostate cancer survival and relapse risk. 
By multi-parameter Elastic Net Cox modeling of biochemical relapse-free survival [@Zou2005; @Simon2011] in the TCGA training cohort employing the collagen pathway genes of interest as an initial explanatory variable set, we established a so called 'Collagen Score'.

Collagen Score included expression values of `r coll_score$coefs %>% nrow` collagen pathway genes. 
Among them, the collagen subunit genes _COL1A1_ and _COL10A1_ and the pro-collagen lysine oxidase _PLOD3_ were found prominently associated with an increased relapse risk. 
In turn, the collagen gene _COL14A1_, collagen-modifying enzyme genes _LOX_ and _PPIB_ as well as _PEPD_ involved in collagen degradation constituted the most important favorable components of the Collagen Score signature (__Figure \@ref(fig:fig-coll-score)A__). 
In the training TCGA cohort and the test cohorts GSE16560 and GSE70769, Collagen Score values correlated significantly with increasing Gleason score (__Supplementary Figure S28A__). 
Additionally, in all investigated collectives, Collagen Score levels were significantly higher in the Collagen^intermediate^ and Collagen^high^ cluster cancers than in the Collagen^low^ cluster. 
The maxima of Collagen Score were consistently found in Collagen^high^ cancers (__Supplementary Figure S28B__). 
These two observations suggest increasing values of the Collagen Score in progressing and collagen-rich prostate tumors.

In univariate survival Cox modeling, Collagen Score was significantly associated with poorer biochemical relapse-free survival in the training TCGA cohort and the test GSE70768 and GSE70768 collectives as indicated by significant hazard ratios and concordance indexes. In those collectives, Collagen score was able to explain 26 - 48% of variance in biochemical relapse-free survival (__Figure \@ref(fig:fig-coll-score)BC__, __Supplementary Figure S29__). 
Furthermore, low, intermediate and high tertile strata of Collagen Score correctly identified individuals with increasing recurrence risk indicative of good calibration of Collagen Score (__Figure \@ref(fig:fig-coll-score-km)__). 
Finally, in the TCGA, GSE70768 and GSE70768 cohorts, Collagen Score retained its prediction performance in multi-parameter Cox regression with age and Gleason score as confounders (__Supplementary Figure S29__). 
In turn, association of Collagen Score with biochemical relapse-free survival in the remaining GSE40272 and GSE16560 cohorts was not significant.

Collectively, the results of survival modeling suggest that a multi-gene Collagen Score being a proxy of enhanced and altered collagen content in prostate cancer may represent an independent marker of biochemical relapse-free survival.

\newpage

# Discussion

-- To be filled in ----

\newpage

# Tables

```{r tab-cohorts, tab.cap = 'Characteristic of the analyzed cohorts.'}

flextable(paper_tbl$cohorts) %>% 
  width(1, width = 3, unit = 'cm') %>% 
  width(2:6, width = 3.3, unit = 'cm') %>% 
  theme_vanilla

```

\newpage

```{r tab-genes, tab.cap = 'Collagen genes of interest and their classification.'}

paper_tbl$genes %>% 
  arrange(`Gene group`) %>% 
  select(`Gene group`, `Gene symbol`, `Entrez ID`) %>% 
  flextable %>% 
  width(1, width = 3.5, unit = 'cm') %>% 
  width(2:3, width = 2.5, unit = 'cm') %>% 
  merge_v(j = 1) %>% 
  theme_vanilla

```

\newpage

# Figures

```{r fig-norm-tumor, fig.width = mm_inch(paper_fig$norm_tumor$w), fig.height = mm_inch(paper_fig$norm_tumor$h), fig.cap = 'Expression of collagen pathway genes in the normal prostate and prostate cancer tissue.'}

paper_fig$norm_tumor$plot

```

__Figure \@ref(fig:fig-norm-tumor). Expression of collagen pathway genes in the normal prostate and prostate cancer tissue.__ 
_Differences in expression of the collagen pathway genes (`r nrow(globals$genes_interest)` transcripts) between the donor-matched samples of the normal prostate and prostate cancer tissue were analyzed by two-tailed paired T test in the TCGA, GSE40272 and GSE70768 cohorts. P values were corrected for multiple testing with Benjamini-Hochberg method._
_(A) Numbers of transcripts significantly up- and down-regulated in the cancer tissue versus normal prostate tissue presented in Venn plots. Transcript significantly regulated in all analyzed cohorts are listed next to the plots._
_(B) Benjamini-Hochberg-corrected significance (pFDR) and log~2~ fold regulation in the cancer tissue versus normal prostate tissue. Significantly up- and downregulated genes are labeled with their symbols. Numbers of tissue donors are indicated under the plots._

\newpage

```{r fig-clusters, fig.width = mm_inch(paper_fig$clusters$w), fig.height = mm_inch(paper_fig$clusters$h), fig.cap = 'Clustering of prostate cancers by collagen pathway gene expression.'}

paper_fig$clusters$plot

```

__Figure \@ref(fig:fig-clusters). Clustering of prostate cancers by collagen pathway gene expression.__ 
_Tumor samples were subjected to semi-supervised clustering in respect to the collagen pathway gene expression with the PAM algorithm with Manhattan distance metric (training: TCGA, test cohorts: GSE16560, GSE40272, GSE70768 and GSE70769). Mean normalized (Z-score) levels of the collagen pathway transcripts in the Collagen^low^, Collagen^intermediate^ and Collagen^high^ clusters are presented as thick lines. The tinted ribbons represent 2$\times$SEM (standard error of the mean) intervals. The collagen pathway genes were grouped according to their biological function as proline synthesis pathway, extracellular matrix (ECM) component and ECM processing protein genes. Numbers of samples in the clusters are indicated under the plots._

\newpage

```{r fig-infil, fig.width = mm_inch(paper_fig$infil$w), fig.height = mm_inch(paper_fig$infil$h), fig.cap = 'Estimates of cancer-associated fibroblast and endothelial cell infiltration in the collagen clusters.'}

paper_fig$infil$plot

```

__Figure \@ref(fig:fig-infil). Estimates of cancer-associated fibroblast and endothelial cell infiltration in the collagen clusters.__ 
_Levels of cancer-associated fibroblasts (CAF) (A) and endothelial cells (B) in the prostate cancer samples were estimated by the MCP counter algorithm. Differences in CAF and endothelial cell content between the Collagen^low^, Collagen^intermediate^ and Collagen^high^ clusters were analyzed with Kruskal-Wallis test with $\eta^2$ effect size statistic. P values were corrected for multiple testing with Benjamini-Hochberg method. Infiltration estimates are presented in violin plots. Points indicate single tumor samples. Red diamonds with whiskers represent medians with interquartile ranges. P and effect size statistic values are displayed in the plot captions. Numbers of samples in the clusters are indicated under the plots._

\newpage

```{r fig-go, fig.width = mm_inch(paper_fig$go$w), fig.height = mm_inch(paper_fig$go$h), fig.cap = 'Common biological process GO terms enriched in Collagen intermediate and Collagen high tumors.'}

paper_fig$go$plot

```

__Figure \@ref(fig:fig-go). Common biological process GO terms enriched in Collagen^intermediate^ and Collagen^high^ tumors.__ 
_Genes differentially regulated between the Collagen^intermediate^ and Collagen^low^ clusters and between the Collagen^high^ and Collagen^low^ clusters were identified by Benjamini-Hochberg-corrected one-way ANOVA and linear modeling. Biological process gene ontology (GO) term enrichment was investigated in the sets of significantly regulated genes with the goana tool._
_GO terms found significantly enriched for the Collagen^intermediate^ (A) and for the Collagen^high^ tumors were identified and their enrichment Benjamini-Hochberg-corrected p values (pFDR) are presented in bubble plots. Point size and color corresponds to the -log~10~pFDR value._

\newpage

```{r fig-spia, fig.width = mm_inch(paper_fig$spia$w), fig.height = mm_inch(paper_fig$spia$h), fig.cap = 'Common regulated signaling pathways in Collagen intermediate and Collagen high tumors.'}

paper_fig$spia$plot

```

__Figure \@ref(fig:fig-spia). Common regulated signaling pathways in Collagen^intermediate^ and Collagen^high^ tumors.__ 
_Genes differentially regulated between the Collagen^intermediate^ and Collagen^low^ clusters and between the Collagen^high^ and Collagen^low^ clusters were identified by Benjamini-Hochberg-corrected one-way ANOVA and linear modeling. Signaling pathway modulation based on fold regulation estimates of the significantly regulated genes was investigated with the SPIA tool._
_Signaling pathways found significantly modulated for the Collagen^intermediate^ (A) and for the Collagen^high^ tumors were identified and their magnitudes of regulation (tA metric, tA > 0 indicates activation) are presented in bubble plots. Point size and color corresponds to the tA value._

\newpage

```{r fig-coll-score, fig.width = mm_inch(paper_fig$coll_score$w), fig.height = mm_inch(paper_fig$coll_score$h), fig.cap = 'Development of the Collagen Score for prediction of prostate cancer survival.'}

paper_fig$coll_score$plot

```

__Figure \@ref(fig:fig-coll-score). Development of the Collagen Score for prediction of prostate cancer survival.__ 
_Collagen Score based on the normalized expression levels of the collagen pathway genes was developed in the TCGA training cohort using the elastic net Cox modeling algorithm with biochemical relapse-free survival (RFS) as a response. The optimal $\lambda$ penalty parameter was found by 200-repeat 10-fold cross-validation._
_(A) Hazard ratio (HR) estimates for non-zero coefficients of the Collagen Score elastic net model in the TCGA are presented in a bubble plot. Point size corresponds to the absolute HR value. The points are labeled with HR values. The optimal $\lambda$ parameter is indicated in the plot caption. Numbers of complete observations and relapses are presented under the plot._
_(B, C) Explanatory performance ($R^2$) (B) and prediction performance (concordance index [C] with 95% confidence intervals) (C) of the Collagen Score in the training TCGA cohort and the GSE16560, GSE40272, GSE70768 and GSE70769 test collectives. Type of the survival response (biochemical relapse-free [RFS] or overall survival [OS]) and numbers of complete observations are indicated in the Y axes of the plots._

\newpage

```{r fig-coll-score-km, fig.width = mm_inch(paper_fig$coll_score_km$w), fig.height = mm_inch(paper_fig$coll_score_km$h), fig.cap = 'Prediction of prostate cancer survival by the Collagen Score.'}

paper_fig$coll_score_km$plot

```

__Figure \@ref(fig:fig-coll-score-km). Prediction of prostate cancer survival by the Collagen Score.__ 
_The Collagen Score values in the training TCGA cohort and the GSE16560, GSE40272, GSE70768 and GSE70769 test collectives were stratified into tertiles (1^st^: low, 2^nd^: intermediate, 3^rd^: high). The observed (solid lines) and Collagen Score-modeled survival in for the Collagen Score strata was compared with the D'Agostino-Nam $\chi^2$ statistic ($\chi^2(DN)$) and presented in Kaplan-Meier plots. Values of concordance indexes (C) with 95% confidence intervals and $\chi^2(DN)$ statistic are presented in the plot captions. Type of the survival response (biochemical relapse-free [RFS] or overall survival [OS]) is indicated in the X axes of the plots. Numbers of complete observations and survival events are shown under the plots._

\newpage

# References

### 2023-07-25 report/markdown/report_supplement.Rmd
Initial commit, report files

---
title: "Prognostic and biologic relevance of collagen biosynthesis pathway in prostate cancer"
subtitle: "Supplementary Material"
author: "Urology study team"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::word_document2:   
    reference_docx: ms_template.docx

bibliography: coll_biblio.bib
csl: frontiers-in-immunology.csl
header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

```{r, setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dev = "png", 
                      dpi = 600, 
                      tab.cap.pre = 'Supplementary Table ', 
                      tab.cap.sep = " ", 
                      tab.cap.style = 'Table Heading')

set_flextable_defaults(font.family = 'Cambria', font.size = 10)

```

\newpage

# Supplementary Methods

## Data analysis software

Bioinformatic and statistical data analysis was done with R version 4.2.3. 
Basic data transformation tasks were accomplished with the _tidyverse_ package bundle [@Wickham2019] and the packages the _rlang_ [@Henry2022] and [_trafo_](https://github.com/PiotrTymoszuk/trafo).
Results were visualized using the _ggplot2_ package [@Wickham2016]. 
Venn plots were generated with _ggvenn_ package [@Yan2021]. 
Volcano and Forest plots were generated using the development package [_microViz_](https://github.com/PiotrTymoszuk/microViz). 
Ribbon and violin graphs were created with the development package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA). 
Kaplan-Meier plots were generated with the _survminer_ package [@Kassambara2016] and the development packages [_kmOptimizer_](https://github.com/PiotrTymoszuk/kmOptimizer) and [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions).
Heat map visualizations of clusters and clustering diagnostic plots were generated with the package _factoextra_ [@Kassambara2020] and the development package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools).

Figures were produced by the _cowplot_ package [@Wilke2019], tables were created with the _flextable_ package [@Gohel2022].
The manuscript and the supplementary material files were written in the _rmarkdown_ environment [@Allaire2022] and rendered as Word or PDF files using the _knitr_ [@Xie2022] and _bookdown_ [@Xie2016] packages together with the _author-info-blocks.lua_ (https://github.com/pandoc/lua-filters/blob/master/author-info-blocks/author-info-blocks.lua) and _scholarly-metadata.lua_ (https://github.com/pandoc/lua-filters/blob/master/scholarly-metadata/scholarly-metadata.lua) LaTEX filters authored by Albert Krewinkel and Jörn Krenzer. 
Figures, tables R expressions and links for the R markdown documents were managed with [_figur_](https://github.com/PiotrTymoszuk/figur).

## Study cohorts and genes of interest

The following cohorts investigated in our previous report [@Eigentler2020] were re-analyzed:

* The Cancer Gene Atlas data set [@Abeshouse2015] (n = `r study_data$tcga$expression %>% filter(tissue == 'tumor') %>% nrow` cancer samples, n = `r study_data$tcga$expression %>% filter(tissue == 'benign') %>% nrow` matched normal prostate tissue samples). This included clinical, overall and biochemical relapse-free survival data and normalized next generation sequencing counts (level 3 data) downloaded with the TCGA Assembler 2 script (https://github.com/compgenome365/TCGA-Assembler-2/blob/master/TCGA-Assembler/).

* GSE16560 data set [@Sboner2010] including normalized microarray transcriptome data, clinical and overall survival information deposited at Gene Expression Omnibus (GEO) (n = `r study_data$GSE16560$expression %>% nrow` cancer samples). The data set was fetched from GEO and imported in R with an in-house developed script.

* GSE40272 data set [@Gulzar2013] including normalized microarray transcriptome data, clinical and biochemical relapse-free survival information deposited at GEO (n = `r study_data$GSE40272$expression %>% filter(tissue == 'tumor') %>% nrow` cancer samples and n = `r study_data$GSE40272$expression %>% filter(tissue == 'benign') %>% nrow` matched normal prostate tissue samples). The data set was fetched from GEO and imported in R with an in-house developed script.

* GSE70768 data set [@Ross-Adams2015] including normalized microarray transcriptome data, clinical and biochemical relapse-free survival information deposited at GEO (n = `r study_data$GSE70768$expression %>% filter(tissue == 'tumor') %>% nrow` cancer samples and n = `r study_data$GSE70768$expression %>% filter(tissue == 'benign') %>% nrow` matched normal prostate tissue samples). The data set was fetched from GEO and imported in R with an in-house developed script.

* GSE70769 data set [@Ross-Adams2015] including normalized microarray transcriptome data, clinical and biochemical relapse-free survival information deposited at GEO (n = `r study_data$GSE70769$expression %>% nrow` cancer samples). The data set was fetched from GEO and imported in R with an in-house developed script.

The clinical information provided by the study authors included age, blood PSA concentration at diagnosis,
Gleason scoring and pathological tumor, node and metastasis staging. 
A detailed characteristic of the study cohorts is provided in __Table 1__ and by Eigentler et al. [@Eigentler2020].
Normalized, study author-provided gene expression values were transformed with the $log_2(Expression)$ and $log_2(Expression + 1)$ functions for the microarray and next generation sequencing studies, respectively. 
Gene symbol: Entrez ID conversion was accomplished with the packages _AnnotationDbi_ [@Pages2022] and _org.Hs.eg.db_ [@Carlson2022].

The list of collagen pathway genes, i.e. genes linked to collagen synthesis, post-transcriptional modification
and degradation was based on the published Collagen Signature by Kocher et al. [@Kocher2021] and the Reactome ‘Collagen
formation’ pathway ([R-HSA-1474290](https://reactome.org/content/detail/R-HSA-1474290)). 
The gene list was constrained to the set of `r nrow(globals$genes_interest)` genes detected in all studied cohorts. 
The collagen pathway genes are listed in __Table 2__.

## Non-malignant cell content in the tumor samples

Levels of non-malignant cells in the tumor samples were assessed with the MCP counter (T cells, CD8^+^ T cells, NK cells, B cells, monocytes, macrophages, myeloid dendritic cells, neutrophils, endothelial cells and cancer-associated fibroblasts) [@Becht2016] and the xCell algorithms (CD4^+^ non-regulatory T cells, CD8^+^ T cells, endothelial cells, cancer-associated fibroblasts, M1 macrophages, M2 macrophages, monocytes, neutrophils, NK cells, NKT cells, gamma-delta T cells, CD4^+^ regulatory T cells) [@Aran2017]. 
The infiltration estimates were computed using the _immunodeconv_ package [@Sturm2019].

## Gene expression in the normal prostate and prostate cancer

Expression of the collagen pathway genes was compared between the matched prostate cancer and normal prostate tissue samples in the TCGA, GSE40272 and GSE70768 cohorts by two-tailed paired T test corrected for multiple testing with Benjamini-Hochberg method [@Benjamini1995] (package [_microViz_](https://github.com/PiotrTymoszuk/microViz)).

## Correlation analysis

Pairwise correlation of the collagen pathway gene expression levels in the tumor tissue was investigated by the Pearson's correlation test. 
Correlation of the collagen score values with the Gleason scores was analyzed by the Spearman's correlation test (package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)). 
The p values were corrected for multiple testing with Benjamini-Hochberg method [@Benjamini1995].

## Semi-supervised clustering

Prostate cancer tissue samples were clustered in respect to normalized (Z-score) expression levels of the collagen pathway genes of interest (__Table 2__). 
Inter-cohort variability (batch effects) between the expression values was reduced by with the `ComBat()` algorithm (package _sva_) [@Leek2012]. 
Clustering of tumor samples the training TCGA cohort was accomplished by the PAM (partition around medoids) algorithm with the Manhattan distance measure (packages _cluster_ and _philentropy_, development package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) [@Drost2018; @Schubert2019]. 
This procedure was characterized by a good explanatory performance measured by the 'explained' clustering variance statistic (ratio of the between-cluster sum of squares to the total sum of squares) and the superior stability measured as the rate of correct cluster assignment in 10-fold cross-validation (package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) [@Lange2004; @Leng2014] in a comparison of several clustering algorithms and distance measures (__Supplementary Figure S\@ref(fig:fig-clust-dev)A__).
The number of clusters, n = 3, was chosen based on the bend of the within-cluster sum of square curve and the peak of the mean silhouette statistic (__Supplementary Figure S\@ref(fig:fig-clust-dev)B__). 
The clusters were further referred to as 'collagen clusters' and were named according to the increasing expression of the collagen subunit genes between Collagen^low^, Collagen^intermediate^ and Collagen^high^ clusters. 
Assignment of tumor samples from the test collectives (GSE16560, GSE40272, GSE70768 and GSE70769) to the collagen clusters defined in the training TCGA cohort was done by the k-NN classifier (k-nearest neighbors, inverse distance weighted, 
GSE16560: k = `r coll_clust$semi_clust$GSE16560$kNN`, GSE40272: k = `r coll_clust$semi_clust$GSE40272$kNN`, GSE70768: k = `r coll_clust$semi_clust$GSE70768$kNN`, GSE70769: k = `r coll_clust$semi_clust$GSE70769$kNN`; package _clustTools_) [@Leng2014]. 
The quality of semi-supervised clustering was assessed by comparing the distribution of the collagen clusters and the 'explained' clustering variance measure described above between the training and the test cohorts (__Supplementary Figure S\@ref(fig:fig-clust-dev)C__).

## Characteristic of the collagen clusters

Since levels of some infiltration estimates were strongly non-normally distributed, the infiltration estimates between the clusters were compared by Kruskal-Wallis test with $\eta^2$ effect size statistic. 
Gleason score and the collagen score distribution in the collagen clusters was investigated by Kruskal-Wallis test and one-way ANOVA, respectively, with $\eta^2$ effect size statistic. 
The analyses were done with the _rstatix_ package [@Kassambara2021] and the development package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)). 
P values were corrected for multiple testing with Benjamini-Hochberg method [@Benjamini1995].

## Differential gene expression

Genes differentially expressed between the collagen clusters were identified by one-way ANOVA and linear modeling with Benjamini-Hochberg correction for multiple testing (FDR: false discovery rate, package [_microViz_](https://github.com/PiotrTymoszuk/microViz)) [@Benjamini1995]. 
In brief, the transcripts significantly regulated in the Collagen^intermediate^ versus Collagen^low^ clusters (__Supplementary Table S\@ref(tab:tab-dge-int)__) and the transcripts significantly regulated in the Collagen^high^ versus Collagen^low^ clusters (__Supplementary Table S\@ref(tab:tab-dge-hi)__) were identified by the following criteria: 
(1) significance in ANOVA (pFDR~ANOVA~ < 0.05) and 
(2) significant regulation versus the Collagen^low^ cluster in linear modeling (pFDR($\beta$) < 0.05).

## GO enrichment analysis

Biological process gene ontology (GO) term enrichment in the gene set significantly regulated in the Collagen^intermediate^ versus Collagen^low^ clusters and in the gene set significantly regulated in the Collagen^high^ versus Collagen^low^ clusters was analyzed with the _goana()_ function from the _limma_ package [@Young2010]. 
Enrichment p values were corrected for multiple testing with Benjamini-Hochberg method [@Benjamini1995]. 
The magnitude of enrichment over genomic frequency for the particular $i$-th GO term was expressed as odds ratio (OR) computed with the following formula:

$$OR_i = \frac{n_{DGE, i} \times n_{total}}{n_i \times n_{DGE}}$$

where $n_{DGE, i}$ denotes the number of differentially expressed genes assigned to the $i$-th GO term, 
$n_{total}$ denotes the total number of investigated genes, 
$n_i$ denotes the total number of genes assigned to the $i$-th GO term, 
and $n_{DGE}$ represents the total number of differentially regulated genes.

Common significantly enriched GO terms were identified as GO terms significantly enriched in at least four cohorts, with at lest weak effect size defined as OR $\geq$ 1.44 [@Cohen2013].

## Signaling pathway modulation and visualization

Modulation of KEGG-listed signaling pathways based on fold-regulation estimates of the genes significantly regulated in the Collagen^intermediate^ versus Collagen^low^ clusters and in the gene significantly regulated in the Collagen^high^ versus Collagen^low^ clusters was done with the _SPIA_ package [@Tarca2009]. 
Significantly activated or inhibited pathways were identified by the Benjamini-Hochberg-corrected combined enrichment and perturbation significance metric pGFdr < 0.05. 
The magnitude of pathway modulation versus the Collagen^low^ clusters was measured with the tA statistic [@Tarca2009].

Pooled estimates of fold-regulation of gene expression versus the Collagen^low^ cluster were computed with the inverse variance method implemented by the package _meta_ [@Balduzzi2019]. 
Based on such pooled regulation estimates, modulation of particular components of the signaling pathways related to cancer biology modulated in at least four study cohorts (focal adhesion: KEGG ID hsa04510, ECM-receptor interaction: hsa04512, regulation of actin cytoskeleton: hsa04810 and small cell lung cancer: hsa05222, pathways in cancer: hsa05200) was visualized with the _pathview_ package [@Luo2013].

## Biochemical reactions

Rules of assignment of genes to biochemical reactions were retrieved from the Recon2 human metabolism model available via the BiGG database [@King2016] and the R package _BiGGR_ [@Gavai2015]. 
Estimates of differential expression for all available genes between the Collagen clusters were calculated by linear modeling as described above. 
Estimates of biochemical pathway fold-regulation were computed by evaluation of the gene assignment rules. 
The 'gene~A~ OR gene~B~' operator was interpreted as arithmetic mean of gene A and gene B estimates of differential expression. 
The 'gene~A~ AND gene~B~' operator was interpreted as minimum of the gene A and gene B estimates of differential expression [@Gavai2015]. 
Standard deviation, 95% confidence intervals and p values for the predicted reaction regulation estimates were obtained by a Monte Carlo simulation with n = 1000 draws from normal distribution of gene expression regulation estimates (mean: model estimate, standard deviation: standard error of the model estimate) [@Gavai2015; @Pichler2023]. 
The analysis was done with the package _BiGGR_ [@Gavai2015] and the development package [_biggrExtra_](https://github.com/PiotrTymoszuk/biggrExtra) [@Pichler2023]. 
The results of biochemical reaction regulation prediction for reactions found significantly regulated in all study cohorts are listed in __Supplementary Table S\@ref(tab:tab-react-int)__ and __Supplementary Table S\@ref(tab:tab-react-hi)__.

Enrichment of significantly activated or inhibited biochemical reactions within the Recon model metabolic subsystems [@King2016] was investigated by Benjamini-Hochberg-corrected Fisher's exact test (packages [_biggrExtra_](https://github.com/PiotrTymoszuk/biggrExtra) and _rstatix_) [@Kassambara2021]. 
The magnitude of enrichment was assessed with the odds ratio statistic (OR) defined for the $i$-th metabolic subsystem as follows:

$$OR_i = \frac{n_{regulated, i} \times n_{total}}{n_i \times n_{regulated}}$$

where $n_{regulated, i}$ denotes the number of significantly regulated (activated or inhibited) biochemical reaction within the $i$-th metabolic subsystem, 
$n_{total}$ denotes the total number of investigated reactions, 
$n_i$ denotes the number of metabolic reactions within the $i$-th metabolic subsystem, and 
$n_{regulated}$ is the total number of significantly regulated biochemical reactions.

Common significantly activated or inhibited metabolic subsystems were identified as metabolic subsystems significantly activated or inhibited in at least four cohorts, with at least weak effect size defined as OR $\geq$ 1.44 [@Cohen2013].

## Development of the collagen score

The collagen score was developed based on normalized (Z-score) expression of the collagen pathway genes in the TCGA training cohort by the Elastic Net Cox modeling with biochemical relapse-free survival as a response variable (package _glmnet_) [@Friedman2010; @Simon2011; @Zou2005]. 
The optimal value of the shrinkage parameter $\lambda$ = `r signif(coll_score$opt_lambda$lambda, 2)` was found by 200-repeat 10-fold cross-validation (CV, function `cv.glmnet()`, folds generated by the _caret_ package) and corresponded to the lowest model deviance [@Kuhn2008; @Friedman2010]. 
The collagen score was defined as the linear predictor score of the Elastic Net model fitted with the optimal $\lambda$ parameter (function `glmnet()`).  
The collagen score was calculated for the training cohort and the test collectives (GSE16560, GSE40272, GSE70768 and GSE70769) with the `predict.glmnet()` method. 
The non-zero Elastic Net model coefficients fitted in the training cohort were extracted with the `coef.glmnet()` method and exponentiated to obtain hazard ratios (HR).

## Univariable and multivariable survival modeling

Association of the collagen score values with overall (GSE16560) or biochemical relapse-free survival (remaining cohorts) was investigated by univariable Cox modeling or by multivariable Cox modeling with age and Gleason score confounders (package _survival_, development package [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions)) [@Therneau2000]. 
Proportional hazard assumption was tested with the `cox.zph()` function from the package _survival_ [@Grambsch1994].

Harrell's concordance index (C-index) and Grafe's integrated Brier score were used as performance metrics of the Cox models [@Harrell1996; @Graf1999] and employed in comparison of the Cox models with age and Gleason score with the Cox models including the Collagen score and clinical confounders as explanatory variables. 
Calibration of the univariable collagen score survival models was tested with a variation of the Hosmer-Lemeshov test. 
The collagen score values were stratified by the score tertiles and the Cox modeling-predicted and observed survival probabilities in the tertile strata compared with the D'Agostino - Nam $\chi^2$ statistic with two degrees of freedom (strata number - 2) [@DAgostino2003]. 
The performance and calibration statistics were computed with the development package [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions).

## Data and code avaibalility

Publicly available data sets were analyzed. 
Cleared data sets used for analyses will be made available upon request to the corresponding author. 
The transcriptome R analysis pipeline is available from GitHub (https://github.com/PiotrTymoszuk/collagen_pca).

\newpage

# Supplementary Tables

```{r tab-dge-int, tab.cap = 'Genes differentially regulated between the Collagen intermediate and Collagen low cluster tumors. The table is available as a supplementary Excel file.'}

suppl_tbl$dge_int[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable

```

```{r tab-dge-hi, tab.cap = 'Genes differentially regulated between the Collagen high and Collagen low cluster tumors. The table is available as a supplementary Excel file.'}

suppl_tbl$dge_hi[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable

```

```{r tab-react-int, tab.cap = 'Biochemical reactions significantly regulated in Collagen intermediate versus Collagen low shared by all study cohorts. The table is available as a supplementary Excel file.'}

suppl_tbl$react_int[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable

```

```{r tab-react-hi, tab.cap = 'Biochemical reactions significantly regulated in Collagen high versus Collagen low shared by all study cohorts. The table is available as a supplementary Excel file.'}

suppl_tbl$react_hi[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable

```

\newpage

# Supplementary Figures

```{r fig-norm-tumor-up, fig.width = mm_inch(suppl_fig$norm_tumor_up$w), fig.height = mm_inch(suppl_fig$norm_tumor_up$h), fig.cap = 'Expression of ALDH18A1, PYCR1, P4HB and PPIB in the normal prostate and prostate cancer tissue.'}

suppl_fig$norm_tumor_up$plot

```

__Supplementary Figure S\@ref(fig:fig-norm-tumor-up). Expression of__ _ALDH18A1_, _PYCR1_, _P4HB_ __and__ _PPIB_ __in the normal prostate and prostate cancer tissue.__ 
_Differences in expression of the collagen pathway genes (`r nrow(globals$genes_interest)` transcripts) between the donor-matched samples of the normal prostate and prostate cancer tissue were analyzed by two-tailed paired T test in the TCGA, GSE40272 and GSE70768 cohorts. P values were corrected for multiple testing with Benjamini-Hochberg method. log~2~-transformed expression levels for ALDH18A1 (A), PYCR1 (B), P4HB (C) and PPIB (D) are presented in plots as points. Sample matching by the tissue donor is indicated by gray lines. P values are presented in the plot captions. Numbers of tissue donors are indicated under the plots._

\newpage

```{r fig-norm-tumor-down, fig.width = mm_inch(suppl_fig$norm_tumor_down$w), fig.height = mm_inch(suppl_fig$norm_tumor_down$h), fig.cap = 'Expression of LAMB3 and COL4A2 in the normal prostate and prostate cancer tissue.'}

suppl_fig$norm_tumor_down$plot

```

__Supplementary Figure S\@ref(fig:fig-norm-tumor-down). Expression of__ _LAMB3_ __and__ _COL4A2_ __in the normal prostate and prostate cancer tissue.__ 
_Differences in expression of the collagen pathway genes (`r nrow(globals$genes_interest)` transcripts) between the donor-matched samples of the normal prostate and prostate cancer tissue were analyzed by two-tailed paired T test in the TCGA, GSE40272 and GSE70768 cohorts. P values were corrected for multiple testing with Benjamini-Hochberg method. log~2~-transformed expression levels for LAMB3 (A) and COL4A2 (B) are presented in plots as points. Sample matching by the tissue donor is indicated by gray lines. P values are presented in the plot captions. Numbers of tissue donors are indicated under the plots._

\newpage

```{r fig-corr1, fig.width = mm_inch(suppl_fig$corr1$w), fig.height = mm_inch(suppl_fig$corr1$h), fig.cap = 'Pairwise correlation of collagen pathway gene expression, TCGA and GSE16560 cohorts.'}

suppl_fig$corr1$plot

```

__Supplementary Figure S\@ref(fig:fig-corr1). Pairwise correlation of collagen pathway gene expression, TCGA and GSE16560 cohorts.__ 
_Pairwise correlation of the collagen pathway genes was investigated with Pearson's correlation test. P values were corrected for multiple testing with Benjamini-Hochberg method. Correlation coefficients (r) for significantly correlated gene pairs are presented in bubble plots. Point size corresponds to the absolute value of r, point color represents r values. Numbers of complete observations are displayed in the plot caption._

\newpage

```{r fig-corr2, fig.width = mm_inch(suppl_fig$corr2$w), fig.height = mm_inch(suppl_fig$corr2$h), fig.cap = 'Pairwise correlation of collagen pathway gene expression, GSE40272 and GSE70768 cohorts.'}

suppl_fig$corr2$plot

```

__Supplementary Figure S\@ref(fig:fig-corr2). Pairwise correlation of collagen pathway gene expression, GSE40272 and GSE70768 cohorts.__ 
_Pairwise correlation of the collagen pathway genes was investigated with Pearson's correlation test. P values were corrected for multiple testing with Benjamini-Hochberg method. Correlation coefficients (r) for significantly correlated gene pairs are presented in bubble plots. Point size corresponds to the absolute value of r, point color represents r values. Numbers of complete observations are displayed in the plot caption._

\newpage

```{r fig-corr3, fig.width = mm_inch(suppl_fig$corr3$w), fig.height = mm_inch(suppl_fig$corr3$h), fig.cap = 'Pairwise correlation of collagen pathway gene expression, GSE70769 cohorts.'}

suppl_fig$corr3$plot

```

__Supplementary Figure S\@ref(fig:fig-corr3). Pairwise correlation of collagen pathway gene expression, GSE70769 cohorts.__ 
_Pairwise correlation of the collagen pathway genes was investigated with Pearson's correlation test. P values were corrected for multiple testing with Benjamini-Hochberg method. Correlation coefficients (r) for significantly correlated gene pairs are presented in bubble plots. Point size corresponds to the absolute value of r, point color represents r values. Numbers of complete observations are displayed in the plot caption._

\newpage

```{r fig-clust-dev, fig.width = mm_inch(suppl_fig$clust_dev$w), fig.height = mm_inch(suppl_fig$clust_dev$h), fig.cap = 'Semi-supervised clustering of prostate cancer samples in respect to collagen pathway gene expression.'}

suppl_fig$clust_dev$plot

```

__Supplementary Figure S\@ref(fig:fig-clust-dev). Semi-supervised clustering of prostate cancer samples in respect to collagen pathway gene expression.__ 
_Tumor samples in the TCGA training cohort were assigned with the PAM (partition around medoids) algorithm with Manhattan distance metric into three clusters: Collagen^low^, Collagen^intermediate^ and Collagen^high^. Assignment of the tumor samples from the training GSE16560, GSE40272, GSE70768 and GSE70769 collectives to the collagen clusters was accomplished by the k-NN classifier (k-nearest neighbors, k = 7)._
_(A) Comparison of performance of several clustering algorithms (PAM: partition around medoids, HCL: hierarchical clustering/Ward D2 and KMEANS) and distance metrics (Euclidean, Manhattan, sum-of-squares and cosine) in the training TCGA cohort was measured by the amount of 'explained' clustering variance (ratio of the between-cluster sum of squares to the total sum of squares) and rate of correct cluster assignment in 10-fold cross-validation (CV). Number of complete observations and the clustering factors are shown next to the plot._
_(B, C) Determination of the cluster number for PAM/Manhattan distance clustering in the TCGA cohort by the bend of the within-cluster sum of squares (WSS) (B) and the peak of the mean silhouette statistic (C)._
_(D) Performance of the PAM/Manhattan distance clustering procedure in semi-supervised clustering in the training cohort and the test collectives measured by the 'explained' clustering variance statistic. Numbers of complete observations are presented in the Y axis._

\newpage

```{r fig-clust-hm, fig.width = mm_inch(suppl_fig$clust_hm$w), fig.height = mm_inch(suppl_fig$clust_hm$h), fig.cap = 'Expression levels of collagen pathway genes in the collagen clusters.'}

suppl_fig$clust_hm$plot

```

__Supplementary Figure S\@ref(fig:fig-clust-hm). Expression levels of collagen pathway genes in the collagen clusters.__ 
_The Collagen^low^, Collagen^intermediate^ and Collagen^high^ clusters were defined by PAM/Manhattan distance semi-supervised clustering in the training TCGA cohort and the GSE16560, GSE40272, GSE70768 and GSE70769 test collectives. Normalized levels (Z-score) of the collagen pathway transcripts in the tumor samples assigned to the collagen clusters are presented as heat maps. Tile intensities correspond to Z-score values. Numbers of samples assigned to the clusters are indicated under the plots._

\newpage

```{r fig-clust-gleason, fig.width = mm_inch(suppl_fig$clust_gleason$w), fig.height = mm_inch(suppl_fig$clust_gleason$h), fig.cap = 'Distribution of Gleason scores in the collagen clusters.'}

suppl_fig$clust_gleason$plot

```

__Supplementary Figure S\@ref(fig:fig-clust-gleason). Distribution of Gleason scores in the collagen clusters.__ 
_The Collagen^low^, Collagen^intermediate^ and Collagen^high^ clusters were defined by PAM/Manhattan distance semi-supervised clustering in the training TCGA cohort and the GSE16560, GSE40272, GSE70768 and GSE70769 test collectives. Distribution of sum Gleason scores between the collagen clusters was compared by Kruskal-Wallis test with $\eta^2$ effect size statistic. P values were corrected for multiple comparisons with Benjamini-Hochberg method. Gleason score values are presented in violin plots. Points depict single tumor samples. Red diamonds with whiskers represent medians with interquartile ranges. P and effect size statistic values are indicated in  the plot captions. Numbers of samples assigned to the clusters are shown under the plots._

\newpage

```{r fig-mcp-infil, fig.width = mm_inch(suppl_fig$mcp_infil$w), fig.height = mm_inch(suppl_fig$mcp_infil$h), fig.cap = 'MCP Counter estimates of non-malignant cell content in the collagen clusters.'}

suppl_fig$mcp_infil$plot

```

__Supplementary Figure S\@ref(fig:fig-mcp-infil). MCP Counter estimates of non-malignant cell content in the collagen clusters.__ 
_Levels of non-malignant cells (CAF: cancer-associated fibroblast, EC: endothelial cells, T cells, CD8^+^ T cells, NK cells: natural killer cells, B cells, macrophages, monocytes, neutrophils and mDC: myeloid dendritic cells) in the prostate cancer samples were estimated by the MCP counter algorithm. Differences in cell content between the Collagen^low^, Collagen^intermediate^ and Collagen^high^ clusters were analyzed with Kruskal-Wallis test. P values were corrected for multiple testing with Benjamini-Hochberg method. Mean minimum/maximum normalized cell amounts in the collagen clusters are depicted as thick lines. Tinted ribbons represent the 2 $\times$ SEM intervals. P values are shown in the Y axes. Numbers of samples assigned to the clusters are indicated under the plots._

\newpage

```{r fig-infil, fig.width = mm_inch(suppl_fig$xcell_infil$w), fig.height = mm_inch(suppl_fig$xcell_infil$h), fig.cap = 'xCell estimates of cancer-associated fibroblast and enbothelial cell infiltration in the collagen clusters.'}

suppl_fig$xcell_infil$plot

```

__Supplementary Figure S\@ref(fig:fig-infil). xCell estimates of cancer-associated fibroblast and endothelial cell infiltration in the collagen clusters.__ 
_Levels of cancer-associated fibroblasts (CAF) (A) and endothelial cells (B) in the prostate cancer samples were estimated by the xCell algorithm. Differences in CAF and endothelial cell content between the Collagen^low^, Collagen^intermediate^ and Collagen^high^ clusters were analyzed with Kruskal-Wallis test with $\eta^2$ effect size statistic. P values were corrected for multiple testing with Benjamini-Hochberg method. Infiltration estimates are presented in violin plots. Points indicate single tumor samples. Red diamonds with whiskers represent medians with interquartile ranges. P and effect size statistic values are displayed in the plot captions. Numbers of samples in the clusters are indicated under the plots._

\newpage

```{r fig-dge1, fig.width = mm_inch(suppl_fig$dge1$w), fig.height = mm_inch(suppl_fig$dge1$h), fig.cap = 'Genes differentially expressed in the collagen clusters, TCGA and GSE16560 cohorts.'}

suppl_fig$dge1$plot

```

__Supplementary Figure S\@ref(fig:fig-dge1). Genes differentially expressed in the collagen clusters, TCGA and GSE16560 cohorts.__ 
_Genes differentially regulated between the Collagen^intermediate^ and Collagen^low^ clusters and between the Collagen^high^ and Collagen^low^ clusters were identified by Benjamini-Hochberg-corrected one-way ANOVA and linear modeling. Benjamini-Hochberg-corrected p values (pFDR) and log~2~ fold regulation estimates are presented as volcano plots. Each points represent a significantly regulated gene. Point color codes for regulation sign (red: upregulated, blue: downregulated). Top most significant transcripts are labeled with their names. Numbers of samples assigned to the clusters are presented in the plot captions. Numbers of regulated transcripts are indicated under the plots._

\newpage

```{r fig-dge2, fig.width = mm_inch(suppl_fig$dge2$w), fig.height = mm_inch(suppl_fig$dge2$h), fig.cap = 'Genes differentially expressed in the collagen clusters, GSE40272 and GSE70768 cohorts.'}

suppl_fig$dge2$plot

```

__Supplementary Figure S\@ref(fig:fig-dge2). Genes differentially expressed in the collagen clusters, GSE40272 and GSE70768 cohorts.__ 
_Genes differentially regulated between the Collagen^intermediate^ and Collagen^low^ clusters and between the Collagen^high^ and Collagen^low^ clusters were identified by Benjamini-Hochberg-corrected one-way ANOVA and linear modeling. Benjamini-Hochberg-corrected p values (pFDR) and log~2~ fold regulation estimates are presented as volcano plots. Each points represent a significantly regulated gene. Point color codes for regulation sign (red: upregulated, blue: downregulated). Top most significant transcripts are labeled with their names. Numbers of samples assigned to the clusters are presented in the plot captions. Numbers of regulated transcripts are indicated under the plots._

\newpage

```{r fig-dge3, fig.width = mm_inch(suppl_fig$dge3$w), fig.height = mm_inch(suppl_fig$dge3$h), fig.cap = 'Genes differentially expressed in the collagen clusters, GSE70769 cohort.'}

suppl_fig$dge3$plot

```

__Supplementary Figure S\@ref(fig:fig-dge3). Genes differentially expressed in the collagen clusters, GSE70769 cohort.__ 
_Genes differentially regulated between the Collagen^intermediate^ and Collagen^low^ clusters and between the Collagen^high^ and Collagen^low^ clusters were identified by Benjamini-Hochberg-corrected one-way ANOVA and linear modeling. Benjamini-Hochberg-corrected p values (pFDR) and log~2~ fold regulation estimates are presented as volcano plots. Each points represent a significantly regulated gene. Point color codes for regulation sign (red: upregulated, blue: downregulated). Top most significant transcripts are labeled with their names. Numbers of samples assigned to the clusters are presented in the plot captions. Numbers of regulated transcripts are indicated under the plots._

\newpage

```{r fig-go1, fig.width = mm_inch(suppl_fig$go1$w), fig.height = mm_inch(suppl_fig$go1$h), fig.cap = 'Top most significantly enriched biological process GO terms in the collagen clusters, TCGA, GSE16560 and GSE40272 cohorts.'}

suppl_fig$go1$plot

```

__Supplementary Figure S\@ref(fig:fig-go1). Top most significantly enriched biological process GO terms in the collagen clusters, TCGA, GSE16560 and GSE40272 cohorts.__ 
_Genes differentially regulated between the Collagen^intermediate^ and Collagen^low^ clusters and between the Collagen^high^ and Collagen^low^ clusters were identified by Benjamini-Hochberg-corrected one-way ANOVA and linear modeling. Biological process gene ontology (GO) term enrichment was investigated in the sets of significantly regulated genes with the goana tool._
_Benjamini-Hochberg corrected p values (pFDR) for the top 20 most significantly enriched GO terms in the Collagen^intermediate^ and in the Collagen^high^ tumors in the TCGA (A), GSE16560 (B) and GSE40272 (C) cohorts are presented as bar plots._

\newpage

```{r fig-go2, fig.width = mm_inch(suppl_fig$go2$w), fig.height = mm_inch(suppl_fig$go2$h), fig.cap = 'Top most significantly enriched biological process GO terms in the collagen clusters, GSE70768 and GSE70769 cohorts.'}

suppl_fig$go2$plot

```

__Supplementary Figure S\@ref(fig:fig-go2). Top most significantly enriched biological process GO terms in the collagen clusters, GSE70768 and GSE70769 cohorts.__ 
_Genes differentially regulated between the Collagen^intermediate^ and Collagen^low^ clusters and between the Collagen^high^ and Collagen^low^ clusters were identified by Benjamini-Hochberg-corrected one-way ANOVA and linear modeling. Biological process gene ontology (GO) term enrichment was investigated in the sets of significantly regulated genes with the goana tool._
_Benjamini-Hochberg corrected p values (pFDR) for the top 20 most significantly enriched GO terms in the Collagen^intermediate^ and in the Collagen^high^ tumors in the GSE70768 (A) and GSE70768 (B) cohorts are presented as bar plots._

\newpage

```{r fig-pathview1, fig.width = mm_inch(suppl_fig$pathview1$w), fig.height = mm_inch(suppl_fig$pathview1$h), fig.cap = 'Components of the focal adhesion pathway regulated in the collagen clusters.'}

suppl_fig$pathview1$plot

```

__Supplementary Figure S\@ref(fig:fig-pathview1). Components of the focal adhesion pathway regulated in the collagen clusters.__ 
_Pooled estimates for fold regulation of gene expression between the Collagen^intermediate^ and Collagen^low^ clusters (A) and between the Collagen^high^ and Collagen^low^ clusters (B) were calculated with the inverse variance method. Modulation of components of the KEGG's focal adhesion pathway based on the common gene regulation estimates was visualized with the pathview tool. Pathway component color corresponds to the magnitude of its regulation._

\newpage

```{r fig-pathview2, fig.width = mm_inch(suppl_fig$pathview2$w), fig.height = mm_inch(suppl_fig$pathview2$h), fig.cap = 'Components of the ECM-receptor interaction pathway regulated in the collagen clusters.'}

suppl_fig$pathview2$plot

```

__Supplementary Figure S\@ref(fig:fig-pathview2). Components of the ECM-receptor interaction pathway regulated in the collagen clusters.__ 
_Pooled estimates for fold regulation of gene expression between the Collagen^intermediate^ and Collagen^low^ clusters (A) and between the Collagen^high^ and Collagen^low^ clusters (B) were calculated with the inverse variance method. Modulation of components of the KEGG's ECM-receptor interaction pathway based on the common gene regulation estimates was visualized with the pathview tool. Pathway component color corresponds to the magnitude of its regulation._

\newpage

```{r fig-pathview3, fig.width = mm_inch(suppl_fig$pathview3$w), fig.height = mm_inch(suppl_fig$pathview3$h), fig.cap = 'Components of the actin cytoskeleton pathway regulated in the collagen clusters.'}

suppl_fig$pathview3$plot

```

__Supplementary Figure S\@ref(fig:fig-pathview3). Components of the actin cytoskeleton pathway regulated in the collagen clusters.__ 
_Pooled estimates for fold regulation of gene expression between the Collagen^intermediate^ and Collagen^low^ clusters (A) and between the Collagen^high^ and Collagen^low^ clusters (B) were calculated with the inverse variance method. Modulation of components of the KEGG's regulation of actin cytoskeleton pathway based on the common gene regulation estimates was visualized with the pathview tool. Pathway component color corresponds to the magnitude of its regulation._

\newpage

```{r fig-pathview4, fig.width = mm_inch(suppl_fig$pathview4$w), fig.height = mm_inch(suppl_fig$pathview4$h), fig.cap = 'Components of the small cell lung cancer pathway pathway regulated in the collagen clusters.'}

suppl_fig$pathview4$plot

```

__Supplementary Figure S\@ref(fig:fig-pathview4). Components of the small cell lung cancer pathway pathway regulated in the collagen clusters.__ 
_Pooled estimates for fold regulation of gene expression between the Collagen^high^ and Collagen^low^ clusters were calculated with the inverse variance method. Modulation of components of the KEGG's small cell lung cancer pathway based on the common gene regulation estimates was visualized with the pathview tool. Pathway component color corresponds to the magnitude of its regulation._

\newpage

\newpage

```{r fig-spia1, fig.width = mm_inch(suppl_fig$spia1$w), fig.height = mm_inch(suppl_fig$spia1$h), fig.cap = 'Modulation of signaling pathways in the collagen clusters, TCGA and GSE16560 cohorts.'}

suppl_fig$spia1$plot

```

__Supplementary Figure S\@ref(fig:fig-spia1). Modulation of signaling pathways in the collagen clusters, TCGA and GSE16560 cohorts.__ 
_Genes differentially regulated between the Collagen^intermediate^ and Collagen^low^ clusters and between the Collagen^high^ and Collagen^low^ clusters were identified by Benjamini-Hochberg-corrected one-way ANOVA and linear modeling. Signaling pathway modulation based on fold regulation estimates of the significantly regulated genes in the TCGA (A) and GSE16560 cohorts (B) was investigated with the SPIA tool._
_Benjamini-Hochberg-corrected p values (pFDR) and estimates of the pathway regulation (tA) versus the Collagen^low^ cluster are presented as point plots. Point color codes for the pathway activation status (red: activated, blue: inhibited, gray: not significant). Top most significantly modulated pathways are labeled with their names. Numbers of significantly activated and inhibited pathways are displayed under the plots._

\newpage

```{r fig-spia2, fig.width = mm_inch(suppl_fig$spia2$w), fig.height = mm_inch(suppl_fig$spia2$h), fig.cap = 'Modulation of signaling pathways in the collagen clusters, GSE40272 and GSE70768 cohorts.'}

suppl_fig$spia2$plot

```

__Supplementary Figure S\@ref(fig:fig-spia2). Modulation of signaling pathways in the collagen clusters, GSE40272 and GSE70768 cohorts.__ 
_Genes differentially regulated between the Collagen^intermediate^ and Collagen^low^ clusters and between the Collagen^high^ and Collagen^low^ clusters were identified by Benjamini-Hochberg-corrected one-way ANOVA and linear modeling. Signaling pathway modulation based on fold regulation estimates of the significantly regulated genes in the GSE40272 (A) and GSE70768 cohorts (B) was investigated with the SPIA tool._
_Benjamini-Hochberg-corrected p values (pFDR) and estimates of the pathway regulation (tA) versus the Collagen^low^ cluster are presented as point plots. Point color codes for the pathway activation status (red: activated, blue: inhibited, gray: not significant). Top most significantly modulated pathways are labeled with their names. Numbers of significantly activated and inhibited pathways are displayed under the plots._

\newpage

```{r fig-spia3, fig.width = mm_inch(suppl_fig$spia3$w), fig.height = mm_inch(suppl_fig$spia3$h), fig.cap = 'Modulation of signaling pathways in the collagen clusters, GSE70769 cohort.'}

suppl_fig$spia3$plot

```

__Supplementary Figure S\@ref(fig:fig-spia3). Modulation of signaling pathways in the collagen clusters, GSE70769 cohort.__ 
_Genes differentially regulated between the Collagen^intermediate^ and Collagen^low^ clusters and between the Collagen^high^ and Collagen^low^ clusters were identified by Benjamini-Hochberg-corrected one-way ANOVA and linear modeling. Signaling pathway modulation based on fold regulation estimates of the significantly regulated genes in the GSE70769 cohort was investigated with the SPIA tool._
_Benjamini-Hochberg-corrected p values (pFDR) and estimates of the pathway regulation (tA) versus the Collagen^low^ cluster are presented as point plots. Point color codes for the pathway activation status (red: activated, blue: inhibited, gray: not significant). Top most significantly modulated pathways are labeled with their names. Numbers of significantly activated and inhibited pathways are displayed under the plots._

\newpage

```{r fig-react-int, fig.width = mm_inch(suppl_fig$react_int$w), fig.height = mm_inch(suppl_fig$react_int$h), fig.cap = 'Differences in predicted biochemical reaction activity between Collagen intermediate and Collagen low cancers.'}

suppl_fig$react_int$plot

```

__Supplementary Figure S\@ref(fig:fig-react-int). Differences in predicted biochemical reaction activity between Collagen intermediate and Collagen low cancers.__ 
_Estimates of biochemical reaction modulation between the Collagen^intermediate^ and Collagen^low^ clusters were computed by evaluating gene association rules in the Recon2 human metabolism model with the BiGGR and biggrExtra tools. Pooled reaction regulation estimates and their 95% confidence interval for reactions found significantly modulated in all study cohorts were calculated with the inverse variance method. The pooled regulation estimates with 95% confidence intervals are shown in a Forest plot. Reactions are labeled with their BiGG database identifiers and were classified as members of aminoacid metabolism (AA), bile acid synthesis (BAS), metabolism of extracellular matrix glycoproteins (ECM), eicosanoid metabolism (ECSM), energy metabolism (Energy), fatty acid metabolism (FAM), fatty acid oxidation (FAOx) and aminoacid amine group metabolism (Urea).__

\newpage

```{r fig-react-hi, fig.width = mm_inch(suppl_fig$react_hi$w), fig.height = mm_inch(suppl_fig$react_hi$h), fig.cap = 'Differences in predicted biochemical reaction activity between Collagen high and Collagen low cancers.'}

suppl_fig$react_hi$plot

```

__Supplementary Figure S\@ref(fig:fig-react-hi). Differences in predicted biochemical reaction activity between Collagen high and Collagen low cancers.__ 
_Estimates of biochemical reaction modulation between the Collagen^high^ and Collagen^low^ clusters were computed by evaluating gene association rules in the Recon2 human metabolism model with the BiGGR and biggrExtra tools. Pooled reaction regulation estimates and their 95% confidence interval for reactions found significantly modulated in all study cohorts were calculated with the inverse variance method. The pooled regulation estimates with 95% confidence intervals are shown in a Forest plot. Reactions are labeled with their BiGG database identifiers and were classified as members of aminoacid metabolism (AA), bile acid synthesis (BAS), metabolism of extracellular matrix glycoproteins (ECM), eicosanoid metabolism (ECSM), energy metabolism (Energy), fatty acid metabolism (FAM), fatty acid oxidation (FAOx) and aminoacid amine group metabolism (Urea)._

\newpage

```{r fig-faox, fig.width = mm_inch(suppl_fig$faox$w), fig.height = mm_inch(suppl_fig$faox$h), fig.cap = 'Predicted regulation of fatty acid oxidation reactions in the Collagen clusters.'}

suppl_fig$faox$plot

```

__Supplementary Figure S\@ref(fig:fig-faox). Predicted regulation of fatty acid oxidation reactions in the Collagen clusters.__ 
_Estimates of biochemical reaction modulation between the Collagen^intermediate^ and Collagen^low^ (A) and between the Collagen^high^ and Collagen^low^ clusters (B) were computed by evaluating gene association rules in the Recon2 human metabolism model with the BiGGR and biggrExtra tools. Predicted regulation estimates for `r nrow(meta_hg$faox_regulation_int[[1]])` reactions involved in fatty acid oxidation (FAOx) are presented as bars. Bar colors code for regulation sign and significance. Numbers of significantly activated and inhibited reactions are displayed in the graph captions._

\newpage

```{r fig-tca, fig.width = mm_inch(suppl_fig$tca$w), fig.height = mm_inch(suppl_fig$tca$h), fig.cap = 'Predicted regulation of TCA cycle reactions in the Collagen clusters.'}

suppl_fig$tca$plot

```

__Supplementary Figure S\@ref(fig:fig-tca). Predicted regulation of TCA cycle reactions in the Collagen clusters.__ 
_Estimates of biochemical reaction modulation between the Collagen^intermediate^ and Collagen^low^ (A) and between the Collagen^high^ and Collagen^low^ clusters (B) were computed by evaluating gene association rules in the Recon2 human metabolism model with the BiGGR and biggrExtra tools. Predicted regulation estimates for the TCA (Tri-Carboxylic Acid) cycle reactions are presented as bars. Bar colors code for regulation sign and significance. The reactions are labeled with their BiGG database IDs._

\newpage

```{r fig-oxphos, fig.width = mm_inch(suppl_fig$oxphos$w), fig.height = mm_inch(suppl_fig$oxphos$h), fig.cap = 'Predicted regulation of oxidative phosphorylation reactions in the Collagen clusters.'}

suppl_fig$oxphos$plot

```

__Supplementary Figure S\@ref(fig:fig-oxphos). Predicted regulation of oxidative phosphorylation reactions in the Collagen clusters.__ 
_Estimates of biochemical reaction modulation between the Collagen^intermediate^ and Collagen^low^ (A) and between the Collagen^high^ and Collagen^low^ clusters (B) were computed by evaluating gene association rules in the Recon2 human metabolism model with the BiGGR and biggrExtra tools. Predicted regulation estimates for the oxidative phosphorylation (OxPhos) reactions are presented as bars. Bar colors code for regulation sign and significance. The reactions are labeled with the OxPhos complex names._

\newpage

```{r fig-cs-gleason, fig.width = mm_inch(suppl_fig$cs_gleason$w), fig.height = mm_inch(suppl_fig$cs_gleason$h), fig.cap = 'Association of the collagen score with the Gleason score and collagen cluster assignment.'}

suppl_fig$cs_gleason$plot

```

__Supplementary Figure S\@ref(fig:fig-cs-gleason). Association of the collagen score with the Gleason score and collagen cluster assignment.__ 
_Collagen Score based on the normalized expression levels of the collagen pathway genes was developed using the elastic net Cox modeling algorithm._
_(A) Correlation of the collagen score values with the sum Gleason score was investigated by Spearman's correlation test. P values were corrected for multiple comparisons by Benjamini-Hochberg method. Collagen score and Gleason score values are presented as points. The points were jittered horizontally to reduce overplotting. Blue lines with gray ribbons represent linear trends with 95% confidence regions. Correlation coeffcients ($\rho$) with 95% confidence intervals, p values and numbers of complete observations are displayed in the plot captions._
_(B) Levels of the collagen score were compared between the Collagen^low^, Collagen^intermediate^ and Collagen^high^ clusters by one-way ANOVA with $\eta^2$ effect size statistic. P values were corrected for multiple comparisons with Benjamini-Hochberg method. Collagen score values are presented in violin plots. Points depict single tumor samples. Red diamonds and whiskers represent medians with interquartile ranges. P and effect size statistic values are shown in the plot captions. Numbers of samples assigned to the clusters are displayed under the plots._

\newpage

```{r fig-cs-multi, fig.width = mm_inch(suppl_fig$cs_multi$w), fig.height = mm_inch(suppl_fig$cs_multi$h), fig.cap = 'Prediction of prostate cancer survival by the collagen score with Gleason score and age confounder variables.'}

suppl_fig$cs_multi$plot

```

__Supplementary Figure S\@ref(fig:fig-cs-multi). Prediction of prostate cancer survival by the collagen score with Gleason score and age confounder variables.__ 
_Association of the collagen score with biochemical relapse-free (RFS: TCGA, GSE40272, GSE70768, GSE70769) or overall survival (OS: GSE16560) were investigated by uni-variable Cox proportional hazard regression or by multi-variable Cox modeling with age and Gleason score as confounders. Hazard ratio of the collagen score estimates with 95% confidence intervals for the uni- and multi-variable models are presented in a Forest plot. Survival response type and numbers of complete observations are indicated in the Y axis._

\newpage

# References

### 2023-07-26 report/markdown/manuscript_figures_tables.Rmd
Parts of the manuscript
Figures, Tables and Supplementary Tables for the manuscript with their legends

---
title: "Prognostic and biologic relevance of collagen biosynthesis pathway in prostate cancer"
subtitle: "Bioinformatic methods, figures and tables"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::word_document2:   
    reference_docx: ms_template.docx

bibliography: coll_biblio.bib

csl: frontiers-in-immunology.csl

header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

```{r, setup, include = FALSE}

library(bookdown)
library(flextable)

knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dev = "png", 
                      dpi = 600, 
                      tab.cap.pre = 'Table ', 
                      tab.cap.sep = " ", 
                      tab.cap.style = 'Table Heading')

set_flextable_defaults(font.family = 'Cambria', font.size = 10)

```

\newpage

# Tables

```{r tab-cohorts, tab.cap = 'Characteristic of the analyzed cohorts. Numeric variables are presented as medians with interquartile ranges (IQR) and ranges. Qualitative variables are presented as percentages of categories within the complete observation set.'}

flextable::flextable(paper_tbl$cohorts) %>% 
  set_widths(c(3, rep(3.5, 5))) %>% 
  merge_v(1) %>% 
  footnote(1, 1, 
           value = as_paragraph('PSA: prostate-specific antigen.'), 
           part = 'header', 
           ref_symbols = 'a') %>%
  theme_vanilla

```

\newpage

```{r tab-genes, tab.cap = 'Collagen genes of interest and their classification.'}

flextable::flextable(paper_tbl$genes) %>% 
  set_widths(rep(4, 3)) %>% 
  italic(j = 1, part = 'body') %>% 
  footnote(1, 3, 
           value = as_paragraph('ECM: extracellular matrix.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

```

\newpage

# Figures

```{r fig-norm-tumor, fig.width = figur::convert(paper_fig$norm_tumor, to = 'in')$w, fig.height = figur::convert(paper_fig$norm_tumor, to = 'in')$h, fig.cap = 'Differences in tissue diffusion capacity and expression of the collagen pathway genes between the prostate cancer and benign tissue.'}

paper_fig$norm_tumor$plot

```

__Figure \@ref(fig:fig-norm-tumor). Differences in tissue diffusion capacity and expression of the collagen pathway genes between the prostate cancer and benign tissue.__

_(A) Representative MRI axial sequences of the prostate of an 83-year-old patient with prostate cancer (left panel). The T2-weighted images (a) show a 23mm suspicious hypo-intense lesion in the left peripheral zone (red star) of the prostate, which has an early enhancement in the contrast-enhanced T1-weighted image (b). The DWI scan (c) and ADC map (d) show a strong diffusion restriction (black and white character) of the tumor area suggesting a PI-RADS 5 lesion. Statistical significance for differences in diffusion capacity in patient-matched tumor and benign tissue measured by ADC was determined by paired T test with Cohen's d effect size statistic (right panel). Single ADC values are visualized as points, grey lines connect measurements of the same donors. The number of measurement pairs, effect size and p value are displayed in the plot caption._

_(B) Differences in $log_2$-transformed expression levels of `r nrow(globals$genes_interest)` genes related to collagen metabolism between donor-matched tumor and benign prostate tissue were investigated by paired T test with Cohen's d effect size statistic in three published cohorts. P values were corrected for multiple testing with the false discovery method. Results for gene significantly regulated in at least two cohorts are presented. Lines represent means, 2 $\times$ SEM (standard error of the mean) intervals are visualized as tinted ribbons. Gene symbols, effect sizes and p values are shown in the Y axes. Significant effects are highlighted in bold. Numbers of tissue pairs are displayed in the plot captions._

\newpage

```{r fig-coll-clusters, fig.width = figur::convert(paper_fig$coll_clusters, to = 'in')$w, fig.height = figur::convert(paper_fig$coll_clusters, to = 'in')$h, fig.cap = 'Collagen clusters of prostate carcinoma and their clinical characteristic.'}

paper_fig$coll_clusters$plot

```

__Figure \@ref(fig:fig-coll-clusters). Collagen clusters of prostate carcinoma and their clinical characteristic.__

_Prostate cancers in the training TCGA cohort were clustered in respect to normalized $log_2$-transformed expression levels of the collagen pathway genes of interest by the PAM algorithm with Manhattan distance between the observations. Three clusters, collagen^low^ (low), collagen^intermediate^ (int) and collagen^high^ (hi) were defined. Cancer samples from the test cohorts (`r globals$study_labels[globals$study_labels != 'TCGA'] %>% paste(collapse = ', ')`) were assigned to the collagen clusters with an inverse distance weighted k-nearest neighbor classifier._

_(A) Mean normalized $log_2$-transformed expression levels of the collagen pathway genes in the collagen clusters and cohorts presented in a heat map._

_(B) Distribution of the collagen clusters expressed as percentage of the cohort cancer samples visualized in a stack plot. Total numbers of cancer samples are displayed in the X axis. _

_(C, D) Distribution of Gleason scores (GS, C) and pathological tumor stages (pT, D) in the collagen clusters. Statistical significance was determined by $\chi^2$ test with Cramer's V effect size statistic. P values were corrected for multiple testing with the false discovery rate method. Gleason score and tumor stage distributions are shown in stack plots. Effect sizes and p values are displayed in the plot captions. Numbers of observations in the clusters are indicated in the X axes._

\newpage

```{r fig-cluster-biology, fig.width = figur::convert(paper_fig$cluster_biology, to = 'in')$w, fig.height = figur::convert(paper_fig$cluster_biology, to = 'in')$h, fig.cap = 'Infiltration, biological processes, signaling and metabolic pathways differentially regulated in the collagen clusters.'}

paper_fig$cluster_biology$plot

```

__Figure \@ref(fig:fig-cluster-biology). Infiltration, biological processes, signaling and metabolic pathways differentially regulated in the collagen clusters.__

_(A) Counts of cancer-associated fibroblasts (CAF) and endothelial cells (EC) estimated by the MCP counter algorithm were found to differ significantly between the collagen clusters in all investigated collectives. Statistical significance was assessed by Kruskal-Wallis test with $\eta^2$ effect size statistic. P values were corrected for multiple comparison with the false discovery rate method. Counts of fibroblasts and endothelial cells in the collagen clusters in the TCGA cohort are shown in violin plots. Red diamonds with whiskers represented medians with interquartile ranges. Single tumor samples are visualized as points. Effect sizes and p values are displayed in the plot captions. Numbers of observations in the clusters are indicated in the X axes._

_(B) Gene set variation analysis (GSVA) with Reactome pathway gene signatures. Statistical significance for differences in signature single sample gene set enrichment analysis scores (ssGSEA) between the collagen clusters was determined by one-way ANOVA with $\eta^2$ effect size statistic. P values were corrected for multiple testing with the false discovery rate method. Mean ssGSEA scores in the clusters and cohorts for signatures significantly regulated in at least four cohorts are presented in a heat map. Full analysis results are presented in Supplementary Table S6._

_(C) Modulation of KEGG-listed signaling pathways in the collagen^high^ (hi) or collagen^intermediate^ (hi) as compared with collagen^low^ tumors was predicted by the SPIA algorithm based on differential gene expression estimates. Predicted magnitude of pathway modulation (tA parameter) for pathways significantly modulated in at least four cohorts was visualized in bubble plots (tA < 0: inhibition, blue; tA > 0, activation, red). Points are labeled with their tA values. Significant effects are highlighted in bold. Full analysis results are presented in Supplementary Table S8._

_(D) Modulation of Recon2 model metabolic reactions in the collagen^high^ (hi) or collagen^intermediate^ (hi) as compared with collagen^low^ tumors was predicted by the BiGGR and biggrExtra algorithms based on differential gene expression estimates. Enrichment of significantly activated and significantly inhibited reactions in the collagen^high^ or collagen^intermediate^ clusters within the Recon model metabolic subsystems was investigated by Fisher's exact test. Enrichment p values were corrected for multiple testing with the false discovery rate method. Enrichment odds ratios for activated (red) and inhibited reaction (blue) for significantly enriched metabolic subsystems shared by at least four cohorts are presented in bubble plots. Points are labeled with their odds ration values. Full analysis results are presented in Supplementary Table S10._

\newpage

# References

### 2023-07-27 report/markdown/manuscript_supplement.Rmd
Supplementary material backbone

---
title: "Prognostic and biologic relevance of collagen biosynthesis pathway in prostate cancer"
subtitle: "Supplementary material, transcriptome part"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::word_document2:   
    reference_docx: ms_template.docx

bibliography: coll_biblio.bib

csl: frontiers-in-immunology.csl

header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

```{r, setup, include = FALSE}

library(bookdown)
library(flextable)

knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dev = "png", 
                      dpi = 600, 
                      tab.cap.pre = 'Table ', 
                      tab.cap.sep = " ", 
                      tab.cap.style = 'Table Heading')

set_flextable_defaults(font.family = 'Cambria', font.size = 10)

```

\newpage

# Supplementary Tables

```{r tab-norm-tumor, tab.cap = "Expression of the collagen pathway genes in the malignant and benign tissue compared by paired T test with Cohen's d effect size statistic. P values were corrected for multiple testing with the false discovery rate method. log2-transformed expression values are presented as medians with interquartile ranges (IQR) and ranges. The table is available as a supplementary Excel file."}

suppl_paper_tbl$norm_tumor[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

```{r tab-cluster, tab.cap = 'Expression of the cluster-defining collagen pathway genes in the collagen clusters of prostate cancer. Statistical significance was assessed by one-way ANOVA with eta-squared effect size statistic. P values were corrected for multiple testing with the false discovery rate method. log2-transformed expression values are presented as medians with interquartile ranges (IQR) and ranges. The table is available as a supplementary Excel file.'}

suppl_paper_tbl$cluster[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

\newpage

```{r tab-clinic, tab.cap = 'Clinical characteristic of the collagen clusters. Numeric variables are presented as medians with interquartile ranges (IQR) and ranges. Nominal variables are presented as percentages and counts of categories within the cluster.'}

flextable::flextable(suppl_paper_tbl$clinic) %>% 
  set_widths(c(2.2, 3.5, rep(4.5, 3), 2.8, 2.8)) %>% 
  merge_v(1:2) %>% 
  footnote(1, 2, 
           value = as_paragraph('PSA: prostate-specific antigen.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 6:7, 
           value = as_paragraph("Categorical variables: \u03C7\u00B2 test with Cramer's V effec size statistic. Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. P values corrected for multiple testing with the false discovery rate method."), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

```

\newpage

```{r tab-mcp, tab.cap = 'Non-malignant cell numbers predicted for the collagen Non-maignant cell fractions predicted for the collagen Statistical significance was assessed by Kruskal-Wallis test with eta-squared effect size statistic. P values were corrected for multiple testing with the false discovery method. The table is available as a supplementary Excel file.'}

suppl_paper_tbl$mcp[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

```{r tab-xcell, tab.cap = 'clusters by the MCP counter algorithm. clusters by the xCell algorithm. Statistical significance was assessed by Kruskal-Wallis test with eta-squared effect size statistic. P values were corrected for multiple testing with the false discovery method. The table is available as a supplementary Excel file.'}

suppl_paper_tbl$xcell[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

```{r tab-gsva, tab.cap = 'Gene set variation analysis with the Reactome pathway gene signatures. Differences between the collagen intermediate or high clusters versus collagen low cancers were investigated by one-way ANOVA with eta-squared effect size statistic and linear modeling. Results for signatures significantly regulated with moderate-to-large effect size (eta-squared at leat 0.06) in at least four cohorts are presented. P values were corrected for multiple testing with the false discovery rate method (FDR). The table is available as a supplementary Excel file.'}

suppl_paper_tbl$gsva[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

```{r tab-dge, tab.cap = 'Genes differentially expressed in the collagen intermediate or high cluster as compared with collagen low cancers were identified by one-way ANOVA and linear modeling with the 1.25-fold regulation cutoff P values were corrected for multiple testing with the false discovery rate method (FDR). The table is available as a supplementary Excel file.'}

suppl_paper_tbl$dge[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

```{r tab-signaling, tab.cap = 'Signaling pathway activity in the collagen clusters investigated by the SPIA algorithm. Resulat for signaling pathways significantly activated or inhibited in at least four cohorts are shown. The table is available as a supplementary Excel file.'}

suppl_paper_tbl$signaling[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

```{r tab-reactions, tab.cap = 'Biochemical reactions predicted to be significantly activated in the collagen high or collagen low cluster as compared with collagen low cancers. Statistical significance was determined by Monte Carlo simulation. P values were corrected for multiple testing with the false discovery rate method. The table is available as a supplementary Excel file.'}

suppl_paper_tbl$reactions[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

```{r tab-subsystems, tab.cap = "Results of enrichment analysis for significantly activated and inhibited biochemical reaction within the Recon metabolism subsystem. Statistical significance was determined by Fisher's exact test corrected for multiple testing with the false discovery rate method (FDR). The table is available as a supplementary Excel file."}

suppl_paper_tbl$subsystems[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

\newpage

# Supplementary Figures

\newpage

# References

### 2023-07-27 report/markdown/manuscript_figures_tables.Rmd
Supplementary material backbone
First supplementary figures
Supplementary figures for the normal - tumor comparison, cluster definition, infiltration prediction, GSVA and differential gene expression

@@
 
-_(A) Mean normalized $log_2$-transformed expression levels of the collagen pathway genes in the collagen clusters and cohorts presented in a heat map._
+_(A) Mean normalized $log_2$-transformed expression levels of the collagen pathway genes in the collagen clusters and cohorts presented in a heat map. ECM: extracellular matrix._
 
@@
 Effect sizes and p values are displayed in the plot captions.
-Numbers of observations in the clusters are indicated in the X axes._
+Numbers of observations in the clusters are indicated in the X axes.
+Full analysis results are presented in Supplementary Table S4._
 
@@
 Mean ssGSEA scores in the clusters and cohorts for signatures significantly regulated in at least four cohorts are presented in a heat map.
-Full analysis results are presented in Supplementary Table S6._
+Full analysis results are presented in Supplementary Table S6.
+ECM: extracellular matrix, GF: growth factor._
 
@@
 Significant effects are highlighted in bold.
-Full analysis results are presented in Supplementary Table S8._
+Full analysis results are presented in Supplementary Table S8.
+ECM: extracellular matrix, ACT: actin, SCLC: small cell lung carcinoma, S.
+aureus: Staphylococcus aureus, SLE: systemic lupus erythematosus._
 
@@
 Points are labeled with their odds ration values.
-Full analysis results are presented in Supplementary Table S10._
+Full analysis results are presented in Supplementary Table S10.
+Val: valine, Leu: leucine, Ile: isoleucine, FAOX: fatty acid oxidation._
 

### 2023-07-27 report/markdown/manuscript_supplement.Rmd
First supplementary figures
Supplementary figures for the normal - tumor comparison, cluster definition, infiltration prediction, GSVA and differential gene expression

@@
 flextable::flextable(suppl_paper_tbl$clinic) %>% 
-  set_widths(c(2.2, 3.5, rep(4.5, 3), 2.8, 2.8)) %>% 
+  set_widths(c(2.2, 3, rep(3.5, 3), 2.8, 2.8)) %>% 
   merge_v(1:2) %>% 

### 2023-07-27 report/markdown/manuscript_figures_tables.Rmd
Supplementary Material, markdown and docx file
Supplementary Material with Supplementary Figures, Supplementary Tables and their legends

@@
 title: "Prognostic and biologic relevance of collagen biosynthesis pathway in prostate cancer"
-subtitle: "Bioinformatic methods, figures and tables"
+subtitle: "Transcriptome methods, figures and tables"
 date: "`r format(Sys.time(), '%Y-%m-%d')`"
@@
 
-_(D) Modulation of Recon2 model metabolic reactions in the collagen^high^ (hi) or collagen^intermediate^ (hi) as compared with collagen^low^ tumors was predicted by the BiGGR and biggrExtra algorithms based on differential gene expression estimates.
+_(D) Modulation of Recon2 model metabolic reactions in the collagen^high^ (hi) or collagen^intermediate^ (hi) as compared with collagen^low^ tumors was predicted by the BiGGR and biggrExtra algorithms based on whole genome differential gene expression estimates.
 Enrichment of significantly activated and significantly inhibited reactions in the collagen^high^ or collagen^intermediate^ clusters within the Recon model metabolic subsystems was investigated by Fisher's exact test.

### 2023-07-27 report/markdown/manuscript_supplement.Rmd
Supplementary Material, markdown and docx file
Supplementary Material with Supplementary Figures, Supplementary Tables and their legends

@@
 
+```{r fig-norm-tumor, fig.width = figur::convert(suppl_paper_fig$norm_tumor, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$norm_tumor, to = 'in')$h, fig.cap = 'Expression of collagen pathway genes in the normal prostate and prostate cancer tissue.'}
+
+suppl_paper_fig$norm_tumor$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-norm-tumor). Expression of collagen pathway genes in the normal prostate and prostate cancer tissue.__
+
+_Differences in $log_2$-transformed expression of `r nrow(globals$genes_interest)` genes related to collagen metabolism between donor-matched pairs of the prostate cancer and benign tissue were assessed by paired T test with Cohen's effect size statistic.
+P values were corrected for multiple testing with the false discovery rate (FDR) method._
+
+_(A) Numbers of significantly up- and downregulated genes in the tumor tissue as compared with the benign tissue in the investigated cohorts presented in Venn plots.
+Genes regulated in at least two cohorts are listed next to the plots._
+
+_(B) Gene expression regulation significance and effect size presented in Volcano plots.
+Points represent single genes.
+Significantly regulated genes are labeled with their symbols.
+Number of analyzed tissue pairs are displayed in the plot captions._
+
+\newpage
+
+```{r fig-clust-dev, fig.width = figur::convert(suppl_paper_fig$clust_dev, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$clust_dev, to = 'in')$h, fig.cap = 'Semi-supervised clustering of prostate cancer samples in respect to expression of the collagen pathway genes.'}
+
+suppl_paper_fig$clust_dev$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-clust-dev). Semi-supervised clustering of prostate cancer samples in respect to expression of the collagen pathway genes.__
+
+_Tumor samples in the TCGA training cohort were clustered in respect to normalized, $log_2$-transformed expression levels of the collagen pathway genes with the PAM (partition around medoids) algorithm with Manhattan distance metric.
+Three clusters were defined: Collagen^low^, Collagen^intermediate^ and Collagen^high^.
+Assignment of the tumor samples from the training GSE16560, GSE40272, GSE70768 and GSE70769 collectives to the collagen clusters was accomplished by an inverse distance-weighted k-nearest neighbor classifier._
+
+_(A) Comparison of performance of several clustering algorithms (PAM: partition around medoids, HCL: hierarchical clustering/Ward D2 and KMEANS) and distance metrics (Euclidean, Manhattan, sum-of-squares and cosine) in the training TCGA cohort was measured by the amount of 'explained' clustering variance (ratio of the between-cluster sum of squares to the total sum of squares) and rate of correct cluster assignment in 10-fold cross-validation (CV).
+Number of complete observations and the clustering factors are shown next to the plot._
+
+_(B) Determination of the cluster number for PAM/Manhattan distance algorithm in the TCGA cohort by the bend of the within-cluster sum of squares (WSS) (B) and the peak of the mean silhouette statistic (C)._
+
+_(C) Distribution of the collagen clusters (left panel) and performance of the PAM/Manhattan distance algorithm in semi-supervised clustering in the training cohort and the test collectives measured by the 'explained' clustering variance statistic (right panel).
+Numbers of complete observations are presented in the plot axes.._
+
+\newpage
+
+```{r fig-clusters, fig.width = figur::convert(suppl_paper_fig$clusters, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$clusters, to = 'in')$h, fig.cap = 'Expression of the collagen pathway genes in the collagen clusters of prostate cancers.'}
+
+suppl_paper_fig$clusters$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-clusters). Expression of the collagen pathway genes in the collagen clusters of prostate cancers.__
+
+_Tumor samples were subjected to semi-supervised clustering in respect to the collagen pathway gene expression with the PAM algorithm with Manhattan distance metric (training: TCGA, test cohorts: GSE16560, GSE40272, GSE70768 and GSE70769).
+Statistical significance of differences in expression of normalized, $log_2$-transformed expression levels (Z-scores) between the collagen clusters were investigated by Kruskal-Wallis test with $\eta^2$ effect statistic.
+P values were corrected for multiple testing with the false discovery rate method.
+Mean expression Z scores in the Collagen^low^, Collagen^intermediate^ (int) and Collagen^high^ (hi) clusters are presented as thick lines.
+The tinted ribbons represent 2$\times$ SEM (standard error of the mean) intervals.
+The collagen pathway genes were grouped according to their biological function as proline synthesis pathway, extracellular matrix (ECM) component and ECM processing protein genes.
+Significantly regulated genes are highlighted in bold.
+Numbers of samples in the clusters are indicated in the plot captions._
+
+\newpage
+
+```{r fig-mcp-infil, fig.width = figur::convert(suppl_paper_fig$mcp_infil, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$mcp_infil, to = 'in')$h, fig.cap = 'MCP Counter estimates of non-malignant cell content in the collagen clusters.'}
+
+suppl_paper_fig$mcp_infil$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-mcp-infil). MCP Counter estimates of non-malignant cell content in the collagen clusters.__
+
+_Counts of non-malignant cells (CAF: cancer-associated fibroblast, EC: endothelial cells, T cells, CD8^+^ T cells, NK cells: natural killer cells, B cells, macrophages, monocytes, neutrophils and mDC: myeloid dendritic cells) in the prostate cancer samples were predicted by the MCP counter algorithm.
+Differences in cell content between the Collagen^low^, Collagen^intermediate^ (in) and Collagen^high^ (hi) clusters were analyzed with Kruskal-Wallis test with $\eta^2$ effect size statistic.
+P values were corrected for multiple testing with the false discovery rate method.
+Means of normalized cell counts in the collagen clusters are depicted as thick lines.
+Tinted ribbons represent the 2 $\times$ SEM (standard error of the mean) intervals.
+Effect sizes and p values are shown in the Y axes.
+Significant effects are highlighted in bold.
+Numbers of samples assigned to the clusters are indicated in the plot captions._
+
+\newpage
+
+```{r fig-infil-mcp, fig.width = 7.086614166, fig.height = 9.055118101, fig.cap = 'Counts of cancer-associated fibroblast and endothelial cell in tumor samples in the  collagen clusters predicted by the MCP counter algorithm.'}
+
+suppl_paper_fig$infil_mcp$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-infil-mcp). Counts of cancer-associated fibroblast and endothelial cell in tumor samples in the  collagen clusters predicted by the MCP counter algorithm.__
+
+_Counts of cancer associated fibroblasts (CAF, A) and endothelial cells (EC, B) in collagen^low^, collagen^intermediate^ (int) and collagen^high^ (hi) prostate cancers were predicted by the MCP counter algorithm.
+Differences between the collagen clusters were assessed by Kruskal-Wallis test with $\eta^2$ effect size statistic.
+Cell counts were presented in violin plots with single observations visualized as points.
+Red diamonds and whiskers represent medians with interquartile ranges.
+Effect sizes and p values are displayed in the plot captions.
+Numbers of samples in the clusters are indicated in the X axes._
+
+\newpage
+
+```{r fig-infil-xcell, fig.width = figur::convert(suppl_paper_fig$infil_xcell, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$infil_xcell, to = 'in')$h, fig.cap = 'Fractions of cancer-associated fibroblast and endothelial cell in tumor samples in the  collagen clusters predicted by the xCell algorithm.'}
+
+suppl_paper_fig$infil_xcell$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-infil-xcell). Fractions of cancer-associated fibroblast and endothelial cell in tumor samples in the  collagen clusters predicted by the xCell algorithm.__
+
+_Fractions of cancer associated fibroblasts (CAF, A) and endothelial cells (EC, B) in the malignant tissue of collagen^low^, collagen^intermediate^ (int) and collagen^high^ (hi) prostate cancers were predicted by the xCell counter algorithm.
+Differences between the collagen clusters were assessed by Kruskal-Wallis test with $\eta^2$ effect size statistic.
+Cell counts were presented in violin plots with single observations visualized as points.
+Red diamonds and whiskers represent medians with interquartile ranges.
+Effect sizes and p values are displayed in the plot captions.
+Numbers of samples in the clusters are indicated in the X axes._
+
+\newpage
+
+```{r fig-top-gsva, fig.width = figur::convert(suppl_paper_fig$top_gsva, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$top_gsva, to = 'in')$h, fig.cap = 'Gene set variation analysis results for representative Reactome pathway gene signatures differentiating between the collagen clusters.'}
+
+suppl_paper_fig$top_gsva$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-top-gsva). Gene set variation analysis results for representative Reactome pathway gene signatures differentiating between the collagen clusters.__
+
+_Gene set variation analysis of the collagen clusters was performed with the Reactome pathway gene signatures.
+Statistical significance for differences in signature single sample gene set enrichment analysis scores (ssGSEA) between the collagen clusters was determined by one-way ANOVA with $\eta^2$ effect size statistic.
+P values were corrected for multiple testing with the false discovery rate method.
+Common regulated signatures were defined as signatures significantly regulated with at least moderate effect size ($\eta^2\geq$ 0.6) in at least four cohorts (Supplementary Table S\@ref(tab:tab-gsva)).
+Mean ssGSEA scores in the clusters and cohorts for the top five strongest regulated signatures for each functional category (extracellular matrix [ECM], collagen; adhesion, motility, scavenging; angiogenesis, development, hemostasis; growth factor [GF] signaling; immune signaling; G protein coupled receptor, WNT, NOTCH signaling; metabolism, transport, translation) are presented in a heat map._
+
+\newpage
+
+```{r fig-dge-numbers, fig.width = figur::convert(suppl_paper_fig$dge_numbers, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$dge_numbers, to = 'in')$h, fig.cap = 'Percentages of the analyzed transcriptome significantly up- and downregulated in the collagen high and collagen low clusters.'}
+
+suppl_paper_fig$dge_numbers$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-dge-numbers). Percentages of the analyzed transcriptome significantly up- and downregulated in the collagen high and collagen low clusters.__
+
+_Genes significantly differentially regulated in the collagen^high^ (hi) and collagen^intermediate^ (int) clusters as compared with collagen^low^ prostate cancers were identified by false discovery rate corrected ANOVA (pFDR < 0.05) and linear modeling (fold-regulation > 1.25).
+Percentages of significantly up- and downregulated genes are presented in bar plots.
+Total numbers of analyzed genes are indicated in the Y axes._
+
+\newpage
+
+```{r fig-dge-top-int, fig.width = figur::convert(suppl_paper_fig$dge_top_int, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$dge_top_int, to = 'in')$h, fig.cap = 'Top strongest differentially expressed genes in the collagen intermediate cluster as compared with collagen low cancers.'}
+
+suppl_paper_fig$dge_top_int$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-dge-top-int). Top strongest differentially expressed genes in the collagen intermediate cluster as compared with collagen low cancers.__
+
+_$log_2$ fold-regulation estimates of differences in expression between collagen^intermediate^ (int) and collagen^low^ cancers with 95% confidence intervals for top 20 strongest up- and downregulated genes were presented in Forest plots.
+Numbers of samples in the clusters are displayed in the plot captions._
+
+\newpage
+
+```{r fig-dge-top-hi, fig.width = figur::convert(suppl_paper_fig$dge_top_hi, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$dge_top_hi, to = 'in')$h, fig.cap = 'Top strongest differentially expressed genes in the collagen high cluster as compared with collagen low cancers.'}
+
+suppl_paper_fig$dge_top_hi$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-dge-top-hi). Top strongest differentially expressed genes in the collagen high cluster as compared with collagen low cancers.__
+
+_$log_2$ fold-regulation estimates of differences in expression between collagen^high^ (high) and collagen^low^ cancers with 95% confidence intervals for top 20 strongest up- and downregulated genes were presented in Forest plots.
+Numbers of samples in the clusters are displayed in the plot captions._
+
+\newpage
+
+```{r fig-reaction-counts, fig.width = figur::convert(suppl_paper_fig$reaction_counts, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$reaction_counts, to = 'in')$h, fig.cap = 'Numbers of significantly activated and inhibited biochemical reactions predicted for the collagen high and collagen intermediate clusters.'}
+
+suppl_paper_fig$reaction_counts$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-reaction-counts). Numbers of significantly activated and inhibited biochemical reactions predicted for the collagen high and collagen intermediate clusters.__
+
+_Modulation of Recon2 model metabolic reactions in the collagen^high^ (hi) or collagen^intermediate^ (hi) as compared with collagen^low^ tumors was predicted by the BiGGR and biggrExtra algorithms based on whole genome differential gene expression estimates.
+Statistical significance was determined by Monte Carlo simulation with n = 1000 draws from distributions of the gene expression regulation standard errors.
+P values were corrected for multiple testing with the false discovery rate method.
+Percentages of predicted significantly activated and inhibited metabolic reactions are shown in bar plots.
+Total reaction numbers are displayed in the plot captions._
+
+\newpage
+
+```{r fig-faox, fig.width = figur::convert(suppl_paper_fig$faox, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$faox, to = 'in')$h, fig.cap = 'Activity of extracellular transport reactions predicted for the collagen high and collagen intermediate clusters.'}
+
+suppl_paper_fig$faox$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-faox). Activity of extracellular transport reactions predicted for the collagen high and collagen intermediate clusters.__
+
+_Regulation of the fatty acid oxidation (FAOX) Recon subsystem reactions (total reactions: n = `r (count(meta$models[[1]][[1]]) %>% filter(subsystem == 'Fatty acid oxidation') %>% .$n_total)[1]`) in the collagen^high^ (hi, A) or collagen^intermediate^ (hi, B) as compared with collagen^low^ tumors was predicted by the BiGGR and biggrExtra algorithms based on whole genome differential gene expression estimates.
+Statistical significance was determined by Monte Carlo simulation with n = 1000 draws from distributions of the gene expression regulation standard errors.
+P values were corrected for multiple testing with the false discovery rate method.
+Estimates of $log_2$ fold-regulation of reaction activity are presented in bar plots.
+Activity status is color coded.
+Numbers of significantly activated and inhibited reactions are displayed in the plot captions._
+
+\newpage
+
+```{r fig-extrans, fig.width = figur::convert(suppl_paper_fig$extrans, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$extrans, to = 'in')$h, fig.cap = 'Activity of extracellular transport reactions predicted for the collagen high and collagen intermediate clusters.'}
+
+suppl_paper_fig$extrans$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-extrans). Activity of extracellular transport reactions predicted for the collagen high and collagen intermediate clusters.__
+
+_Regulation of the extracellular transport (ExTrans) Recon subsystem reactions (total reactions: n = `r (count(meta$models[[1]][[1]]) %>% filter(subsystem == 'Transport, extracellular') %>% .$n_total)[1]`) in the collagen^high^ (hi, A) or collagen^intermediate^ (hi, B) as compared with collagen^low^ tumors was predicted by the BiGGR and biggrExtra algorithms based on whole genome differential gene expression estimates.
+Statistical significance was determined by Monte Carlo simulation with n = 1000 draws from distributions of the gene expression regulation standard errors.
+P values were corrected for multiple testing with the false discovery rate method.
+Estimates of $log_2$ fold-regulation of reaction activity are presented in bar plots.
+Activity status is color coded.
+Numbers of significantly activated and inhibited reactions are displayed in the plot captions._
+
 \newpage

### 2023-08-07 analysis scripts/clinic.R
R code update
Inclusion of the GSE116918 cohort. Two-cluster solution for the collagen clusters of prostate carcinoma

@@ -73,7 +73,7 @@
   ## correlation
   
   coll_clinic$age_corr <- 
-    coll_clinic$analysis_tbl[c('GSE16560', 'GSE40272', 'GSE70768', 'tcga')] %>% 
+    coll_clinic$analysis_tbl[c("GSE16560", "GSE70768", "GSE116918", "tcga")] %>% 
     future_map(function(cohort) map(coll_clinic$responses, 
                                     ~c('age', .x)) %>% 
                  map_dfr(~correlate_variables(cohort, 
@@ -101,10 +101,11 @@
   ## correlation plots
   
   coll_clinic$age_plots <- 
-    list(data = coll_clinic$analysis_tbl[c("GSE16560", "GSE40272", "GSE70768", "tcga")], 
+    list(data = coll_clinic$analysis_tbl[names(coll_clinic$age_corr)], 
          test_results = coll_clinic$age_corr, 
-         point_color = globals$study_colors[c("GSE16560", "GSE40272", "GSE70768", "tcga")]) %>% 
-    pmap(plot_clinical_correlation)
+         point_color = globals$study_colors[names(coll_clinic$age_corr)]) %>% 
+    future_pmap(plot_clinical_correlation, 
+                .options = furrr_options(seed = TRUE))
   
 # Collagen genes and PSA an diagnosis: Spearman correlation -----
   
@@ -113,7 +114,7 @@
   ## correlation
   
   coll_clinic$psa_corr <- 
-    coll_clinic$analysis_tbl[c('GSE40272', 'GSE70768', 'GSE70769', 'tcga')] %>% 
+    coll_clinic$analysis_tbl[c("GSE70768", "GSE70769", "GSE116918", "tcga")] %>% 
     future_map(function(cohort) map(coll_clinic$responses, 
                                     ~c('psa_at_diagnosis', .x)) %>% 
                  map_dfr(~correlate_variables(cohort[.x] %>% 
@@ -142,10 +143,11 @@
   ## correlation plots
   
   coll_clinic$psa_plots <- 
-    list(data = coll_clinic$analysis_tbl[c("GSE40272", "GSE70768", "GSE70769", "tcga")], 
+    list(data = coll_clinic$analysis_tbl[names(coll_clinic$psa_corr)], 
          test_results = coll_clinic$psa_corr, 
-         point_color = globals$study_colors[c("GSE40272", "GSE70768", "GSE70769", "tcga")]) %>% 
-    pmap(plot_clinical_correlation)
[…]

### 2023-08-07 analysis scripts/clust_devel.R

@@ -56,7 +56,7 @@
     clust_dev$dists %>% 
     map(~hcluster(data = clust_dev$analysis_tbl$tcga, 
                   distance_method = .x,
-                  k = 3, 
+                  k = 2, 
                   hc_method = 'ward.D2'))
   
   ## K-MEANS
@@ -65,7 +65,7 @@
     clust_dev$dists %>% 
     map(~kcluster(data = clust_dev$analysis_tbl$tcga, 
                   distance_method = .x,
-                  k = 3, 
+                  k = 2, 
                   clust_fun = 'kmeans'))
   
   ## PAM
@@ -74,7 +74,7 @@
     clust_dev$dists %>% 
     map(~kcluster(data = clust_dev$analysis_tbl$tcga, 
                   distance_method = .x,
-                  k = 3, 
+                  k = 2, 
                   clust_fun = 'pam'))
 
 # Variances of the algorithms ------

### 2023-08-07 analysis scripts/collagen_cluster_clinic.R

@@ -32,10 +32,10 @@
                                      'psa_at_diagnosis' = 'AU'; 
                                      'gleason' = 'score'; 
                                      'collagen_score' = 'Collagen Score'")), 
-           test_type = ifelse(format == 'numeric', 'kruskal_etasq', 'cramer_v'), 
+           test_type = ifelse(format == 'numeric', 'cohen_d', 'cramer_v'), 
            plot_type = ifelse(format == 'numeric', 'violin', 'stack'))
   
-  ## variables by thier format: the vestor is used later for plot styling
+  ## variables by their format: the vector is used later for plot styling
   
   cs_cluster$variables <- cs_cluster$lexicon %>% 
     blast(format) %>% 
@@ -62,10 +62,12 @@
          left_join, by = 'patient_id') %>% 
     map(mutate, 
         clust_id = stri_replace(clust_id, fixed = 'Collagen ', replacement = ''), 
-        clust_id = factor(clust_id, c('low', 'int', 'hi')), 
+        clust_id = factor(clust_id, c('low', 'hi')), 
         gleason_factor = cut(gleason, 
                              c(-Inf, 6, 7, Inf), 
-                             c('5 - 6', '7', '8+')))
+                             c('5 - 6', '7', '8+'))) %>% 
+    map(map_dfc, 
+        function(x) if(is.factor(x)) droplevels(x) else x)
   
   ## vectors with variables for single analysis tables
   

### 2023-08-07 analysis scripts/collagen_cluster_infiltration.R

@@ -17,7 +17,7 @@
   
   insert_msg('Globals')
 
-  ## infiltration variables, skipping the cytotoxicity score
+  ## infiltration variables
   
   clust_infil$variables[c('xcell', 'mcp')] <- 
     infil[c("xcell_types", "mcp_counter_types")]
@@ -27,7 +27,8 @@
   clust_infil$assignment <- coll_clust$assignment %>% 
     map(mutate, 
         clust_id = stri_extract(clust_id, regex = 'low|int|hi'), 
-        clust_id = factor(clust_id, c('low', 'int', 'hi')))
+        clust_id = factor(clust_id, c('low', 'int', 'hi')), 
+        clust_id = droplevels(clust_id))
   
   clust_infil$analysis_tbl$mcp <- clust_infil$assignment
   clust_infil$analysis_tbl$xcell <- clust_infil$analysis_tbl$mcp
@@ -72,13 +73,13 @@
                  .options = furrr_options(seed = TRUE)) %>% 
       map(reduce, left_join, by = 'variable') %>% 
       map(set_names, 
-          c('variable', levels(clust_infil$analysis_tbl[[1]][[1]]$clust_id)))
-    
+          c('variable', levels(clust_infil$assignment[[1]]$clust_id)))
+
   }
   
-# Kruskal-Wallis test ------
+# Mann-Whitney test ------
   
-  insert_msg('Kruskal-Wallis test')
+  insert_msg('Mann-whitney test')
   
   for(i in names(clust_infil$analysis_tbl)) {
     
@@ -87,7 +88,7 @@
                                     variables = clust_infil$variables[[i]][clust_infil$variables[[i]] %in% names(.x)], 
                                     split_factor = 'clust_id', 
                                     what = 'eff_size',
-                                    types = 'kruskal_eta', 
+                                    types = 'wilcoxon_r', 
                                     ci = FALSE, 
                                     exact = FALSE, 
                                     pub_styled = TRUE, 

### 2023-08-07 analysis scripts/collagen_cluster_survival.R

@@ -75,7 +75,7 @@
   ## test
   
   clust_surv$test <- clust_surv$survfit_obj %>% 
-    map(map, surv_pvalue) %>% 
+    map(map, surv_pvalue, method = 'S1') %>% 
     map(compress, names_to = 'cohort') %>% 
     compress(names_to = 'comparison') %>% 
     mutate(comparison = factor(comparison, c('overall', 'hi_lo'))) %>% 
@@ -95,7 +95,7 @@
   insert_msg('KM plots')
 
   clust_surv$km_plots <- list(fit = clust_surv$survfit_obj$overall, 
-                              pval = clust_surv$km_p_labs, 
+                              pval = clust_surv$test$overall$significance, 
                               title = paste('Collagen clusters and survival,', 
                                             globals$study_labels[names(clust_surv$analysis_tbl)]), 
                               xlab = c('Overall survival, months', 

### 2023-08-07 analysis scripts/collagen_clustering.R

@@ -23,19 +23,17 @@
   
   ## optimal clustering structure
 
-  coll_clust$clust_obj$tcga <- clust_dev$algos$pam_manhattan
+  coll_clust$clust_obj$tcga <- clust_dev$algos$pam_cosine
   
   ## renaming
   
   coll_clust$clust_obj$tcga$clust_assignment <- 
     coll_clust$clust_obj$tcga$clust_assignment  %>% 
     mutate(clust_id = car::recode(clust_id, 
-                                  "'1' = 'Collagen int'; 
-                                  '2' = 'Collagen hi'; 
-                                  '3' = 'Collagen low'"), 
+                                  "'1' = 'Collagen hi'; 
+                                  '2' = 'Collagen low'"), 
            clust_id = factor(clust_id, 
                              c('Collagen hi', 
-                               'Collagen int', 
                                'Collagen low')))
   
   ## tables used for the clustering structure assignment
@@ -69,10 +67,8 @@
   set.seed(12345)
   
   coll_clust$semi_clust <- 
-    list(newdata = coll_clust$clust_tbl[c('GSE16560', 
-                                          'GSE40272', 
-                                          'GSE70768', 
-                                          'GSE70769')]) %>% 
+    list(newdata = coll_clust$clust_tbl[c("GSE16560", "GSE70768", 
+                                          "GSE70769", "GSE116918")]) %>% 
     pmap(adapt_knn, 
          object = coll_clust$clust_obj$tcga, 
          type = 'propagation', 
@@ -80,14 +76,10 @@
          resolve_ties = TRUE, 
          max_kNN = 27)
   
-  coll_clust$clust_obj[c('GSE16560', 
-                          'GSE40272', 
-                          'GSE70768', 
-                          'GSE70769')] <- 
-    coll_clust$semi_clust[c('GSE16560', 
-                            'GSE40272', 
-                            'GSE70768', 
-                            'GSE70769')] %>% 
+  coll_clust$clust_obj[c("GSE16560", "GSE70768", 
+                         "GSE70769", "GSE116918")] <- 
+    coll_clust$semi_clust[c("GSE16560", "GSE70768", 
+                            "GSE70769", "GSE116918")] %>% 
     map(~.x$object)
   
   coll_clust$clust_obj <- coll_clust$clust_obj[names(coll_clust$clust_tbl)]
@@ -123,8 +115,8 @@
     scale_fill_manual(values = c(test = 'steelblue', 
                                  training = 'darkolivegreen4'), 
                       name = 'Cohort') + 
[…]

### 2023-08-07 analysis scripts/collagen_score.R

@@ -24,7 +24,7 @@
   
   coll_score$variables <- globals$genes_interest$gene_symbol
   
-  ## analysis tables: survival informaiton
+  ## analysis tables: survival information
   
   coll_score$analysis_tbl <- study_data %>% 
     map(~.x$expression) %>% 

### 2023-08-07 analysis scripts/correlation.R

@@ -157,34 +157,36 @@
 # Traditional scatter plots for each gene pair -------
   
   insert_msg('Gene pair scatter plots')
+  
+  ## not done for the while: memory issue!
 
-  for(i in names(corr$analysis_tbl)) {
+  #for(i in names(corr$analysis_tbl)) {
     
-    corr$plots[[i]] <- 
-      list(variables = map2(corr$test_result[[i]]$variable1, 
-                            corr$test_result[[i]]$variable2, c), 
-           plot_title = corr$test_results[[i]]$plot_title, 
-           plot_subtitle = corr$test_results[[i]]$plot_cap, 
-           x_lab = paste0('<em>', corr$test_results[[i]]$variable1, 
-                          '</em>, log<sub>2</sub> expression'), 
-           y_lab = paste0('<em>', corr$test_results[[i]]$variable2, 
-                          '</em>, log<sub>2</sub> expression')) %>% 
-      future_pmap(plot_correlation, 
-                  data = corr$analysis_tbl[[i]], 
-                  type = 'correlation', 
-                  point_color = globals$study_colors[[i]], 
-                  cust_theme = globals$common_theme, 
-                  .options = furrr_options(seed = TRUE, 
-                                           packages = c('tidyverse', 
-                                                        'exda'))) %>% 
-      map(~.x + 
-            theme(plot.tag = element_blank(), 
-                  plot.title = element_markdown(), 
-                  axis.title.x = element_markdown(), 
-                  axis.title.y = element_markdown())) %>% 
-      set_names(corr$test_results[[i]]$pair)
+   # corr$plots[[i]] <- 
+    #  list(variables = map2(corr$test_result[[i]]$variable1, 
+     #                       corr$test_result[[i]]$variable2, c), 
+      #     plot_title = corr$test_results[[i]]$plot_title, 
+       #    plot_subtitle = corr$test_results[[i]]$plot_cap, 
+        #   x_lab = paste0('<em>', corr$test_results[[i]]$variable1, 
+         #                 '</em>, log<sub>2</sub> expression'), 
+          # y_lab = paste0('<em>', corr$test_results[[i]]$variable2, 
+           #               '</em>, log<sub>2</sub> expression')) %>% 
+  #    future_pmap(plot_correlation, 
+   #               data = corr$analysis_tbl[[i]], 
+    #              type = 'correlation', 
+     #             point_color = globals$study_colors[[i]], 
+      #            cust_theme = globals$common_theme, 
[…]

### 2023-08-07 analysis scripts/mri.R

@@ -14,16 +14,19 @@
   insert_msg('Analysis globals')
   
   ## analysis table
-  ## Warning: at the moment only simulated data with the mean and SD
-  ## from the GraphPad plot
+
+  mri$analysis_tbl <- read_xlsx('./data/MRT Langer.xlsx')
   
-  mri$analysis_tbl <- 
-    tibble(patient_id = paste0('pat_', rep(1:10, 2)), 
-           adc = c(rnorm(10, mean = 700, sd = 150), 
-                   rnorm(10, mean = 1240, sd = 250)), 
-           tissue = c(rep('Tumor', 10), rep('Benign', 10))) %>% 
-    mutate(tissue = factor(tissue, c('Tumor', 'Benign')))
-  
+  mri$analysis_tbl <- mri$analysis_tbl[1:10, ] %>% 
+    transmute(patient_id = as.character(Patientennummer), 
+              date = `Datum des MRT`, 
+              tumor = `SI Avg Tumor`, 
+              benign = `SI Avg kontralat gesunde Seite`) %>% 
+    pivot_longer(cols = all_of(c('tumor', 'benign')), 
+                 names_to = 'tissue', 
+                 values_to = 'adc') %>%
+    mutate(tissue = factor(tissue, c('tumor', 'benign')))
+
 # Descriptive stats ------
   
   insert_msg('Descriptive stats')
@@ -64,8 +67,8 @@
                   plot_title = 'Tissue diffusion capacity', 
                   plot_subtitle = mri$test$plot_cap, 
                   y_lab = expression('mean ADC, 10'^6 * ' mm'^2 * '/s')) + 
-    scale_fill_manual(values = c(Tumor = 'coral3', 
-                                 Benign = 'darkolivegreen4'), 
+    scale_fill_manual(values = c(tumor = 'coral3', 
+                                 benign = 'darkolivegreen4'), 
                       name = '') + 
     theme(axis.title.x = element_blank(), 
           plot.tag = element_blank())

### 2023-08-07 analysis scripts/normal_tumor.R

@@ -22,8 +22,7 @@
   ## compatible with paired T test
   
   norm_tumor$analysis_tbl <- 
-    study_data[c('GSE40272', 
-                 'GSE70768', 
+    study_data[c('GSE70768', 
                  'tcga')] %>% 
     map(~.x$expression) %>% 
     map(select, 
@@ -103,20 +102,11 @@
     map(map, ~.x$variable) %>% 
     transpose
   
-  ## transcripts regulated in at least two cohorts
+  ## transcripts regulated in both cohorts
   
-  norm_tumor$cmm_sets <- names(norm_tumor$test_results) %>% 
-    combn(m = 2, simplify = FALSE)
-  
-  for(i in names(norm_tumor$significant)) {
-    
-    norm_tumor$common[[i]] <- norm_tumor$cmm_sets %>% 
-      map(~norm_tumor$significant[[i]][.x]) %>% 
-      map(reduce, intersect) %>% 
-      reduce(union)
-    
-  }
-  
+  norm_tumor$common <- norm_tumor$significant %>% 
+    map(reduce, intersect)
+
 # volcano plots -------
   
   insert_msg('Volcano plots')
@@ -137,7 +127,8 @@
                       shape = 16) + 
            geom_text_repel(aes(label = plot_lab), 
                            size = 2.5, 
-                           fontface = 'italic') + 
+                           fontface = 'italic', 
+                           show.legend = FALSE) + 
            scale_color_manual(values = c(upregulated = 'firebrick', 
                                          downregulated = 'steelblue', 
                                          ns = 'gray60'), 
@@ -162,6 +153,7 @@
                         'Downregulated transcripts'), 
          text = norm_tumor$common) %>% 
     pmap(plot_venn, 
+         show_text = FALSE, 
          colors = globals$study_colors[names(norm_tumor$analysis_tbl)], 
          plot_subtitle = 'Tumor vs benign', 
          fct_per_line = 1, 
@@ -214,7 +206,7 @@
     summarise(estimate = mean(estimate), .by = variable) %>% 
     arrange(estimate)
   
-  ## axis labels: bold if significant
+  ## axis labels: all are significant, no extra labeling!
   
   norm_tumor$ribbon_plots$ax_labs <- norm_tumor$test_results %>% 
     map(filter, variable %in% norm_tumor$ribbon_plots$variables) %>% 
@@ -223,9 +215,10 @@
         significance = ifelse(stri_detect(significance, fixed = 'ns'), 
[…]

### 2023-08-07 analysis scripts/survival_cut.R

@@ -172,83 +172,86 @@
          y = list(expression('-log'[10] * ' pFDR, Mentel-Henszel test'), 
                   expression('-log'[10] * ' p raw, Mentel-Henszel test'))) %>% 
     pmap(function(x, y) list(data = surv_cut$summary_tbl, 
-                            plot_title = paste('Survival markers', 
-                                               globals$study_labels[names(surv_cut$summary_tbl)], 
-                                               sep = ', '), 
-                            plot_subtitle = map2(c('OS', rep('RFS', 4)), 
-                                                 surv_cut$n_tags, 
-                                                 paste, sep = ', ')) %>% 
-          pmap(plot_signifcant, 
-               p_variable = x, 
-               label_variable = 'variable', 
-               top_significant = 30, 
-               x_lab = y, 
-               cust_theme = globals$common_theme) %>% 
-          map(~.x + 
-                theme(axis.text.y = element_markdown(), 
-                      plot.tag = element_blank()) + 
-                scale_y_discrete(labels = function(x) exchange(x, 
-                                                               surv_cut$var_labs, 
-                                                               value = 'html_label')) + 
-                scale_fill_manual(values = c(ns = 'gray60', 
-                                             significant = 'coral3'), 
-                                  labels = c(ns = 'ns', 
-                                             significant = 'p < 0.05'))))
+                             plot_title = paste('Survival markers', 
+                                                globals$study_labels[names(surv_cut$summary_tbl)], 
+                                                sep = ', '), 
+                             plot_subtitle = map2(c('OS', rep('RFS', 4)), 
+                                                  surv_cut$n_tags, 
+                                                  paste, sep = ', ')) %>% 
+           future_pmap(plot_signifcant, 
+                       p_variable = x, 
+                       label_variable = 'variable', 
+                       top_significant = 30, 
+                       x_lab = y, 
+                       cust_theme = globals$common_theme, 
+                       .options = furrr_options(seed = TRUE)) %>% 
+           map(~.x + 
+                 theme(axis.text.y = element_markdown(), 
+                       plot.tag = element_blank()) + 
[…]

### 2023-08-07 analysis scripts/survival_multi.R

@@ -54,23 +54,23 @@
   
   surv_multi$formulas$multi <- 
     list(GSE16560 = Surv(vitality_fup, death) ~ collagen_score + age + gleason_factor, 
-         GSE40272 = Surv(relapse_fup, relapse) ~ collagen_score + age + gleason_factor, 
          GSE70768 = Surv(relapse_fup, relapse) ~ collagen_score + age + gleason_factor, 
          GSE70769 = Surv(relapse_fup, relapse) ~ collagen_score + gleason_factor, 
+         GSE116918 = Surv(relapse_fup, relapse) ~ collagen_score + age + gleason_factor,
          tcga = Surv(relapse_fup, relapse) ~ collagen_score + age + gleason_factor)
   
   surv_multi$formulas$uni <- 
     list(GSE16560 = Surv(vitality_fup, death) ~ collagen_score, 
-         GSE40272 = Surv(relapse_fup, relapse) ~ collagen_score, 
          GSE70768 = Surv(relapse_fup, relapse) ~ collagen_score, 
          GSE70769 = Surv(relapse_fup, relapse) ~ collagen_score, 
+         GSE116918 = Surv(relapse_fup, relapse) ~ collagen_score, 
          tcga = Surv(relapse_fup, relapse) ~ collagen_score)
   
   surv_multi$formulas$clinical <- 
     list(GSE16560 = Surv(vitality_fup, death) ~ age + gleason_factor, 
-         GSE40272 = Surv(relapse_fup, relapse) ~ age + gleason_factor, 
          GSE70768 = Surv(relapse_fup, relapse) ~ age + gleason_factor, 
          GSE70769 = Surv(relapse_fup, relapse) ~ gleason_factor, 
+         GSE116918 = Surv(relapse_fup, relapse) ~ age + gleason_factor, 
          tcga = Surv(relapse_fup, relapse) ~ age + gleason_factor)
   
   ## model colors and labels

### 2023-08-07 analysis scripts/survival_uni.R

@@ -83,9 +83,9 @@
   
   surv_cox$formulas <- 
     c(GSE16560 = 'Surv(vitality_fup, death)', 
-      GSE40272 = 'Surv(relapse_fup, relapse)', 
-      GSE70768 = 'Surv(relapse_fup, relapse)',
-      GSE70769 = 'Surv(relapse_fup, relapse)', 
+      GSE70768 = 'Surv(relapse_fup, relapse)', 
+      GSE70769 = 'Surv(relapse_fup, relapse)',
+      GSE116918 = 'Surv(relapse_fup, relapse)', 
       tcga = 'Surv(relapse_fup, relapse)') %>% 
     map(function(response) surv_cox$variables %>% 
           map(~paste(response, .x, sep = ' ~ ')) %>% 
@@ -217,7 +217,7 @@
            labs(title = paste('Inference,', y), 
                 subtitle = paste0(resp, ', univariable Cox modeling, ', z), 
                 x = paste('HR \u00B1 95%CI, normalized')))
-  
+
 # END -----
   
   plan('sequential')

### 2023-08-07 analysis.R

@@ -8,6 +8,7 @@
   library(trafo)
   library(rlang)
   library(stringi)
+  library(readxl)
 
   library(ggtext)
   library(ggrepel)
@@ -62,9 +63,26 @@
   ## and infiltration
   
   c('./analysis scripts/collagen_score.R', ## development and properties of the Collagen Score
-    './analysis scripts/collagen_score_plots.R', ## graphs for features of the Collagen Score
-    './analysis scripts/clinic.R', ## Collagen genes and Collagen Score as a function of clinical features
-    './analysis scripts/survival_uni.R', ## Uni-variable Cox modeling of survival
+    './analysis scripts/collagen_score_plots.R') %>% ## graphs for features of the Collagen Score
+    source_all(message = TRUE, crash = TRUE) %>% 
+    print
+  
+  if(file.exists('./cache/coll_clinic.RData')) {
+    
+    insert_msg('Loading cached clinical association results')
+  
+    load('./cache/coll_clinic.RData')
+    
+  } else {
+    
+    ## Collagen genes and Collagen Score as a function of clinical features
+    
+    c('./analysis scripts/clinic.R') %>% 
+      source_all(message = TRUE, crash = TRUE)
+    
+  }
+  
+  c('./analysis scripts/survival_uni.R', ## Uni-variable Cox modeling of survival
     './analysis scripts/survival_cut.R', ## survival in dichotomous collagen gene and Collagen Score strata
     './analysis scripts/survival_multi.R', ## multi-variable survival modeling, Collagen Score
     './analysis scripts/collagen_score_infiltration.R') %>% ## correlation of the Collagen Score and non-malignant cell infiltration)
@@ -75,7 +93,7 @@
   
   c('./analysis scripts/clust_devel.R', ## development of the collagen clusters
     './analysis scripts/collagen_clustering.R', ## semi-supervised clustering
-    './analysis scripts/collagen_cluster_survival.R', ## diffferences in survival between the Collagen Clusters
+    './analysis scripts/collagen_cluster_survival.R', ## differences in survival between the Collagen Clusters
     './analysis scripts/collagen_cluster_clinic.R', ## differences in clinics and Collagen score between the clusters
     './analysis scripts/collagen_cluster_infiltration.R') %>% ## infiltration in the clusters
     source_all(message = TRUE, crash = TRUE) %>% 

### 2023-08-07 clearing scripts/data_clearing_geo.R

@@ -2,185 +2,204 @@
 
   insert_head()
 
-# GSE40272: merging the sub-experiments ----
+# container ------
 
-  insert_msg('GSE40272: merging the sub-experiments')
-
-  geo_data$GSE40272$expression <- geo_data[c('GSE40272-GPL15971', 
-                                             'GSE40272-GPL15972', 
-                                             'GSE40272-GPL15973', 
-                                             'GSE40272-GPL9497')] %>% 
-    map(function(x) x$expression) %>% 
-    map_dfr(mutate, 
-            dfs_months = as.numeric(dfs_months))
-  
-  geo_data$GSE40272$annotation <- geo_data$`GSE40272-GPL15971`$annotation
+  geo_cleared <- list()
 
 # integrating the expression as geometric mean of multiple probes -----
 
   insert_msg('Integrating the gene expression')
   
-  study_data <- geo_data[c('GSE16560', 
-                           'GSE40272', 
-                           'GSE70768', 
-                           'GSE70769')] %>% 
+  geo_cleared <- geo_data %>% 
     map(function(x) integrate_expression(expression_tbl = x$expression, 
                                          annotation_tbl = x$annotation, 
                                          gene_identifier = 'symbol', 
-                                         .parallel = T))
+                                         .parallel = TRUE))
   
 # clearing the GSE16560 study metadata ----
   
   insert_msg('Clearing the GSE16560 metadata')
   
-  study_data$GSE16560$expression <- study_data$GSE16560$expression %>% 
-    mutate(death = ifelse(vitality == 'Alive', 0, 1), 
-           vitality_fup = followup_months, 
-           patient_id = provided_ID)
+  geo_cleared$GSE16560$clinic <- geo_data$GSE16560$clinic %>% 
+    transmute(sample_ID = geo_accession, 
+              patient_id = title, 
+              gleason = extract_pfield(characteristics_ch1, 'numeric'), 
+              major_gleason = extract_pfield(characteristics_ch1.1, 'numeric'), 
+              minor_gleason = extract_pfield(characteristics_ch1.2, 'numeric'), 
+              cancer_percent = extract_pfield(characteristics_ch1.5, 'numeric'), 
+              age = extract_pfield(characteristics_ch1.6, 'numeric'), 
+              vitality = extract_pfield(characteristics_ch1.8, 'factor', 
+                                        c('Alive', 'Dead')), 
+              death = ifelse(vitality == 'Alive', 0, 1), 
[…]

### 2023-08-07 clearing scripts/data_clearing_tcga.R

@@ -2,6 +2,10 @@
 
   insert_head()
 
+# container ------
+
+  tcga_cleared <- list()
+
 # merging the clinical and expression data ------
 
   insert_msg('Merging the clinical and expression data')
@@ -58,7 +62,7 @@
   
   insert_msg('Integrating the TCGA data')
   
-  study_data$tcga <- integrate_expression(expression_tbl = tcga_data$expression, 
+  tcga_cleared <- integrate_expression(expression_tbl = tcga_data$expression, 
                                           annotation_tbl = tcga_data$annotation %>% 
                                             set_names(c('probe_ID', 'symbol', 'protein')), 
                                           gene_identifier = 'symbol', 
@@ -69,13 +73,13 @@
   
   insert_msg('Duplicate removal')
   
-  study_data$tcga$expression <- study_data$tcga$expression %>% 
+  tcga_cleared$expression <- tcga_cleared$expression %>% 
     dlply(.(tissue), 
           rm_duplicates, 
           patient_identifier = 'patient_id', 
           reposit_identifier = 'entity_submitter_id') %>% 
     reduce(rbind)
-
+  
 # END ----
   
   insert_msg()

### 2023-08-07 clearing scripts/infiltration.R

@@ -13,8 +13,7 @@
   
   infil$analysis_tbl <- study_data %>% 
     map(~.x$expression) %>% 
-    map(function(x) if('tissue' %in% names(x)) filter(x, tissue == 'tumor') else x)
-  
+    map(filter, tissue == 'tumor')
   
   infil$variables <- study_data %>% 
     map(~.x$annotation) %>% 
@@ -30,13 +29,8 @@
     map(column_to_rownames, 'patient_id') %>%
     map(t)
   
-  infil$analysis_tbl[c('GSE16560', 
-                       'GSE40272', 
-                       'GSE70768', 
-                       'GSE70769')] <- infil$analysis_tbl[c('GSE16560', 
-                                                            'GSE40272', 
-                                                            'GSE70768', 
-                                                            'GSE70769')] %>% 
+  infil$analysis_tbl[c("GSE16560", "GSE70768", "GSE70769", "GSE116918")] <- 
+    infil$analysis_tbl[c("GSE16560", "GSE70768", "GSE70769", "GSE116918")] %>% 
     map(~2^.x)
   
   infil$analysis_tbl$tcga <- 2^infil$analysis_tbl$tcga - 1

### 2023-08-07 dge scripts/dge_go.R

@@ -27,14 +27,10 @@
   
   plan('multisession')
   
-  dge_go$test$int <- dge$entrez_collagen_int %>% 
+  dge_go$test <- dge$dge_entrez %>% 
     future_map(goana, 
                .options = furrr_options(seed = TRUE))
-  
-  dge_go$test$high <- dge$entrez_collagen_hi %>% 
-    future_map(goana, 
-               .options = furrr_options(seed = TRUE))
-  
+
   plan('sequential')
   
 # formatting of the results: BD GO and FDR correction -----
@@ -44,29 +40,20 @@
   ## appending the result tables with the total number of genes
   ## and number of regulated genes
   
-  dge_go$test$int <- 
-    list(x = dge_go$test$int, 
-         y = dge$entrez_collagen_int, 
+  dge_go$test <- 
+    list(x = dge_go$test, 
+         y = dge$dge_entrez, 
          z = dge$annotation) %>% 
     pmap(function(x, y, z) x %>% 
            mutate(n_de = length(y), 
                   n_total = nrow(z)))
-  
-  dge_go$test$high <- 
-    list(x = dge_go$test$high, 
-         y = dge$entrez_collagen_hi, 
-         z = dge$annotation) %>% 
-    pmap(function(x, y, z) x %>% 
-           mutate(n_de = length(y), 
-                  n_total = nrow(z)))
-  
+
   ## effect sizes of the enrichment: odds ratio
   ## this is computed with the following formula
   ## (n(DE, GO)/n(GO))/(n(DE)/n(total))
   
   dge_go$test <- dge_go$test %>% 
-    map(map, 
-        mutate, 
+    map(mutate, 
         OR = (DE/N)/(n_de/n_total))
   
   ## multiple testing adjustment
@@ -74,19 +61,19 @@
   ## effect size (Cohen 1988): at least small
   
   dge_go$test <- dge_go$test %>% 
-    map(~map(.x, rownames_to_column, 'go_id') %>% 
-          map(mutate, 
-              p_adjusted = p.adjust(P.DE), 
-              eff_size = interpret_oddsratio(OR, rules = 'cohen1988')) %>% 
-          map(filter, 
-              Ont == 'BP') %>% 
-          map(as_tibble))
-  
+    map(rownames_to_column, 'go_id') %>% 
+    map(mutate, 
+        p_adjusted = p.adjust(P.DE), 
[…]

### 2023-08-07 dge scripts/dge_pathview.R

@@ -20,82 +20,61 @@
   
   ## KEGG IDs of the regulated samples
   
-  dge_path$kegg_ID <- 
-    map2(dge_spia$test, 
-         dge_spia$common %>% 
-           reduce(union), 
-         function(data, name) data[[1]] %>% 
-           filter(Name %in% name) %>% 
-           .$ID)
-  
+  dge_path$kegg_ID <- dge_spia$test %>% 
+    map(filter, Name %in% unique(unlist(dge_spia$common))) %>% 
+    map(~.x$ID) %>% 
+    reduce(union)
+
   ## vectors of differentially regulated genes in at least one cohort
   
-  dge_path$collagen_genes[c('int', 'hi')] <- 
-    dge[c("dge_collagen_int", "dge_collagen_hi")] %>% 
-    map(map, ~.x$entrez_id) %>% 
-    map(reduce, union)
-  
+  dge_path$collagen_genes <- dge$dge_entrez %>% 
+    reduce(union)
+
   ## analysis tables
   
   dge_path$collagen_tbl <- dge$test_results %>% 
-    map(~.x$lm) %>% 
-    map(filter, 
-        level %in% c('Collagen hi', 'Collagen int')) %>% 
-    map(blast, level) %>% 
-    map(function(data) map2(data, 
-                            dge_path$collagen_genes, 
-                            ~filter(.x, 
-                                    entrez_id %in% .y))) %>% 
-    transpose %>% 
-    set_names(c('int', 'hi')) %>% 
-    map(compress, names_to = 'cohort') %>% 
-    map(select, cohort, entrez_id, estimate, se)
+    map(filter, entrez_id %in% dge_path$collagen_genes) %>% 
+    compress(names_to = 'cohort') %>% 
+    select(cohort, entrez_id, estimate, se)
 
 # calculation of the meta-regulation estimates -----
   
   insert_msg('Computing meta estimates')
-  
-  for(i in names(dge_path$collagen_tbl)) {
-    
-    dge_path$meta_tbl[[i]] <- dge_path$collagen_tbl[[i]] %>% 
-      blast(entrez_id) %>% 
-      future_map(~safely(metagen)(TE = .x$estimate, 
-                                  seTE = .x$se)) %>% 
-      map(~.$result) %>% 
-      map2_dfr(., names(.), ~tibble(entrez_id = .y, 
-                                    estimate = .x$TE.common, 
[…]

### 2023-08-07 dge scripts/dge_pathways.R

@@ -1,49 +1,43 @@
 # Modulation of signaling pathways with SPIA
+# Significantly regulated pathways are identified as pFDR < 0.05 and magnitude
+# of regulation expressed as tA > 2
 
   insert_head()
   
 # container -------
   
   dge_spia <- list()
-  
-# analysis globals -------
-  
-  insert_msg('Analysis globals')
-  
-  ## plot title prefixes
-  
-  dge_spia$prefixes <- 
-    c(int = 'Collagen int vs low', 
-      hi = 'Collagen high vs low')
-  
+
 # serial analysis -------
   
   insert_msg('Serial analysis with SPIA')
   
   plan('multisession')
   
-  dge_spia$test$int <- list(de = dge$regulation_int, 
-                            all = dge$all_vectors) %>% 
+  dge_spia$test <- list(de = dge$regulation_vectors, 
+                        all = dge$all_vectors) %>% 
     future_pmap(spia, 
                 verbose = FALSE, 
                 .options = furrr_options(seed = TRUE)) %>% 
     map(as_tibble)
   
-  dge_spia$test$hi <- list(de = dge$regulation_hi, 
-                            all = dge$all_vectors) %>% 
-    future_pmap(spia, 
-                verbose = FALSE, 
-                .options = furrr_options(seed = TRUE)) %>% 
-    map(as_tibble)
-
   plan('sequential')
-  
+
 # identification of the significantly regulated pathways and common ones -----
   
   insert_msg('Significant pathways')
   
+  dge_spia$test <- dge_spia$test %>% 
+    map(mutate, 
+        significant = ifelse(pGFdr < 0.05 & abs(tA) > 2, 
+                             'yes', 'no'), 
+        regulation = ifelse(significant == 'yes', 
+                            ifelse(tA > 0, 'activated', 'inhibited'), 
+                            'ns'), 
+        regulation = factor(regulation, c('activated', 'inhibited', 'ns')))
+  
   dge_spia$significant <- dge_spia$test %>% 
-    map(map, filter, pGFdr < 0.05)
+    map(filter, significant == 'yes')
 
 # Common regulated pathways -------
   
@@ -51,125 +45,118 @@
   
   ## in at least 4 out of five cohorts
   
-  dge_spia$cmm_sets <- names(dge_spia$significant[[1]]) %>% 
-    combn(m = 4, simplify = FALSE) %>% 
[…]

### 2023-08-07 dge scripts/dge_plots.R

@@ -18,13 +18,7 @@
   dge_plots$n_caps <- dge_plots$n_numbers %>% 
     map(~map2_chr(.x[[1]], .x[[2]], paste, sep = ': n = ')) %>% 
     map_chr(paste, collapse = ', ')
-  
-  ## plot title prefixes
-  
-  dge_plots$prefixes[c("dge_collagen_int", "dge_collagen_hi")] <- 
-    c('Collagen int vs low', 
-      'Collagen high vs low')
-  
+
 # N number plots -------  
   
   insert_msg('Numbers of differentialy regulated genes')
@@ -32,130 +26,94 @@
   ## stats
   
   dge_plots$dge_numbers$regulated <- 
-    dge[c("dge_collagen_int", "dge_collagen_hi")] %>% 
-    map(map, count, regulation) %>% 
-    map(compress, names_to = 'cohort')
-  
+    dge$signif_results %>% 
+    map(count, regulation) %>% 
+    compress(names_to = 'cohort')
+
   dge_plots$dge_numbers$total <- dge$annotation %>% 
     map_dbl(nrow) %>% 
     compress(names_to = 'cohort',
              values_to = 'n_total')
   
-  dge_plots$dge_numbers$regulated <- dge_plots$dge_numbers$regulated %>% 
-    map(left_join, 
-        dge_plots$dge_numbers$total, 
-        by = 'cohort') %>% 
-    map(mutate, 
-        cohort_lab = paste(globals$study_labels[cohort], 
-                           n_total, sep = '\nn = '), 
-        percent = n/n_total * 100, 
-        cohort = factor(cohort, names(dge$analysis_tbl)))
+  dge_plots$dge_numbers$regulated <- 
+    left_join(dge_plots$dge_numbers$regulated, 
+              dge_plots$dge_numbers$total, 
+              by = 'cohort') %>% 
+    mutate(cohort_lab = paste(globals$study_labels[cohort], 
+                              n_total, sep = '\nn = '), 
+           percent = n/n_total * 100, 
+           cohort = factor(cohort, names(dge$analysis_tbl)))
+
+  ## plot
   
-  ## plots
-  
-  dge_plots$dge_numbers$plots <- 
-    list(x = dge_plots$dge_numbers$regulated, 
-         y = dge_plots$prefixes) %>% 
-    pmap(function(x, y) x %>% 
-           ggplot(aes(x = percent, 
-                      y = cohort, 
-                      fill = regulation)) + 
-           geom_bar(stat = 'identity', 
[…]

### 2023-08-07 dge scripts/dge_reactome.R

@@ -39,104 +39,63 @@
   
   insert_msg('Testing')
   
-  ## one-way ANOVA: the ssGSVA scores have usually a nice normal distribution
+  ## two-tailed T test: the ssGSVA scores have usually
+  ## a nice normal distribution
+  ## computation of Cohen's d used later for selection of common significant
+  ## signatures and plot ordering
   
   dge_gsva$test <- dge_gsva$analysis_tbl %>% 
-    future_map(test_anova, 
+    future_map(test_two_groups,
+               type = 't', 
                split_fct = 'clust_id', 
                variables = dge_gsva$lexicon$variable, 
                adj_method = 'BH', 
                .parallel = FALSE, 
-               .options = furrr_options(seed = TRUE))
+               .options = furrr_options(seed = TRUE)) %>% 
+    map2(., dge_gsva$analysis_tbl, 
+         ~mutate(.x,  
+                 n = nrow(.y), 
+                 se = abs(estimate/stat), 
+                 sd = se * sqrt(n), 
+                 eff_size = abs(estimate/sd), 
+                 eff_size_desc = interpret_cohens_d(eff_size), 
+                 regulation = ifelse(significant == 'yes' & 
+                                       eff_size > 0.3, 
+                                     ifelse(estimate > 0, 
+                                            'upregulated', 'downregulated'), 
+                                     'ns'), 
+                 regulation = factor(regulation, 
+                                     c('upregulated', 'downregulated', 'ns'))))
+         
+# Significant and common signatures ------
   
-# Effect size computation ------
+  insert_msg('Significant and common significant signatures')
   
-  insert_msg('Eta-squared')
+  ## pFDR < 0.05 and at least small effect size
   
-  for(i in names(dge_gsva$analysis_tbl)) {
-    
-    dge_gsva$eff_size[[i]] <- dge_gsva$lexicon$variable %>% 
-      future_map_dbl(get_etasq, 
-                     split_factor = 'clust_id', 
-                     data = dge_gsva$analysis_tbl[[i]], 
-                     .options = furrr_options(seed = TRUE))
-    
-  }
-  
-  ## formatting the results
-  ## significant effect is defined as eta^2 >= 0.06
[…]

### 2023-08-07 dge scripts/dge_testing.R

@@ -1,8 +1,8 @@
 # Identification of differentially regulated genes 
 # between the Collagen clusters
-# Differentially regulated genes are identified by one-way ANOVA
+# Differentially regulated genes are identified by two-tailed T test
 # corrected for multiple testing with the Benjamini-Hochberg method
-# Regulation cutoff: 1.25-fold
+# Regulation cutoff: 1.2-fold
 
   insert_head()
   
@@ -83,15 +83,16 @@
         clust_id = factor(clust_id, 
                           c('Collagen low', 
                             'Collagen int', 
-                            'Collagen hi')))
+                            'Collagen hi')), 
+        clust_id = droplevels(clust_id))
   
-# Serial ANOVA -------
+# Serial t test -------
   
   insert_msg('Serial testing')
   
   dge$test_results <- list(data = dge$analysis_tbl, 
                            variables = dge$variables) %>% 
-    pmap(test_anova, 
+    pmap(test_two_groups, 
          split_fct = 'clust_id', 
          type = 't', 
          adj_method = 'BH', 
@@ -101,107 +102,65 @@
   
   insert_msg('Annotation of the results')
   
-  for(i in names(dge$analysis_tbl)) {
-    
-    dge$test_results[[i]]$anova <- dge$test_results[[i]]$anova %>% 
-      mutate(gene_symbol = response, 
-             entrez_id = exchange(gene_symbol, 
-                                  dge$annotation[[i]], 
-                                  key = 'gene_symbol', 
-                                  value = 'entrez_id'))
-    
-    dge$test_results[[i]]$lm <- dge$test_results[[i]]$lm %>% 
-      mutate(gene_symbol = response, 
-             entrez_id = exchange(gene_symbol, 
-                                  dge$annotation[[i]], 
-                                  key = 'gene_symbol', 
-                                  value = 'entrez_id'))
-    
-    
-  }
+  dge$test_results <- 
+    map2(dge$test_results, 
+         dge$annotation, 
+         ~mutate(.x, 
+                 gene_symbol = response, 
+                 entrez_id = exchange(gene_symbol, 
+                                      .y, 
+                                      key = 'gene_symbol', 
+                                      value = 'entrez_id')))
+
 
 # Formatting the results and identifying differentially regulated genes ------
   
[…]

### 2023-08-07 dge.R

@@ -58,7 +58,7 @@
   
   if(file.exists('./cache/dge_spia.RData')) {
     
-    insert_msg('Loading cached signalig results')
+    insert_msg('Loading cached signaling results')
     
     load('./cache/dge_spia.RData')
     

### 2023-08-07 import scripts/geo_import.R

@@ -1,62 +1,100 @@
-# This script imports prostate cancer data from GEO (GSE16560, GSE40272, GSE70768 and GSE70769)
+# Import of the GEO datasets 
 
-  insert_head()
-  
+
+insert_head()
+
+# tools ------
+
+  library(tidyverse)
+  library(GEOquery)
+
 # data containers -----
-  
+
   geo_globals <- list()
   geo_data <- list()
-    
-# definition of folders with data to import and strings to clear from the design data sets ----
   
-  geo_globals$study_list <- c('GSE16560', 
-                              'GSE40272-GPL15971',
-                              'GSE40272-GPL15972', 
-                              'GSE40272-GPL15973', 
-                              'GSE40272-GPL9497', 
-                              'GSE70768', 
-                              'GSE70769') %>% 
-    paste('./input data', ., sep = '/')
+# fetching the GEO objects ------
   
-  geo_globals$clearing_strings <- c('gleason', ': ', 'major.', 'fup.month', 'diag.yr', 'status.all', 'extreme', 'cancer.percent', 
-                                    'age', 'batch', 'minor.', 'ethnicity', 'individual', 'disease state', 'pre-operation treatment', 
-                                    'tissue', 'disease free survival', 'sample type', 'tumour', '%', 'iclusterplus group', 
-                                    'extra-capsular extension (ece)', 'positive surgical margins (psm)', 'biochemical relapse (bcr)',
-                                    'time to bcr (months)', 'tmprss2ERG gene fusion status:', 'at diag', 'psa ', 'clinical ', 'pathology ', 
-                                    'total follow up (months)', 'st', 'extra capsular extension (ece)', 'derived data ()')
+  insert_msg('Fetching the GEO objects')
   
-# definition of objects exported to the processor cluster for parallel computing ---- 
+  geo_globals$ids <- 
+    c('GSE16560', 
+      'GSE70768', 
+      'GSE70769', 
+      'GSE116918')
   
-  geo_globals$cl_export <- c('transpose_tibble', 'read_expression', 'read_design', 'read_annotation', 
-                             'untangle_gleason_sum', 'read_study', 'read_excel', 'parse_guess', 'as_tibble', 'set_names', 
[…]

### 2023-08-07 import.R

@@ -23,6 +23,8 @@
   library(doParallel)
   library(soucer)
 
+  explore <- exda::explore
+
   c('./tools/import_engine.R', 
     './tools/tcga_tools.R', 
     './tools/project_tools.R') %>% 
@@ -72,6 +74,10 @@
     c('./clearing scripts/data_clearing_geo.R', 
       './clearing scripts/data_clearing_tcga.R') %>% 
       source_all(message = TRUE, crash = TRUE)
+    
+    study_data <- 
+      c(geo_cleared, 
+        list(tcga = tcga_cleared))
     
     save(study_data, file = './data/study_data.RData')
     
@@ -135,5 +141,7 @@
     source_all(message = TRUE, crash = TRUE)
   
 # END -----
+  
+  rm(geo_data, tcga_data)
   
   insert_tail()

### 2023-08-07 metabolism scripts/metab_enrichment.R

@@ -18,19 +18,17 @@
   insert_msg('Enrichment testing')
   
   meta_sub$test <- meta$models %>% 
-    map(map, 
-        suba, 
+    map(suba, 
         method = 'fisher', 
         signif_type = 'fdr', 
         .parallel = TRUE)
   
 # Numbers of all and significantly regulated reactions ------
   
-  insert_msg('Numbers of all and significantly regulated reactions')
+  insert_msg('Numbers of all, and activated and inhibited reactions')
   
   meta_sub$react_n <- meta_sub$test %>% 
-    map(map, 
-        summarise, 
+    map(summarise, 
         n_all = sum(n), 
         n_all_total = sum(n_total), 
         .by = status)
@@ -43,17 +41,14 @@
   ## this is computed with the following formula
   ## (n(DE, subsystem)/n(subsystem))/(n(DE)/n(total))
   
-  for(i in names(meta_sub$test)) {
-    
-    meta_sub$test[[i]] <- 
-      map2(meta_sub$test[[i]], 
-           meta_sub$react_n[[i]], 
-           left_join, by = 'status') %>% 
-      map(mutate, 
-          OR = (n/n_total)/(n_all/n_all_total), 
-          eff_size = interpret_oddsratio(OR, rules = 'cohen1988'))
-    
-  }
+  meta_sub$test <- 
+    map2(meta_sub$test, 
+         meta_sub$react_n, 
+         left_join, 
+         by = 'status') %>% 
+    map(mutate, 
+        OR = (n/n_total)/(n_all/n_all_total), 
+        eff_size = interpret_oddsratio(OR, rules = 'cohen1988'))
   
 # Identification of significantly activated and inhibited subsystems ------
   
@@ -62,14 +57,13 @@
   ## pFDR < 0.05, OR > 1 and at least small effect size (Cohen1988)
   
   meta_sub$significant <- meta_sub$test %>% 
-    map(map, 
-        filter, 
+    map(filter, 
         status %in% c('activated', 'inhibited'), 
         p_adjusted < 0.05, 
         OR > 1, 
         eff_size %in% c('small', 'medium', 'large')) %>% 
-    map(map, blast, status) %>% 
-    map(transpose)
+    map(blast, status) %>% 
+    transpose
   
 # Identification of common activated and inhibited subsystems -----
   
@@ -77,20 +71,16 @@
   
   ## significantly enriched in at least 4 out of 5 investigated cohorts
[…]

### 2023-08-07 metabolism scripts/metab_genes.R

@@ -7,12 +7,6 @@
   
   meta_gene <- list()
   
-# parallel backend ------
-  
-  insert_msg('Parallel backend')
-  
-  plan('multisession')
-  
 # globals ------
   
   insert_msg('Globals setup')
@@ -20,12 +14,15 @@
   ## common regulated subsystemts
   
   meta_gene$subsystems <- meta_sub$common %>% 
-    map(reduce, union) %>% 
-    reduce(union)
+    unlist %>% 
+    unique
+  
+  meta_gene$subsystems <- 
+    c(meta_gene$subsystems, 'Oxidative phosphorylation')
   
   ## reactions of the common regulated subsystems
   
-  meta_gene$reactions <- meta$regulation[[1]][[1]] %>% 
+  meta_gene$reactions <- meta$regulation[[1]] %>% 
     filter(subsystem %in% meta_gene$subsystems) %>% 
     mutate(subsystem = factor(subsystem, meta_gene$subsystems)) %>% 
     blast(subsystem) %>% 
@@ -34,43 +31,20 @@
   ## genes
   
   meta_gene$entrez_id <- meta_gene$reactions %>% 
-    map(react_to_gene, meta$models[[1]][[1]]) %>% 
+    map(react_to_gene, meta$models[[1]]) %>% 
     map(~.x$entrez_id) %>% 
     map(reduce, union)
-  
-  ## plot title suffixes
-  
-  meta_gene$title_suffix <- 
-    c(int = 'Collagen int vs low', 
-      hi = 'Collagen high vs low')
-  
+
 # Extraction of regulation estimates -------
   
   insert_msg('Extraction of gene regulation estimates')
   
-  meta_gene$estimates <- dge$test_results %>% 
-    map(~.x$lm) %>% 
-    map(mutate, 
-        regulation = ifelse(significant == 'no', 
-                            'ns', 
-                            ifelse(estimate > log2(1.25), 
-                                   'upregulated', 
-                                   ifelse(estimate < log2(1.25), 
-                                          'downregulated', 'ns'))), 
-        regulation = factor(regulation, c('upregulated', 'downregulated', 'ns')))
-  
-  meta_gene$estimates <- meta_gene$estimates %>% 
-    map(function(data) meta_gene$entrez_id %>% 
-          map(~filter(data, 
-                      entrez_id %in% .x, 
-                      level %in% c('Collagen hi', 'Collagen int'))) %>% 
[…]

### 2023-08-07 metabolism scripts/metab_hg.R

@@ -8,12 +8,6 @@
   
   meta_hg <- list()
 
-# parallel backend ------
-  
-  insert_msg('Parallel backend')
-  
-  plan('multisession')
-  
 # Globals -------
   
   insert_msg('Globals setup')
@@ -29,14 +23,14 @@
   ## subsystems of interest, reactions and metabolites
   ## the common ones, citric acid cycle and OxPhos
   
-  meta_hg$subsystems <- meta_sub$common[c("int", "hi")] %>% 
+  meta_hg$subsystems <- meta_sub$common[c("activated", "inhibited")] %>% 
     unlist %>% 
     unique
   
   meta_hg$subsystems <- 
-    c(meta_hg$subsystems, c('Citric acid cycle', 'Oxidative phosphorylation'))
+    c(meta_hg$subsystems, c('Oxidative phosphorylation'))
   
-  meta_hg$reactions <- meta$regulation[[1]][[1]] %>% 
+  meta_hg$reactions <- meta$regulation[[1]] %>% 
     filter(subsystem %in% meta_hg$subsystems) %>% 
     mutate(subsystem = factor(subsystem, meta_hg$subsystems)) %>% 
     blast(subsystem) %>% 
@@ -50,72 +44,46 @@
     map(unlist) %>% 
     map(unique)
 
-  ## plot title suffixes
-  
-  meta_hg$title_suffix <- 
-    c(int = 'Collagen int vs low', 
-      hi = 'Collagen high vs low')
-  
-  meta_hg$short_suffix <- 
-    c(int = 'int vs low', 
-      hi = 'high vs low')
-  
 # regulation estimates ------
   
   insert_msg('Regulation estimates')
   
-  for(i in names(meta$regulation)) {
-    
-    ## labeling of the reactions: those annotated with R_*, i.e. 
-    ## with no enzymes attached to them, are labeled as NA and skipped
-    ## in the Forest plots of the top regulated reactions
-    
-    meta_hg$estimates[[i]] <- meta$regulation[[i]] %>% 
-      map(mutate, 
-          log_reg = log2(fold_reg), 
-          log_lower = log2(lower_ci), 
-          log_upper = log2(upper_ci), 
-          react_name = annotate_bigg(react_id), 
-          react_lab = ifelse(stri_detect(react_name, regex = '^RE\\d+'), 
-                             NA, 
-                             paste(react_name, react_id, sep = '\n'))) %>% 
-      map(filter, 
[…]

### 2023-08-07 metabolism scripts/metab_plots.R

@@ -5,131 +5,47 @@
 # container -----
   
   meta_plots <- list()
-  
-# globals -----
-  
-  insert_msg('Globals')
-  
-  ## common regulated reactions
-  
-  meta_plots$common_reactions <- 
-    meta$common_reactions %>% 
-    map(reduce, union)
-  
-  ## labels
-  
-  meta_plots$clust_labels <- 
-    c(int = 'Collagen int vs Collagen low', 
-      hi = 'Collagen high vs Collagen low')
-  
+
 # Diagnostic plots: errors ------
   
   insert_msg('Diagnostic plots: errors')
   
   meta_plots$errors <- meta$models %>% 
-    map(map, 
-        plot, 
+    map(plot, 
         type = 'errors', 
         cust_theme = globals$common_theme)
-
-  ## plot_titles
-  
-  for(i in names(meta_plots$errors)) {
-    
-    meta_plots$errors[[i]] <- 
-      map2(meta_plots$errors[[i]], 
-           paste('Errors and estimates,', globals$study_labels), 
-           ~.x + 
-             labs(title = .y, 
-                  subtitle = meta_plots$clust_labels))
-    
-  }
 
 # General regulation plots -------
   
   insert_msg('General regulation plots')
   
   meta_plots$regulation <- meta$models %>% 
-    map(map, 
-        plot, 
+    map(plot, 
         type = 'regulation', 
         show_fit = FALSE, 
-        cust_theme = globals$common_theme)
-  
-  ## plot titles
-  
-  for(i in names(meta_plots$regulation)) {
-    
-    meta_plots$regulation[[i]] <- 
-      map2(meta_plots$regulation[[i]], 
-           paste(meta_plots$clust_labels[[i]], 
-                 globals$study_labels, 
-                 sep = ', '), 
-           ~.x + 
-             theme(panel.grid.major.x = element_blank()) + 
-             labs(title = .y,, 
-                  y = expression('log'[2] * 'fold-regulation')))
-    
-  }
+        cust_theme = globals$common_theme) %>% 
+    map2(., 
+         globals$study_labels[names(meta$models)], 
+         ~.x + 
+           theme(panel.grid.major.x = element_blank()) + 
+           labs(title = .y,, 
+                y = expression('log'[2] * 'fold-regulation')))
 
 # Top regulated reactions ------
   
   insert_msg('Top regulated reactions')
   
   meta_plots$top_plots <- meta$models %>% 
-    map(map, 
[…]

### 2023-08-07 metabolism scripts/metab_testing.R

@@ -2,7 +2,7 @@
 # in the Recon2 BiGG model. 
 # SEM brought back to the linear scale (from log2) by the first-differential 
 # equation log(2) * SEM(log2-Expression) * 2^log2-Expression
-# No multiple-testing correction!!!
+# FDR multiple testing correction
 
   insert_head()
   
@@ -18,18 +18,10 @@
   ## annotation with Entrez ID
   
   meta$analysis_tbl <- dge$test_results %>% 
-    map(~.x$lm) %>% 
-    map(filter, level != '(Intercept)') %>% 
     map(mutate, 
-        level = stri_extract(level, regex = 'int|hi'), 
-        level = factor(level, c('int', 'hi')), 
         se = log(2) * se * 2^estimate, 
         estimate = 2^estimate) %>% 
-    map(~.x[c('level', 'gene_symbol', 'entrez_id', 'estimate', 'se')])
-
-  meta$analysis_tbl <- meta$analysis_tbl %>% 
-    map(blast, level) %>% 
-    transpose
+    map(~.x[c('gene_symbol', 'entrez_id', 'estimate', 'se')])
 
 # construction of the models ------
   
@@ -37,7 +29,7 @@
   
   set.seed(1234)
   
-  meta$models$int <- meta$analysis_tbl$int %>% 
+  meta$models <- meta$analysis_tbl %>% 
     map(~build_geneSBML(x = set_names(.x$estimate, .x$entrez_id), 
                         err = set_names(.x$se, .x$entrez_id), 
                         database = Recon2D, 
@@ -45,19 +37,7 @@
                         and_fun = 'min', 
                         x_default = 1, 
                         err_method = 'mc', 
-                        n_iter = 1010, 
-                        ci_method = 'bca', 
-                        .parallel = TRUE))
-  
-  meta$models$hi <- meta$analysis_tbl$hi %>% 
-    map(~build_geneSBML(x = set_names(.x$estimate, .x$entrez_id), 
-                        err = set_names(.x$se, .x$entrez_id), 
-                        database = Recon2D, 
-                        or_fun = 'mean', 
-                        and_fun = 'min', 
-                        x_default = 1, 
-                        err_method = 'mc', 
-                        n_iter = 1010, 
+                        n_iter = 2010, 
                         ci_method = 'bca', 
                         .parallel = TRUE))
   
@@ -65,38 +45,30 @@
   
[…]

### 2023-08-07 metabolism.R

@@ -62,7 +62,7 @@
     
   } else {
     
-    c('/metabolism scripts/metab_enrichment.R') %>% 
+    c('./metabolism scripts/metab_enrichment.R') %>% 
       source_all(message = TRUE, crash = TRUE)
     
   }

### 2023-08-07 mutation scripts/counts.R

@@ -51,7 +51,7 @@
     compare_variables(variables = mut_counts$lexicon$variable, 
                       split_factor = 'clust_id', 
                       what = 'eff_size', 
-                      types = 'kruskal_etasq', 
+                      types = 'wilcoxon_r', 
                       exact = FALSE, 
                       ci = FALSE, 
                       pub_styled = TRUE, 

### 2023-08-07 mutation scripts/tables.R

@@ -179,7 +179,7 @@
   ## mutations: present in at least 2% of participants
   ## CNV: at least 10% of participants
   ## amplifications: at least 4% of participants
-  ## deletions: at least 10% of particiants
+  ## deletions: at least 10% of participants
   
   mut_tables$top[c('mutations', 
                    'cnv', 

### 2023-08-07 report scripts/links.R

@@ -19,7 +19,8 @@
          'microViz' = 'https://github.com/PiotrTymoszuk/microViz', 
          'biggrExtra' = 'https://github.com/PiotrTymoszuk/biggrExtra', 
          'coxExtensions' = 'https://github.com/PiotrTymoszuk/coxExtensions', 
-         'kmOptimizer' = 'https://github.com/PiotrTymoszuk/kmOptimizer') %>%
+         'kmOptimizer' = 'https://github.com/PiotrTymoszuk/kmOptimizer', 
+         'gseaTools' = 'https://github.com/PiotrTymoszuk/gseaTools') %>%
     compress(names_to = 'obj_name',
              values_to = 'x') %>%
     mutate(ref_name = paste0('_', obj_name, '_'))

### 2023-08-07 report scripts/paper_figures.R

@@ -18,8 +18,6 @@
                 theme(plot.margin = globals$common_margin), 
               plot_grid(mri$plot + 
                           theme(legend.position = 'none'), 
-                        get_legend(norm_tumor$ribbon_plots$plots[[1]] + 
-                                     theme(legend.position = 'bottom')), 
                         nrow = 2, 
                         rel_heights = c(0.7, 0.3)), 
               ncol = 2, 
@@ -27,26 +25,18 @@
               align = 'hv', 
               axis = 'tblr')
   
-  ## bottom panel: ribbon plots with the expression comparison
-  ## common regulated collagen pathway genes
+  ## bottom panel: heat map of mean normalized expression 
+  ## for the common regulated collagen pathway genes
   
-  paper_fig$norm_tumor$bottom <- norm_tumor$ribbon_plots$plots %>% 
-    map(~.x + 
-          labs(title = paste(.x$labels$title, .x$labels$subtitle, sep = ', ')) + 
-          theme(legend.position = 'none', 
-                plot.subtitle = element_blank())) %>% 
-    plot_grid(plotlist = ., 
-              ncol = 3, 
-              align = 'hv',
-              axis = 'tblr')
-  
+  paper_fig$norm_tumor$bottom <- norm_tumor$hm_plot$plot
+
   ## the entire figure
   
   paper_fig$norm_tumor <- 
     plot_grid(paper_fig$norm_tumor$upper, 
               paper_fig$norm_tumor$bottom, 
               nrow = 2, 
-              rel_heights = c(1, 1.4),
+              rel_heights = c(1.4, 1),
               labels = LETTERS, 
               label_size = 10) %>% 
     as_figure(label = 'figure_1_normal_tumor', 
@@ -55,7 +45,7 @@
                               'and expression of the collagen pathway genes', 
                               'between the prostate cancer and benign tissue.'), 
               w = 180, 
-              h = 220)
+              h = 160)
   
 # Figure 2: collagen clusters ------
   
@@ -65,10 +55,13 @@
   ## and cluster distribution
   
   paper_fig$coll_clusters$left <- 
-    list(coll_clust$mean_hm$plot, 
-         coll_clust$n_plot) %>% 
-    map(~.x + 
-          theme(legend.position = 'bottom', 
+    list(coll_clust$umap_layouts$tcga + 
+           labs(x = 'UMAP 1', 
+                y = 'UMAP 2'), 
+         coll_clust$mean_hm$plot) %>% 
+    map2(., c('right', 'bottom'), 
[…]

### 2023-08-07 report scripts/paper_supplement.R

@@ -10,7 +10,7 @@
   
   insert_msg('Figure S1: comparison of expression, tumor versus normal')
   
-  ## upper panel, Venn plots, genes regulated in at least two cohort
+  ## upper panel, Venn plots, genes regulated in both two cohort
   ## are listed
   
   suppl_paper_fig$norm_tumor$upper <- 
@@ -19,16 +19,17 @@
               align = 'hv', 
               axis = 'tblr')
   
-  ## bottom panel, Volcano plots with significance 
-  ## and effect size of the regulation
+  ## bottom panel, ribbon plots
   
   suppl_paper_fig$norm_tumor$bottom <- 
-    norm_tumor$volcano_plots %>% 
+    norm_tumor$ribbon_plots$plots %>% 
     map(~.x + 
-          theme(legend.position = 'none')) %>% 
-    c(list(legend = get_legend(norm_tumor$volcano_plots[[1]]))) %>% 
+          theme(legend.position = 'none', 
+                axis.text.y = element_markdown(size = 7))) %>% 
+    c(list(legend = get_legend(norm_tumor$ribbon_plots$plots[[1]]))) %>% 
     plot_grid(plotlist = ., 
-              ncol = 2, 
+              ncol = 3, 
+              rel_widths = c(1, 1, 0.25), 
               align = 'hv', 
               axis = 'tblr')
   
@@ -40,7 +41,7 @@
               nrow = 2, 
               align = 'hv', 
               axis = 'tblr', 
-              rel_heights = c(1, 2), 
+              rel_heights = c(0.25, 1), 
               labels = LETTERS, 
               label_size = 10) %>% 
     as_figure(label = 'figure_s1_normal_tumor', 
@@ -114,23 +115,26 @@
   insert_msg('Figure S3: cluster defining factors')
   
   suppl_paper_fig$clusters <- 
-    coll_clust$ribbon_plots[c('tcga', 
-                              'GSE16560', 
-                              'GSE40272', 
-                              'GSE70768', 
-                              'GSE70769')] %>% 
+    coll_clust$ribbon_plots[c("tcga", 
+                              "GSE16560", 
+                              "GSE70768", 
+                              "GSE70769", 
+                              "GSE116918")] %>% 
     map2(., 
          c('TCGA, training', 
            'GSE16560, test', 
-           'GSE40272, test', 
            'GSE70768, test', 
-           'GSE70769, test'), 
+           'GSE70769, test', 
+           'GSE116918, test'), 
[…]

### 2023-08-07 report scripts/paper_tables.R

@@ -22,24 +22,33 @@
                             'percentages of categories within the', 
                             'complete observation set.'))
   
-# Table 2: collagen genes of interest -----
+# Table S1: collagen genes of interest -----
   
-  insert_msg('Table 2: collagen genes of interest')
+  insert_msg('Table S1: collagen genes of interest')
   
-  paper_tbl$genes <- globals$genes_interest %>% 
+  suppl_paper_tbl$genes <- globals$genes_interest %>% 
     mutate(entrez_id = mapIds(org.Hs.eg.db, 
                               keys = gene_symbol, 
                               column = 'ENTREZID', 
-                              keytype = 'SYMBOL')) %>% 
-    set_names(c('Gene symbol', 'Entrez ID', 'Gene group')) %>% 
-    mdtable(label = 'table_2_genes', 
+                              keytype = 'SYMBOL'),
+           gene_group = car::recode(gene_group, "'Pro' = 'proline turnover'"), 
+           gene_group = factor(gene_group, 
+                               c('proline turnover', 
+                                 'collagen modification', 
+                                 'ECM component', 
+                                 'ECM processing', 
+                                 'adhesion'))) %>% 
+    arrange(gene_group) %>% 
+    select(gene_group, gene_symbol, entrez_id) %>% 
+    set_names(c('Gene group', 'Gene symbol', 'Entrez ID')) %>% 
+    mdtable(label = 'table_s2_genes', 
             ref_name = 'genes', 
-            caption = paste('Collagen genes of interest', 
-                            'and their classification.'))
+            caption = paste('Investigated collagen pathway genes and', 
+                            'their classification.'))
   
-# Table S1: normal-tumor ------
+# Table S2: normal-tumor ------
   
-  insert_msg('Table S1: normal - tumor')
+  insert_msg('Table S2: normal - tumor')
   
   suppl_paper_tbl$norm_tumor <- 
     map2(norm_tumor$stats, 
@@ -57,7 +66,7 @@
     select(cohort, variable, tumor, benign, significance, est_lab) %>% 
     set_names(c('Cohort', 'Variable', 'Tumor', 'Benign', 
                 'Significance', 'Effect size')) %>% 
-    mdtable(label = 'table_s1_normal_tumor', 
+    mdtable(label = 'table_s2_normal_tumor', 
             ref_name = 'norm_tumor', 
             caption = paste("Expression of the collagen pathway genes", 
                             "in the malignant and benign tissue compared", 
@@ -71,26 +80,26 @@
[…]

### 2023-08-07 report.R

@@ -34,13 +34,14 @@
   insert_msg('Paper scripts')
   
   ## files for the analysis report
+  ## skipped for the moment
   
-  c('./report scripts/links.R', 
-    './report scripts/report_tables.R', 
-    './report scripts/report_figures.R', 
-    './report scripts/report_supplement.R') %>% 
-    source_all(message = TRUE, crash = TRUE) %>% 
-    print
+  #c('./report scripts/links.R', 
+   # './report scripts/report_tables.R', 
+    #'./report scripts/report_figures.R', 
+  #  './report scripts/report_supplement.R') %>% 
+   # source_all(message = TRUE, crash = TRUE) %>% 
+    #print
   
   ## files for the manuscript
   

### 2023-08-07 tools/globals.R

@@ -46,16 +46,67 @@
   
   ## classification of the genes by their function
   
+  globals$gene_classes <-
+    c('ALDH18A1' = 'Pro', 
+      'PYCR1' = 'Pro', 
+      'PLOD2' = 'collagen modification', 
+      'P4HA1' = 'collagen modification', 
+      'P4HA2' = 'collagen modification', 
+      'PEPD' = 'Pro', 
+      'COL3A1' = 'ECM component', 
+      'COL2A1' = 'ECM component', 
+      'COL11A1' = 'ECM component', 
+      'COL11A2' = 'ECM component', 
+      'COL5A2' = 'ECM component', 
+      'COL5A1' = 'ECM component', 
+      'COL1A2' = 'ECM component', 
+      'COL7A1' = 'ECM component', 
+      'BMP1' = 'ECM processing', 
+      'LAMA3' = 'ECM component', 
+      'LAMC2' = 'ECM component', 
+      'LAMB3' = 'ECM component', 
+      'COL4A2' = 'ECM component', 
+      'COL4A1' = 'ECM component', 
+      'COL4A5' = 'ECM component', 
+      'COL4A3' = 'ECM component', 
+      'COL4A6' = 'ECM component', 
+      'MMP13' = 'ECM processing', 
+      'MMP9' = 'ECM processing', 
+      'CTSS' = 'ECM processing', 
+      'MMP7' = 'ECM processing', 
+      'COL18A1' = 'ECM component', 
+      'COL15A1' = 'ECM component', 
+      'COL14A1' = 'ECM component', 
+      'COL6A3' = 'ECM component', 
+      'COL6A1' = 'ECM component', 
+      'COL6A2' = 'ECM component', 
+      'DST' = 'adhesion', 
+      'COL17A1' = 'ECM component', 
+      'ITGA6' = 'adhesion', 
+      'ITGB4' = 'adhesion', 
+      'CD151' = 'adhesion', 
+      'COL9A2' = 'ECM component', 
+      'COL9A1' = 'ECM component', 
+      'COL9A3' = 'ECM component', 
+      'PCOLCE' = 'ECM processing', 
+      'LOXL2' = 'collagen modification', 
+      'LOXL1' = 'collagen modification', 
+      'LOX' = 'collagen modification', 
+      'COL16A1' = 'ECM component', 
+      'COL19A1' = 'ECM component', 
+      'PLOD3' = 'collagen modification', 
[…]

### 2023-08-07 tools/import_engine.R

@@ -389,7 +389,7 @@
                             reposit_identifier = 'sample_ID') {
     
     ## screens the expression table for duplicates and chooses the sample
-    ## with the latest reportitory id
+    ## with the latest repository id
     
     start_time <- Sys.time()
     message(paste('Screening for duplicates within', 

### 2023-08-07 tools/project_tools.R

@@ -17,6 +17,29 @@
 
 # Clinical data formatting -------
 
+  extract_pfield <- function(x, 
+                             format = c('numeric', 'factor', 'character'), 
+                             levels = NULL) {
+    
+    format <- match.arg(format[1], 
+                        c('numeric', 'factor', 'character'))
+
+    if(is.null(levels)) {
+      
+      levels <- sort(unique(as.character(x)))
+      
+    }
+        
+    x <- stri_replace(x, regex = '.*:\\s{1}', replacement = '')
+    
+    switch(format, 
+           numeric = as.numeric(x), 
+           factor = factor(x, levels = levels), 
+           character = as.character(x))
+    
+    
+  }
+
   regex_replacer <- function(x, regex, replacement = '') {
     
     if(is.character(x)) {
@@ -153,6 +176,7 @@
 
   plot_venn <- function(plotting_lst,  
                         text = NULL, 
+                        show_text = TRUE, 
                         colors = c('blue', 'yellow', 'green', 'red'), 
                         plot_title = NULL, 
                         plot_subtitle = NULL, 
@@ -170,10 +194,12 @@
              fill_color = unname(colors), 
              set_name_size = 2.75, 
              text_size = 2.75) + 
-      theme(plot.title = element_text(size = 10, face = 'bold'), 
+      theme(plot.title = element_text(size = 8, face = 'bold'), 
             plot.subtitle = globals$common_text) + 
       labs(title = plot_title, 
            subtitle = plot_subtitle)
+    
+    if(!show_text) return(venn_plot)
     
     ## feature listing
     
@@ -402,11 +428,16 @@
               'Valine' = 'Val', 
               'leucine' = 'Leu', 
               'and ' = '', 
-              'metabolism' = '\nmetabolism', 
+              ' infection' = '\ninfection', 
+              ' metabolism' = '\nmetabolism', 
+              ' synthesis' = '\nsynthesis', 
+              'Phosphatidylinositol phosphate' = 'PIP', 
               'interconversion' = '\ninterconversion', 
               'Transport, extracellular' = 'Extracellular\ntransport', 
               'adhesion' = '\nadhesion', 
-              'Regulation of actin cytoskeleton' = 'ACT\ncytoskeleton')
+              'Regulation of actin cytoskeleton' = 'ACT\ncytoskeleton', 
+              'TGF-beta signaling pathway' = 'TGF-\u03B2', 
+              ' in' = '\nin')
[…]

### 2023-08-07 report/markdown/manuscript_figures_tables.Rmd
Figures, Tables, Supplementary Material

--- 2023-07-26 report/markdown/manuscript_figures_tables.Rmd
@@
 title: "Prognostic and biologic relevance of collagen biosynthesis pathway in prostate cancer"
-subtitle: "Bioinformatic methods, figures and tables"
+subtitle: "Methods, figures and tables for the transcriptome part"
 date: "`r format(Sys.time(), '%Y-%m-%d')`"
@@
 
+# Methods
+
+## Magnetic resonance imaging of the prostate
+
+_Dear co-authors, it's your part as experts:)_
+
+ADC was compared between patient-matched measurements of benign prostate and malignancy by paired T test with Cohen's d effect size statistic.
+
+## Transcriptome analyses
+
+The transcriptome analysis was performed with R version 4.2.3. 
+Its details are presented in __Supplementary Methods__.
+
+### Datasets
+
+Five publicly available transcriptome datasets of primary prostate cancer samples, 
+the TCGA prostate cancer cohort [@Abeshouse2015] (n = `r study_data$tcga$expression %>% filter(tissue == 'tumor') %>% nrow`), 
+GSE16560 (n = `r nrow(study_data$GSE16560$expression)`) [@Sboner2010], 
+GSE70768 (n = `r study_data$GSE70768$expression %>% filter(tissue == 'tumor') %>% nrow`) [@Ross-Adams2015], 
+GSE70769 (n = `r nrow(study_data$GSE70769$expression)`) [@Ross-Adams2015], and
+GSE116918 (n = `r nrow(study_data$GSE116918$expression)`) [@Jain2018],  were re-analyzed. 
+Normalized gene expression data were fetched from the GDC data portal for the TCGA or from the Gene Expression Omnibus for the remaining cohorts. 
+Donor-matched benign and malignant tissue samples were available for the TCGA 
+(n = `r study_data$tcga$expression %>% filter(tissue == 'benign') %>% nrow` specimen pairs) and 
+GSE70768 cohort (n = `r study_data$GSE70768$expression %>% filter(tissue == 'benign') %>% nrow`). 
+Expression values were $log_2$-transformed prior to analyses. 
+Non-immune cell content in cancer samples was estimated with the _MCP-counter_ and _xCell_ algorithms [@Aran2017; @Becht2016].
+single sample gene set enrichment scores (ssGSEA) for the Reactome pathway gene signatures retrieved from the MSig database version 7.5.1 were computed with the _GSVA_ algorithm [@Hanzelmann2013].
+
+The collagen pathway genes were retrieved from the published Collagen Signature [@Kocher2021] and the Reactome collagen formation pathway (R-HSA-1474290) and restricted to the gene set present in all investigated datasets (__Supplementary Table S2__).
+
+### Differential expression in benign and malignant prostate tissue
+
+Expression of the collagen pathway gene between patient-matched benign and malignant specimens was compared by the false discovery rate method (FDR) [@Benjamini1995] corrected paired two-tailed T test with Cohen's d effect size statistic [@Cohen2013].
+
+### Collagen clusters
+
+Cancer samples in the training TCGA cohort were clustered in respect to COMBAT-batch adjusted, normalized $log_2$-transformed gene expression levels of the collagen pathway genes with the PAM algorithm with cosine distance between the samples [@Schubert2019; @Leek2012].
+The algorithm choice was motivated by its high explained clustering variance and stability in 10-fold cross-validation as compared with several other procedures [@Lange2004; @Leng2014]. 
+Two sample clusters were defined [@Rousseeuw1987]: collagen^high^ and collagen^low^ named after differences in expression of collagen subunit genes. 
+Cancers samples in the GSE16560, GSE70768, GSE70769 and GSE116918 collectives were assigned to the collagen clusters by an inverse distance weighted k-nearest neighbor classifier [@Leng2014].
+
+Expression of the collagen pathway genes in the collagen clusters was compared by FDR-adjusted two-tailed T test with Cohen's d effect size statistic. 
+Differences in clinical variables between the collagen clusters were assessed by FDR-corrected Mann-Whitney test with r effect size statistic or FDR-corrected $\chi^2$ test with Cramer V effect size statistic.
+
+### Gene set variation analysis and differential gene expression in the collagen clusters
+
+ssGSEA scores of the Reactome signatures and $log_2$-transformed gene expression levels were compared between the collagen clusters by FDR-corrected two-tailed T tests with Cohen's d effect size statistic.
+Signatures and genes significantly regulated between the clusters were defined by the significance cutoff (pFDR < 0.05) and at least weak effect size of the differences between the clusters (d $\geq$ 0.2).
+
+### Activity of signaling pathways and metabolic reactions
+
+Modulation of activity of signaling pathways between the collagen clusters was investigated by the _SPIA_ algorithm. 
+Differentially activated or inhibited pathways were identified by the FDR-corrected combined p value for enrichment and pathway perturbation (pGFDR). 
+The magnitude of signaling pathway modulation was measured with the tA parameter [@Tarca2009].
+
+Modulation of metabolic reactions of the Recon2 human metabolism model based on whole-genome differential gene expression was evaluated with the _BiGGR_ and [_biggrExtra_](https://github.com/PiotrTymoszuk/biggrExtra) R packages as described [@Pichler2023; @Gavai2015; @King2016].
+Statistical significance of reaction activity estimates in the collagen^high^ cluster as compared with the collagen^low^ subset was assessed by a Monte Carlo simulation with n = 1000 draws from normal distribution of gene expression.
+P values were corrected for multiple testing with the FDR method.
+
+Enrichment of significantly activated or inhibited biochemical reactions within the Recon model metabolic subsystems [@King2016] was investigated by FDR-corrected Fisher's exact test with odds ratio (OR) effect size statistic.
+Significantly activated and inhibited metabolic subsystems were defined by the significance cutoff (pFDR < 0.05) and at least weak effect size of the enrichment (OR $\geq$ 1.44).
+
+### Data and code availability
+
+Publicly available data sets were analyzed. 
+Cleared data sets used for analyses will be made available upon request to the corresponding author. 
+The transcriptome R analysis pipeline is available from GitHub (https://github.com/PiotrTymoszuk/collagen_pca).
+
+\newpage
+
 # Tables
@@
            value = as_paragraph('PSA: prostate-specific antigen.'), 
-           part = 'header', 
-           ref_symbols = 'a') %>%
-  theme_vanilla
-
-```
-
-\newpage
-
-```{r tab-genes, tab.cap = 'Collagen genes of interest and their classification.'}
-
-flextable::flextable(paper_tbl$genes) %>% 
-  set_widths(rep(4, 3)) %>% 
-  italic(j = 1, part = 'body') %>% 
-  footnote(1, 3, 
-           value = as_paragraph('ECM: extracellular matrix.'), 
            part = 'header', 
@@
 
-_(B) Differences in $log_2$-transformed expression levels of `r nrow(globals$genes_interest)` genes related to collagen metabolism between donor-matched tumor and benign prostate tissue were investigated by paired T test with Cohen's d effect size statistic in three published cohorts.
+_(B) Differences in $log_2$-transformed expression levels of `r nrow(globals$genes_interest)` genes related to collagen metabolism between donor-matched tumor and benign prostate tissue were investigated by paired T test with Cohen's d effect size statistic in the GSE70768 and TCGA cohort.
 P values were corrected for multiple testing with the false discovery method.
-Results for gene significantly regulated in at least two cohorts are presented.
-Lines represent means, 2 $\times$ SEM (standard error of the mean) intervals are visualized as tinted ribbons.
-Gene symbols, effect sizes and p values are shown in the Y axes.
-Significant effects are highlighted in bold.
-Numbers of tissue pairs are displayed in the plot captions._
+Normalized mean $log_2$-transformed expression levels in benign and cancer tissue for `r length(unique(unlist(norm_tumor$common)))` genes significantly regulated in both cohorts are presented in a heat map.
+Numbers of tissue pairs are displayed in the plot caption._
 
@@
 
-_Prostate cancers in the training TCGA cohort were clustered in respect to normalized $log_2$-transformed expression levels of the collagen pathway genes of interest by the PAM algorithm with Manhattan distance between the observations.
-Three clusters, collagen^low^ (low), collagen^intermediate^ (int) and collagen^high^ (hi) were defined.
+_Prostate cancers in the training TCGA cohort were clustered in respect to normalized $log_2$-transformed expression levels of the collagen pathway genes of interest by the PAM algorithm with cosine distance between the observations.
+Two clusters, collagen^low^ (low) and collagen^high^ (hi) were defined.
 Cancer samples from the test cohorts (`r globals$study_labels[globals$study_labels != 'TCGA'] %>% paste(collapse = ', ')`) were assigned to the collagen clusters with an inverse distance weighted k-nearest neighbor classifier._
 
-_(A) Mean normalized $log_2$-transformed expression levels of the collagen pathway genes in the collagen clusters and cohorts presented in a heat map._
+_(A) UMAP (uniform manifold approximation and projection) of the normalized $log_2$-transformed dataset of the collagen gene expression in the training TCGA cohort.
+Single cancer samples are visualized as points, the collagen cluster assignment is color coded.
+Numbers of samples in the clusters are indicated in the plot legend._
 
-_(B) Distribution of the collagen clusters expressed as percentage of the cohort cancer samples visualized in a stack plot. Total numbers of cancer samples are displayed in the X axis. _
+_(B) Mean normalized $log_2$-transformed expression levels of the collagen pathway genes in the collagen clusters and cohorts presented in a heat map.
+ECM: extracellular matrix, Pro: proline turnover, Adh.
+adhesion._
 
@@
 _(A) Counts of cancer-associated fibroblasts (CAF) and endothelial cells (EC) estimated by the MCP counter algorithm were found to differ significantly between the collagen clusters in all investigated collectives.
-Statistical significance was assessed by Kruskal-Wallis test with $\eta^2$ effect size statistic.
+Statistical significance was assessed by Mann-Whitney test with r effect size statistic.
 P values were corrected for multiple comparison with the false discovery rate method.
@@
 Effect sizes and p values are displayed in the plot captions.
-Numbers of observations in the clusters are indicated in the X axes._
+Numbers of observations in the clusters are indicated in the X axes.
+Full infiltration analysis results are presented in Supplementary Table S5._
 
 _(B) Gene set variation analysis (GSVA) with Reactome pathway gene signatures.
-Statistical significance for differences in signature single sample gene set enrichment analysis scores (ssGSEA) between the collagen clusters was determined by one-way ANOVA with $\eta^2$ effect size statistic.
+Statistical significance for differences in signature single sample gene set enrichment analysis scores (ssGSEA) between the collagen clusters was determined by two-tailed T test with Cohen's d effect size statistic.
 P values were corrected for multiple testing with the false discovery rate method.
-Mean ssGSEA scores in the clusters and cohorts for signatures significantly regulated in at least four cohorts are presented in a heat map.
-Full analysis results are presented in Supplementary Table S6._
+Mean ssGSEA scores in the clusters and cohorts for signatures significantly regulated in at least four cohorts with at least weak effect size (d $\geq$ 0.2) are presented in a heat map.
+Full analysis results are presented in Supplementary Table S7.
+ECM: extracellular matrix, GF: growth factor, GPCR: G protein-coupled receptor._
 
-_(C) Modulation of KEGG-listed signaling pathways in the collagen^high^ (hi) or collagen^intermediate^ (hi) as compared with collagen^low^ tumors was predicted by the SPIA algorithm based on differential gene expression estimates.
-Predicted magnitude of pathway modulation (tA parameter) for pathways significantly modulated in at least four cohorts was visualized in bubble plots (tA < 0: inhibition, blue; tA > 0, activation, red).
+_(C) Modulation of KEGG-listed signaling pathways in collagen^high^ (hi) tumors as compared with collagen^low^ cancers was predicted by the SPIA algorithm based on differential gene expression estimates.
+Predicted magnitude of pathway modulation (tA parameter) for pathways significantly modulated in at least four cohorts was visualized in a bubble plot (tA < 0: inhibition, blue; tA > 0, activation, red, ns: not significant, gray).
 Points are labeled with their tA values.
-Significant effects are highlighted in bold.
-Full analysis results are presented in Supplementary Table S8._
+Full analysis results are presented in Supplementary Table S9.
+ECM: extracellular matrix, ACT: actin, SCLC: small cell lung carcinoma, S.
+aureus: Staphylococcus aureus, SLE: systemic lupus erythematosus, TGF-$\beta$: transforming growth factor-beta.._
 
-_(D) Modulation of Recon2 model metabolic reactions in the collagen^high^ (hi) or collagen^intermediate^ (hi) as compared with collagen^low^ tumors was predicted by the BiGGR and biggrExtra algorithms based on differential gene expression estimates.
-Enrichment of significantly activated and significantly inhibited reactions in the collagen^high^ or collagen^intermediate^ clusters within the Recon model metabolic subsystems was investigated by Fisher's exact test.
+_(D) Modulation of Recon2 model metabolic reactions in collagen^high^ (hi) tumors as compared with collagen^low^ cancers was predicted by the BiGGR and biggrExtra algorithms based on whole-genome differential gene expression estimates.
+Enrichment of significantly activated and significantly inhibited reactions in the collagen^high^ within the Recon model metabolic subsystems was investigated by Fisher's exact test.
 Enrichment p values were corrected for multiple testing with the false discovery rate method.
-Enrichment odds ratios for activated (red) and inhibited reaction (blue) for significantly enriched metabolic subsystems shared by at least four cohorts are presented in bubble plots.
-Points are labeled with their odds ration values.
-Full analysis results are presented in Supplementary Table S10._
+Enrichment odds ratios (OR) for significantly enriched metabolic subsystems with at least weak effect size (OR $\geq$ 1.44)  shared by at least four cohorts are presented in a bubble plot (inhibited: blue, activated: red, not significant: gray).
+Points are labeled with their odds ratio values.
+Full analysis results are presented in Supplementary Table S11.
+Val: valine, Leu: leucine, Ile: isoleucine, FAOX: fatty acid oxidation, PIP: phosphoinositol phosphate._
 

### 2023-08-07 report/markdown/manuscript_supplement.Rmd
Figures, Tables, Supplementary Material

--- 2023-07-27 report/markdown/manuscript_supplement.Rmd
@@
 
+# Supplementary Methods
+
+## Software
+
+The analysis was done with R version 4.2.3 (R Foundation).
+
+Tabular data were handled with the packages _tidyverse_ [@Wickham2019], _rlang_ [@Henry2022] and [_trafo_](https://github.com/PiotrTymoszuk/trafo). 
+Text data were handled with _stringi_ [@Gagolewski2021].
+
+Import of the TCGA dataset was accomplished with the [_TCGA-Assembler-2_ script](https://github.com/compgenome365/TCGA-Assembler-2/tree/master/TCGA-Assembler). 
+Transcriptome datasets from the Gene Expression Omnibus were fetched with the _GEOquery_ package [@Sean2007]. 
+Gene and probe annotation was accomplished with the _AnnotationDbi_ [@Pages2022] and _org.Hs.eg.db_ packages [@Carlson2022]. 
+For prediction of non-malignant cell counts and fractions in cancer samples, the R implementation of the _MCP-counter_ and _xCell_ algorithms provided by the _immunedeconv_ package was used [@Sturm2020; @Aran2017; @Becht2016].
+Single sample gene set enrichment analysis scores (ssGSEA) were computed with the _GSVA_ algorithm [@Hanzelmann2013] implemented by the development package [_gseaTools_](https://github.com/PiotrTymoszuk/gseaTools).
+Expression data of collegen genes were adjusted for batch effects with the _COMBAT_ algorithm implemented by the _sva_ package [@Leek2012].
+
+Semi supervised clustering was accomplished with the development package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) employing algorithms provided by the _philentropy_, _cluster_ and _factorextra_ packages [@Kassambara2020; @Schubert2019; @Drost2018; @Konopka2022].
+
+For statistical hypothesis testing, effect size calculation, gene set variation analysis (GSVA) and differential gene expression analysis, the packages _rstatix_ [@Kassambara2021], [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) and [_microViz_](https://github.com/PiotrTymoszuk/microViz) were employed.
+Signaling pathway analysis was done with the package _SPIA_ [@Tarca2009].  
+Metabolic reaction activity modeling and metabolic subsystem enrichment analysis were performed with the packages _BiGGR_ [@Gavai2015; @King2016] and [_biggrExtra_](https://github.com/PiotrTymoszuk/biggrExtra).
+
+For visualization of the results, the packages 
+_ggplot2_ [@Wickham2019] (scatter, bubble and bar plots, heat maps), 
+[_ExDA_](https://github.com/PiotrTymoszuk/ExDA) (violin, stack and ribbon plots), 
+[_microViz_](https://github.com/PiotrTymoszuk/microViz) (Forest plots) and 
+[_biggrExtra_](https://github.com/PiotrTymoszuk/biggrExtra) (fold-regulation estimates of metabolic reactions). 
+Figures and tables were created with the packages _cowplot_ [@Wilke2019] and _flextable_ [@Gohel2022].
+
+## Data sources and data import
+
+The TCGA prostate cancer dataset [@Abeshouse2015] consisting of 
+`r study_data$tcga$expression %>% filter(tissue == 'tumor') %>% nrow` normalized expression (level 3) data for primary cancer samples and clinical information was imported in R from the GDC repository with the [_TCGA-Assembler-2_ script](https://github.com/compgenome365/TCGA-Assembler-2/tree/master/TCGA-Assembler).
+The TCGA cohort included also 
+`r study_data$tcga$expression %>% filter(tissue == 'benign') %>% nrow` donor-matched samples of the benign prostate and prostate cancer tissue. 
+The available clinical information for the TCGA cohort included age, pathological staging for the tumor, node and distant metastases, tumor grade, Gleason scoring, blood concentrations of prostate-specific antigen (PSA) and diagnosis, overall and biochemical relapse-free survival.
+The GSE16560, GSE70768, GSE70769 and GSE116918 normalized whole-transcriptome datasets deposited by the study authors were fetched from GEO with the _GEOquery_ package (function `getGEO()`) [@Sean2007].
+Transcriptomes of the GSE16560 primary cancer samples (n = `r study_data$GSE16560$expression %>% nrow`) [@Sboner2010] were measured with a custom microarray developed by Illumina. 
+The accompanying clinical information included age, Gleason score and overall survival. 
+Transcriptome measurements in primary cancer samples (n = `r study_data$GSE70768$expression %>% filter(tissue == 'tumor') %>% nrow`) in the GSE70768 cohort were done with the Illumina HumanHT-12 V4.0 expression beadchip microarray.
+The GSE70768 dataset [@Ross-Adams2015] included also `r study_data$GSE70768$expression %>% filter(tissue == 'benign') %>% nrow` donor-matched normal prostate and cancer tissue specimens. 
+The clinical information for the GSE70768 dataset encompassed age, clinical and pathological staging of the tumor, lymph nodes and distant metastases, PSA concentrations at diagnosis, data on extra capsular extension and positive surgery margins, Gleason scoring as well as biochemical relapse-free survival.
+The GSE70769 dataset [@Ross-Adams2015] consisted of `r study_data$GSE70769$expression %>% nrow` primary cancer specimens, whose transcriptomes were gauged with Illumina HumanHT-12 V4.0 expression beadchip.
+The clinical information for the GSE70769 dataset encompassed clinical and pathological staging of the tumor, lymph nodes and distant metastases, PSA concentrations at diagnosis, data on extra capsular extension and positive surgery margins, Gleason scoring as well as biochemical relapse-free survival.
+The GSE116918 dataset [@Jain2018] included `r study_data$GSE116918$expression %>% nrow` prostate cancer biopsy specimens. 
+The transcriptome measurements were performed with a custom-made microarray by Affymetrix (Almac Diagnostics Prostate Disease Specific Array). 
+The accompanying clinical information consisted of age, blood PSA concentrations at diagnosis, pathological staging of the tumor, Gleason scoring and biochemical relapse-free survival.
+Clinical characteristic of the analyzed cohorts is provided in __Table 1__.
+
+Gene expression levels were transformed with the $log_2(expression)$ and $log_2(FKPM + 1)$ function for the microarray and RNA sequencing datasets prior to the analysis. 
+Prior to semi-supervised clustering, inter-cohort variability in normalized $log_2$-transformed expression of collagen pathway genes was reduced by with the `ComBat()` function from the package _sva_ [@Leek2012].
+
+The MCP-counter and xCell estimates of non-malignant cell content in cancer samples were computed with the `deconvolute()` function from the _immunedeconv_ package. 
+Gene signatures of the [Reactome pathways](https://reactome.org/) were obtained from the [MSig database](https://www.gsea-msigdb.org/gsea/msigdb/index.jsp), version 7.5.1. 
+ssGSEA scores were computed with the `calculate()` function from the [_gseaTools_](https://github.com/PiotrTymoszuk/gseaTools) package.
+
+The collagen pathway genes investigated in the current report included the published Collagen Signature [@Kocher2021] and the Reactome Collagen formation pathway genes (R-HSA-1474290) and was constrained to `r nrow(globals$genes_interest)` genes detected in all expression datasets (__Supplementary Table S\@ref(tab:tab-genes)__).
+
+## Statistical hypothesis testing, effect size and statistical significance
+
+Differences in numeric variables were investigated by paired and unpaired two-tailed T tests with Cohen's d effect size statistic or Mann-Whitney tests with r effect size statistic. 
+Differences in distribution of categorical variables were assessed by $\chi^2$ test with Cramer's V effect size statistic. 
+Statistical significance for enrichment of regulated metabolic reactions in metabolic subsystems was investigated by Fisher's exact test with odds ratio (OR) as an effect size statistic. 
+Effect sizes were interpreted as follows [@Cohen2013]:
+
+* Cohen's d, weak: 0.2 - 0.5, moderate: 0.5 - 0.8, large: $\geq$ 0.8
+
+* r, weak: 0.1 - 0.3, moderate: 0.3 - 0.5, large: $\geq$ 0.5
+
+* Cramer's V, weak: 0.1 - 0.3, moderate: 0.3 - 0.5, large: $\geq$ 0.5
+
+* OR, weak: 1.44 - 2.48, moderate: 2.48 - 4.27, large: $\geq$ 4.27
+
+P values were corrected for multiple testing with the false discovery rate method (FDR) [@Benjamini1995] within each analysis step and cohort. 
+Effects were considered statistically significant for FDR-corrected p values < 0.05.
+
+## Comparison of collagen pathway gene expression between the benign and cancer tissue
+
+Differences in $log_2$-transformed expression levels of the collagen pathway genes between donor-matched cancer and benign prostate specimens were investigated with paired two-tailed T test with Cohen's effect size statistic (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)) in the GSE70768 and TCGA cohorts.
+The analysis results are presented in __Figure 1B__, __Supplementary Figure S\@ref(fig:fig-norm-tumor)__ and __Supplementary Table S\@ref(tab:tab-norm-tumor)__.
+
+## Collagen clusters of prostate cancer samples
+
+Prostate cancer samples in the TCGA cohort were clustered by their COMBAT-adjusted normalized $log_2$-transformed expression levels of the collagen pathway genes with the PAM (partition around medoids) algorithm with cosine distance between the samples (function `kcluster()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)).
+The choice of the clustering algorithm was motivated by the trade-off between the high explanatory performance measured by 'explained' clustering variance (ratio of the total between-cluster sum of squares to the total sum of squares) and the optimal reproducibility in 10-fold cross-validation (mean rate of the correct cluster assignment in the folds) [@Lange2004; @Leng2014] as compared with hierarchical clustering and KMEANS procedures (__Supplementary Figure S\@ref(fig:fig-clust-dev)A__).
+For variance computation and cross-validation, the methods `var()` and `cv()` provided by the [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) package were used.
+The cluster number choice was based on the bend of the curve of within-cluster sum of squares [@Kassambara2020] and the peak of the average silhouette statistic [@Rousseeuw1987] (method `plot()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools), __Supplementary Figure S\@ref(fig:fig-clust-dev)B__).
+By this means, two clusters of the samples were identified: 'collagen^high^' and 'collagen^low^' cancers.
+
+Assignment of cancers samples of the GSE16560, GSE70768, GSE70769 and GSE116918 collectives to the collagen clusters based on COMBAT-adjusted normalized $log_2$-transformed expression levels of the collagen pathway genes was done with an inverse distance weighted k-nearest neighbor classifier (method `predict()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) [@Leng2014].
+The choice of the nearest neighbor number k was motivated by the maximum value of the 'explained' clustering variance defined above (
+GSE16560: k = `r coll_clust$semi_clust$GSE16560$kNN`, 
+GSE70768: k = `r coll_clust$semi_clust$GSE70768$kNN`, 
+GSE70769: k = `r coll_clust$semi_clust$GSE70769$kNN`, 
+GSE116918: k = `r coll_clust$semi_clust$GSE116918$kNN`). 
+The quality of the cluster assignment was assessed by comparing the collagen clusters distribution and 'explained' clustering variances between the training TCGA datasets and the test cohorts (__Supplementary Figure S\@ref(fig:fig-clust-dev)C__).
+Separation of the clusters was assessed by a visual inspection of the UMAP layout plots (method `plot()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) and heat maps of the mean levels of the clustering factors in the collagen clusters (__Figure 2AB__).
+
+Differences in the $log_2$-transformed collagen pathway gene expression between the collagen clusters were investigated by two-tailed T test with Cohen's d effect size statistic. 
+The comparison results are presented in __Supplementary Figure S\@ref(fig:fig-clusters)__ and __Supplementary Table S\@ref(tab:tab-cluster)__.
+
+## Clinical characteristic of the collagen clusters
+
+Differences in numeric clinical variables (age and PSA) between the collagen clusters were assessed by Mann-Whitney test with r effect size statistic. 
+Differences in qualitative clinical variables between the collagen clusters were investigated by $\chi^2$ test with Cramer's V effect size statistic (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)).
+The clinical characteristics of the collagen clusters are presented in __Figure 2CD__ and __Supplementary Table S\@ref(tab:tab-clinic)__).
+
+## Non-malignant cell infiltration in the collagen clusters
+
+Non-malignant cell counts and non-malignant cell fractions in the cancer samples were predicted by the _MCP-counter_ [@Becht2016] and _xCell_ [@Aran2017] algorithms, respectively.
+Differences in the predicted infiltration levels between the collagen clusters were investigated by Mann-Whitney test with r effect size statistic (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)).
+The analysis results are shown in __Figure 3A__, __Supplementary Figures S\@ref(fig:fig-mcp-infil)__ - __S\@ref(fig:fig-infil-xcell)__ and __Supplementary Tables S\@ref(tab:tab-mcp)__ - __S\@ref(tab:tab-xcell)__.
+
+## Gene set variance analysis of Reactome pathway gene signatures
+
+ssGSEA scores [@Hanzelmann2013] of the Reactome pathway gene signatures were compared between the collagen clusters by two-tailed T test with Cohen's d effect size statistic. 
+Signatures found to be significantly regulated (pFDR < 0.05) with at least weak effect size (d $\geq$ 0.2) in at least four cohort were further investigated. 
+The analysis results are presented in __Figure 3B__, __Supplementary Figure S\@ref(fig:fig-top-gsva)__ and __Supplementary Table S\@ref(tab:tab-gsva)__.
+
+## Differential gene expression in the collagen clusters
+
+Differences in whole-transcriptome gene expression between the collagen clusters were investigated by FDR-corrected two-tailed T test with Cohen's d effect size statistic (function `test_two_groups()`, package [_microViz_](https://github.com/PiotrTymoszuk/microViz)).
+Genes significantly differentially expressed between the clusters were identified by the significance cutoff (pFDR < 0.05) and at least weak effect size of the expression differences (d $\geq$ 0.2). 
+Percentages of the investigated transcriptomes found to be significantly differentially expressed between the collagen clusters are displayed in __Supplementary Figure S\@ref(fig:fig-dge-numbers)__. 
+Expression regulation estimates for the strongest up- and downregulated genes in collagen^high^ vs collagen^low^ cancers are presented in __Supplementary Figure S\@ref(fig:fig-top-dge)__. 
+Significantly differentially regulated genes are listed in __Supplementary Table S\@ref(tab:tab-dge)__.
+
+## Signaling pathway activity modulation in the collagen clusters
+
+Modulation of activity of KEGG-listed signaling pathways was predicted by the _SPIA_ algorithm (function `spia()`) [@Tarca2009] based on $log_2$ fold-regulation estimates of expression regulation for the significantly differentially regulated genes.
+The `tA` parameter served as a measure of the predicted pathway activity (inhibited: tA < 0, activated: tA > 0). 
+Statistical significance of the activity modulation was assessed by the combined enrichment/perturbation p value corrected for multiple testing with the FDR method (pGFDR) [@Tarca2009]. 
+Common modulated signaling pathways were defined as signaling pathways predicted to be significantly activated or inhibited in at least four cohorts. 
+The analysis results are presented in __Figure 3C__ and __Supplementary Table S\@ref(tab:tab-signaling)__.
+
+## Activity of metabolic reactions in the collagen clusters
+
+Rules of assignment of genes to biochemical reactions were retrieved from the Recon2 human metabolism model available via the BiGG database [@King2016] and the R package _BiGGR_ [@Gavai2015]. 
+Estimates of differential expression for all available genes between the collagen clusters were calculated by two-tailed T test as describe above. 
+Estimates of biochemical pathway fold-regulation were computed by evaluation of the gene assignment rules in the Recon2 model. 
+The 'gene~A~ OR gene~B~' operator was interpreted as arithmetic mean of expression regulation estimates for the genes A and B. 
+The 'gene~A~ AND gene~B~' operator was interpreted as minimum of expression regulation estimates for the gene A and gene B [@Gavai2015]. 
+Standard deviation, 95% confidence intervals and p values for the predicted reaction regulation estimates were obtained by a Monte Carlo simulation with n = 1000 draws from normal distribution of gene expression regulation estimates (mean: expression regulation estimate, standard deviation: standard error of the expression regulation estimate) [@Gavai2015; @Pichler2023].
+P values were corrected for multiple testing with the FDR method. 
+The analysis was done with the package _BiGGR_ [@Gavai2015] and the development package [_biggrExtra_](https://github.com/PiotrTymoszuk/biggrExtra) [@Pichler2023] (function `build_geneSBML()`). 
+Biochemical reactions predicted to be significantly modulated between the collagen clusters are listed in __Supplementary Table S\@ref(tab:tab-reactions)__.
+
+Enrichment of significantly activated or inhibited biochemical reactions within the Recon model metabolic subsystems [@King2016] was investigated by FDR-corrected Fisher's exact test (function `suba()`, package [_biggrExtra_](https://github.com/PiotrTymoszuk/biggrExtra)).
+The magnitude of enrichment was assessed with OR defined for the $i$-th metabolic subsystem as follows:
+
+$$OR_i = \frac{n_{regulated, i} \times n_{total}}{n_i \times n_{regulated}}$$
+
+where $n_{regulated, i}$ denotes the number of significantly regulated (activated or inhibited) biochemical reactions within the $i$-th metabolic subsystem, 
+$n_{total}$ denotes the total number of investigated reactions, 
+$n_i$ denotes the number of metabolic reactions within the $i$-th metabolic subsystem, and 
+$n_{regulated}$ is the total number of significantly regulated biochemical reactions. 
+Common significantly activated or inhibited metabolic subsystems were identified as metabolic subsystems significantly activated or inhibited in at least four cohorts, with at least weak effect size defined as OR $\geq$ 1.44 [@Cohen2013] (__Figure 3D__, __Supplementary Table S\@ref(tab:tab-subsystems)__).
+
+## Data and code availability
+
+Publicly available datasets were analyzed. 
+Cleared datasets used for analyses will be made available upon request to the corresponding author. 
+The transcriptome R analysis pipeline is available from GitHub (https://github.com/PiotrTymoszuk/collagen_pca).
+
+\newpage
+
 # Supplementary Tables
+
+```{r tab-genes, tab.cap = 'Investigated collagen pathway genes and their classification.'}
+
+suppl_paper_tbl$genes %>% 
+  flextable %>% 
+  set_widths(widths = c(4.5, 3.5, 3)) %>% 
+  italic(j = 2, part = 'body') %>% 
+  merge_v(1) %>% 
+  theme_vanilla
+
+```
+
+\newpage
 
@@
 
-```{r tab-cluster, tab.cap = 'Expression of the cluster-defining collagen pathway genes in the collagen clusters of prostate cancer.
-Statistical significance was assessed by one-way ANOVA with eta-squared effect size statistic.
+```{r tab-cluster, tab.cap = "Expression of the cluster-defining collagen pathway genes in the collagen clusters of prostate cancer.
+Statistical significance was assessed by two-tailed T test with Cohen's d effect size statistic.
 P values were corrected for multiple testing with the false discovery rate method.
 log2-transformed expression values are presented as medians with interquartile ranges (IQR) and ranges.
-The table is available as a supplementary Excel file.'}
+The table is available as a supplementary Excel file."}
 
@@
 flextable::flextable(suppl_paper_tbl$clinic) %>% 
-  set_widths(c(2.2, 3, rep(3.5, 3), 2.8, 2.8)) %>% 
+  set_widths(c(2.2, 3, rep(4, 2), 2.8, 2.8)) %>% 
   merge_v(1:2) %>% 
@@
            ref_symbols = 'a') %>% 
-  footnote(1, 6:7, 
+  footnote(1, 5:6, 
            value = as_paragraph("Categorical variables: \u03C7\u00B2 test with Cramer's V effec size statistic.
-Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic.
+Numeric variables: Mann-whitney test with r effect size statistic.
 P values corrected for multiple testing with the false discovery rate method."),
@@
 
-```{r tab-mcp, tab.cap = 'Non-malignant cell numbers predicted for the collagen Non-maignant cell fractions predicted for the collagen Statistical significance was assessed by Kruskal-Wallis test with eta-squared effect size statistic.
+```{r tab-mcp, tab.cap = 'Non-malignant cell numbers predicted for the collagen clusters by the MCP counter algorithm.
+Statistical significance was assessed by Mann-Whitney test with r effect size statistic.
 P values were corrected for multiple testing with the false discovery method.
@@
 
-```{r tab-xcell, tab.cap = 'clusters by the MCP counter algorithm.
-clusters by the xCell algorithm.
-Statistical significance was assessed by Kruskal-Wallis test with eta-squared effect size statistic.
+```{r tab-xcell, tab.cap = 'Non-maignant cell fractions predicted for the collagen clusters by the xCell algorithm.
+Statistical significance was assessed by Mann-Whitney test with r effect size statistic.
 P values were corrected for multiple testing with the false discovery method.
@@
 
-```{r tab-gsva, tab.cap = 'Gene set variation analysis with the Reactome pathway gene signatures.
-Differences between the collagen intermediate or high clusters versus collagen low cancers were investigated by one-way ANOVA with eta-squared effect size statistic and linear modeling.
-Results for signatures significantly regulated with moderate-to-large effect size (eta-squared at leat 0.06) in at least four cohorts are presented.
+```{r tab-gsva, tab.cap = "Gene set variation analysis with the Reactome pathway gene signatures.
+Differences in ssGSEA scores between collagen high and collagen low cancers were investigated by two-tailed T test with Cohen's d effect size statistic.
+Results for signatures significantly regulated with at least weak effect size (d at least 0.2) in at least four cohorts are presented.
 P values were corrected for multiple testing with the false discovery rate method (FDR).
-The table is available as a supplementary Excel file.'}
+The table is available as a supplementary Excel file."}
 
@@
 
-```{r tab-dge, tab.cap = 'Genes differentially expressed in the collagen intermediate or high cluster as compared with collagen low cancers were identified by one-way ANOVA and linear modeling with the 1.25-fold regulation cutoff P values were corrected for multiple testing with the false discovery rate method (FDR).
-The table is available as a supplementary Excel file.'}
+```{r tab-dge, tab.cap = "Genes differentially expressed in the collagen high cluster as compared with collagen low cancers were identified by two-tailed T test with the 1.25-fold regulation cutoff and the Cohen's d effect size statistic of at least 0.2 (at least weak effect size).
+P values were corrected for multiple testing with the false discovery rate method (FDR).
+The table is available as a supplementary Excel file."}
 
@@
 ```{r tab-signaling, tab.cap = 'Signaling pathway activity in the collagen clusters investigated by the SPIA algorithm.
-Resulat for signaling pathways significantly activated or inhibited in at least four cohorts are shown.
+Results for signaling pathways significantly activated or inhibited in at least four cohorts are shown.
 The table is available as a supplementary Excel file.'}
@@
 
-```{r tab-reactions, tab.cap = 'Biochemical reactions predicted to be significantly activated in the collagen high or collagen low cluster as compared with collagen low cancers.
-Statistical significance was determined by Monte Carlo simulation.
+```{r tab-reactions, tab.cap = 'Biochemical reactions predicted to be significantly activated in the collagen high as compared with collagen low cancers.
+Statistical significance was determined by a Monte Carlo simulation.
 P values were corrected for multiple testing with the false discovery rate method.
@@
 
+```{r fig-norm-tumor, fig.width = figur::convert(suppl_paper_fig$norm_tumor, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$norm_tumor, to = 'in')$h, fig.cap = 'Expression of collagen pathway genes in the normal prostate and prostate cancer tissue.'}
+
+suppl_paper_fig$norm_tumor$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-norm-tumor). Expression of collagen pathway genes in the normal prostate and prostate cancer tissue.__
+
+_Differences in $log_2$-transformed expression of `r nrow(globals$genes_interest)` genes related to collagen metabolism between donor-matched pairs of the prostate cancer and benign tissue were assessed by paired T test with Cohen's effect size statistic in the GSE70768 and TCGA cohorts.
+P values were corrected for multiple testing with the false discovery rate (FDR) method._
+
+_(A) Numbers of significantly up- and downregulated genes in the tumor tissue as compared with the benign tissue in the investigated cohorts presented in Venn plots._
+
+_(B) Expression of the collagen pathway genes significantly regulated in both analyzed cohorts.
+Mean normalized $log_2$-transformed expression values are presented as lines.
+Tinted ribbons represent the 2 $\times$ SEM (standard error of the mean) intervals.
+Effect sizes and p values are indicated in the Y axis._
+
+\newpage
+
+```{r fig-clust-dev, fig.width = figur::convert(suppl_paper_fig$clust_dev, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$clust_dev, to = 'in')$h, fig.cap = 'Semi-supervised clustering of prostate cancer samples in respect to expression of the collagen pathway genes.'}
+
+suppl_paper_fig$clust_dev$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-clust-dev). Semi-supervised clustering of prostate cancer samples in respect to expression of the collagen pathway genes.__
+
+_Tumor samples in the TCGA training cohort were clustered in respect to normalized, $log_2$-transformed expression levels of the collagen pathway genes with the PAM (partition around medoids) algorithm with cosine distance metric.
+Two clusters were defined: collagen^low^ (low) and collagen^high^ (hi).
+Assignment of the tumor samples in the training GSE16560, GSE40272, GSE70768 and GSE70769 collectives to the collagen clusters was accomplished by an inverse distance-weighted k-nearest neighbor classifier._
+
+_(A) Comparison of performance of several clustering algorithms (PAM: partition around medoids, HCL: hierarchical clustering/Ward D2 and KMEANS) and distance metrics (Euclidean, Manhattan, sum-of-squares and cosine) in the training TCGA cohort was measured by the amount of 'explained' clustering variance (ratio of the between-cluster sum of squares to the total sum of squares) and rate of correct cluster assignment in 10-fold cross-validation (CV).
+Number of complete observations and the clustering factors are shown next to the plot._
+
+_(B) Determination of the cluster number for PAM/Manhattan distance algorithm in the TCGA cohort by the bend of the within-cluster sum of squares (WSS) (B) and the peak of the mean silhouette statistic (C)._
+
+_(C) Distribution of the collagen clusters (left panel) and performance of the PAM/cosine distance algorithm at semi-supervised clustering in the training cohort and the test collectives measured by the 'explained' clustering variance statistic (right panel).
+Numbers of complete observations are presented in the plot axes._
+
+\newpage
+
+```{r fig-clusters, fig.width = figur::convert(suppl_paper_fig$clusters, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$clusters, to = 'in')$h, fig.cap = 'Expression of the collagen pathway genes in the collagen clusters of prostate cancers.'}
+
+suppl_paper_fig$clusters$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-clusters). Expression of the collagen pathway genes in the collagen clusters of prostate cancers.__
+
+_Tumor samples were subjected to semi-supervised clustering in respect to the collagen pathway gene expression with the PAM algorithm with cosine distance metric (training: TCGA, test cohorts: GSE16560, GSE40272, GSE70768 and GSE70769).
+Statistical significance of differences in expression of normalized, $log_2$-transformed expression levels (Z-scores) between the collagen clusters were investigated by two-tailed T test with Cohen's d effect size statistic.
+P values were corrected for multiple testing with the false discovery rate method.
+Mean expression Z scores in the collagen^low^ (low) and collagen^high^ (hi) clusters are presented as thick lines.
+The tinted ribbons represent 2$\times$ SEM (standard error of the mean) intervals.
+The collagen pathway genes were grouped according to their biological function as proline turnover (Pro), collagen modification, extracellular matrix (ECM) component, ECM processing and adhesion (Adh.) protein genes.
+Significantly regulated genes are highlighted in bold.
+Numbers of samples in the clusters are indicated in the plot captions._
+
+\newpage
+
+```{r fig-mcp-infil, fig.width = figur::convert(suppl_paper_fig$mcp_infil, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$mcp_infil, to = 'in')$h, fig.cap = 'MCP Counter estimates of non-malignant cell content in the collagen clusters.'}
+
+suppl_paper_fig$mcp_infil$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-mcp-infil). MCP Counter estimates of non-malignant cell content in the collagen clusters.__
+
+_Counts of non-malignant cells (CAF: cancer-associated fibroblast, EC: endothelial cells, T cells, CD8^+^ T cells, NK cells: natural killer cells, B cells, macrophages, monocytes, neutrophils and mDC: myeloid dendritic cells) in the prostate cancer samples were predicted by the MCP counter algorithm.
+Differences in cell content between the collagen^low^ (low) and collagen^high^ (hi) clusters were analyzed with Mann-Whitney test with r effect size statistic.
+P values were corrected for multiple testing with the false discovery rate method.
+Means of normalized cell counts in the collagen clusters are depicted as thick lines.
+Tinted ribbons represent the 2 $\times$ SEM (standard error of the mean) intervals.
+Effect sizes and p values are shown in the Y axes.
+Significant effects are highlighted in bold.
+Numbers of samples assigned to the clusters are indicated in the plot captions._
+
+\newpage
+
+```{r fig-infil-mcp, fig.width = figur::convert(suppl_paper_fig$infil_mcp, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$infil_mcp, to = 'in')$h, fig.cap = 'Counts of cancer-associated fibroblast and endothelial cell in tumor samples in the  collagen clusters predicted by the MCP counter algorithm.'}
+
+suppl_paper_fig$infil_mcp$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-infil-mcp). Counts of cancer-associated fibroblast and endothelial cell in tumor samples in the  collagen clusters predicted by the MCP counter algorithm.__
+
+_Counts of cancer associated fibroblasts (CAF, A) and endothelial cells (EC, B) in collagen^low^ (low) and collagen^high^ (hi) prostate cancers were predicted by the MCP counter algorithm.
+Differences between the collagen clusters were assessed by Mann-Whitney test with r effect size statistic.
+Cell counts were presented in violin plots with single observations visualized as points.
+Red diamonds and whiskers represent medians with interquartile ranges.
+Effect sizes and p values are displayed in the plot captions.
+Numbers of samples in the clusters are indicated in the X axes._
+
+\newpage
+
+```{r fig-infil-xcell, fig.width = figur::convert(suppl_paper_fig$infil_xcell, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$infil_xcell, to = 'in')$h, fig.cap = 'Fractions of cancer-associated fibroblast and endothelial cell in tumor samples in the  collagen clusters predicted by the xCell algorithm.'}
+
+suppl_paper_fig$infil_xcell$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-infil-xcell). Fractions of cancer-associated fibroblast and endothelial cell in tumor samples in the  collagen clusters predicted by the xCell algorithm.__
+
+_Fractions of cancer associated fibroblasts (CAF, A) and endothelial cells (EC, B) in the malignant tissue of collagen^low^ (low) and collagen^high^ (hi) prostate cancers were predicted by the xCell counter algorithm.
+Differences between the collagen clusters were assessed by Mann-Whitney test with r effect size statistic.
+Cell counts were presented in violin plots with single observations visualized as points.
+Red diamonds and whiskers represent medians with interquartile ranges.
+Effect sizes and p values are displayed in the plot captions.
+Numbers of samples in the clusters are indicated in the X axes._
+
+\newpage
+
+```{r fig-top-gsva, fig.width = figur::convert(suppl_paper_fig$top_gsva, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$top_gsva, to = 'in')$h, fig.cap = 'Gene set variation analysis results for representative Reactome pathway gene signatures differentiating between the collagen clusters.'}
+
+suppl_paper_fig$top_gsva$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-top-gsva). Gene set variation analysis results for representative Reactome pathway gene signatures differentiating between the collagen clusters.__
+
+_Gene set variation analysis of the collagen clusters was performed with the Reactome pathway gene signatures.
+Statistical significance for differences in signature single sample gene set enrichment analysis scores (ssGSEA) between the collagen clusters was determined by two-tailed T test with Cohen's d effect size statistic.
+P values were corrected for multiple testing with the false discovery rate method.
+Common regulated signatures were defined as signatures significantly regulated with at least weak effect size (d $\geq$ 0.2) in at least four cohorts (Supplementary Table S\@ref(tab:tab-gsva)).
+Mean ssGSEA scores in the clusters and cohorts for the top five strongest regulated signatures for each functional category are presented in a heat map._
+_ECM: extracellular matrix, GF: growth factor, GPCR: G protein-coupled receptor._
+
+\newpage
+
+```{r fig-dge-numbers, fig.width = figur::convert(suppl_paper_fig$dge_numbers, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$dge_numbers, to = 'in')$h, fig.cap = 'Percentages of the analyzed transcriptome significantly up- and downregulated in collagen high vs collagen low cluster cancers.'}
+
+suppl_paper_fig$dge_numbers$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-dge-numbers). Percentages of the analyzed transcriptome significantly up- and downregulated in collagen high vs collagen low cluster cancers.__
+
+_Genes significantly differentially regulated in collagen^high^ cancers (hi) as compared with collagen^low^ prostate tumors (low) were identified by false discovery rate corrected two-tailed T test (pFDR < 0.05) with Cohen's d effect size statistic (d $\geq$ 0.2).
+Percentages of significantly up- and downregulated genes are presented in bar plots.
+Total numbers of analyzed genes are indicated in the Y axes._
+
+\newpage
+
+```{r fig-top-dge, fig.width = figur::convert(suppl_paper_fig$top_dge, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$top_dge, to = 'in')$h, fig.cap = 'Top strongest differentially expressed genes in the collagen high cluster as compared with collagen low cancers.'}
+
+suppl_paper_fig$top_dge$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-top-dge). Top strongest differentially expressed genes in the collagen high cluster as compared with collagen low cancers.__
+
+_$log_2$ fold-regulation estimates of differences in expression between collagen^high^ (hi) and collagen^low^ (low) cancers with 95% confidence intervals for top 20 strongest up- and downregulated genes were presented in Forest plots.
+Numbers of samples in the clusters are displayed in the plot captions._
+
+\newpage
+
+```{r fig-reaction-counts, fig.width = figur::convert(suppl_paper_fig$reaction_counts, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$reaction_counts, to = 'in')$h, fig.cap = 'Numbers of significantly activated and inhibited biochemical reactions predicted for the collagen high cluster as compared with collagen low cancers.'}
+
+suppl_paper_fig$reaction_counts$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-reaction-counts).
+Numbers of significantly activated and inhibited biochemical reactions predicted for the collagen high cluster as compared with collagen low cancers.__
+
+_Modulation of Recon2 model metabolic reactions in collagen^high^ cancers (hi) as compared with collagen^low^ tumors (low) was predicted by the BiGGR and biggrExtra algorithms based on whole-genome differential gene expression estimates.
+Statistical significance was determined by Monte Carlo simulation with n = 1000 draws from normal distributions of the gene expression regulation estimates.
+P values were corrected for multiple testing with the false discovery rate method.
+Percentages of predicted significantly activated and inhibited metabolic reactions are shown in bar plots.
+The total reaction number is displayed in the plot caption._
+
+\newpage
+
+```{r fig-faox, fig.width = figur::convert(suppl_paper_fig$faox, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$faox, to = 'in')$h, fig.cap = 'Activity of fatty acid oxidation and citric acid cycle reactions predicted for the collagen high cluster as compared with collagen low cancers.'}
+
+suppl_paper_fig$faox$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-faox). Activity of fatty acid oxidation and citric acid cycle reactions predicted for the collagen high cluster as compared with collagen low cancers.__
+
+_Regulation of reactions of the fatty acid oxidation (FAOX, A) and citric acid cycle (tri-carboxylic acid cycle/TCA, B) Recon subsystems in collagen^high^ cancers (hi) as compared with collagen^low^ tumors (low) was predicted by the BiGGR and biggrExtra algorithms based on whole-genome differential gene expression estimates.
+Statistical significance was determined by Monte Carlo simulation with n = 1000 draws from distributions of the gene expression estimates.
+P values were corrected for multiple testing with the false discovery rate method.
+Estimates of $log_2$ fold-regulation of reaction activity are presented in bar plots.
+Activity status is color coded.
+Numbers of significantly activated and inhibited reactions are displayed in the plot captions.
+Total reaction numbers in the investigated metabolic subsystem are indicated in the X axes._
+
+\newpage
+
+```{r fig-extrans, fig.width = figur::convert(suppl_paper_fig$extrans, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$extrans, to = 'in')$h, fig.cap = 'Activity of chondroitin synthesis and extracellular transport reactions predicted for the collagen high cluster as compared with collagen low cancers.'}
+
+suppl_paper_fig$extrans$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-extrans). Activity of chondroitin synthesis and extracellular transport reactions predicted for the collagen high cluster as compared with collagen low cancers.__
+
+_Regulation of reactions of the chondroitin synthesis (CS, A) and extracellular transport (ExTrans, B) Recon subsystems in collagen^high^ cancers (hi) as compared with collagen^low^ tumors (low) was predicted by the BiGGR and biggrExtra algorithms based on whole-genome differential gene expression estimates.
+Statistical significance was determined by Monte Carlo simulation with n = 1000 draws from distributions of the gene expression estimates.
+P values were corrected for multiple testing with the false discovery rate method.
+Estimates of $log_2$ fold-regulation of reaction activity are presented in bar plots.
+Activity status is color coded.
+Numbers of significantly activated and inhibited reactions are displayed in the plot captions.
+Total reaction numbers in the investigated metabolic subsystem are indicated in the X axes._
+
 \newpage

### 2024-01-01 report/markdown/manuscript_figures_tables.Rmd
Report update
Main figures as discussed with the study team, abbreviated Supplementary Material

--- 2023-07-27 report/markdown/manuscript_figures_tables.Rmd
@@
 ---
-title: "Prognostic and biologic relevance of collagen biosynthesis pathway in prostate cancer"
-subtitle: "Bioinformatic methods, figures and tables"
+title: "Prognostic and biological relevance of collagen biosynthesis pathway in prostate cancer"
+subtitle: "Methods, figures and tables for the transcriptome part"
 date: "`r format(Sys.time(), '%Y-%m-%d')`"
@@
 
-# Tables
+# Methods
 
-```{r tab-cohorts, tab.cap = 'Characteristic of the analyzed cohorts.
-Numeric variables are presented as medians with interquartile ranges (IQR) and ranges.
-Qualitative variables are presented as percentages of categories within the complete observation set.'}
+## Magnetic resonance imaging of the prostate
 
-flextable::flextable(paper_tbl$cohorts) %>% 
-  set_widths(c(3, rep(3.5, 5))) %>% 
-  merge_v(1) %>% 
-  footnote(1, 1, 
-           value = as_paragraph('PSA: prostate-specific antigen.'), 
-           part = 'header', 
-           ref_symbols = 'a') %>%
-  theme_vanilla
+_Dear co-authors, it's your part as experts:)_
 
-```
+ADC was compared between patient-matched measurements of benign prostate and malignancy by paired T test with Cohen's d effect size statistic.
 
-\newpage
+## Transcriptome analyses
 
-```{r tab-genes, tab.cap = 'Collagen genes of interest and their classification.'}
+The transcriptome analysis was performed with R version 4.2.3. 
+Its details are presented in __Supplementary Methods__.
 
-flextable::flextable(paper_tbl$genes) %>% 
-  set_widths(rep(4, 3)) %>% 
-  italic(j = 1, part = 'body') %>% 
-  footnote(1, 3, 
-           value = as_paragraph('ECM: extracellular matrix.'), 
-           part = 'header', 
-           ref_symbols = 'a') %>% 
-  theme_vanilla
+## Transcriptome data sets
 
-```
+Six publicly available transcriptome data sets of primary prostate cancer samples, 
+the TCGA prostate cancer cohort [@Liu2018; @Abeshouse2015] 
+(n = `r tcga$expression %>% filter(tissue_type == 'tumor') %>% nrow`, RNA sequencing), 
+GSE54460 [@Long2014] (n = `r gse54460$expression %>% filter(tissue_type == 'tumor') %>% nrow`, RNA sequencing), 
+GSE70768  [@Ross-Adams2015] (n = `r gse70768$expression %>% filter(tissue_type == 'tumor') %>% nrow`, microarray), 
+GSE70769  [@Ross-Adams2015] (n = `r gse70769$expression %>% filter(tissue_type == 'tumor') %>% nrow`, microarray), 
+GSE220095 [@Schimmelpfennig2023] (n = `r gse220095$expression %>% filter(tissue_type == 'tumor') %>% nrow`, RNA sequencing), 
+DKFZ [@Gerhauser2018] (n = `r dkfz$expression %>% filter(tissue_type == 'tumor') %>% nrow`, RNA sequencing) 
+were re-analyzed (__Supplementary Table S1__). 
+Normalized gene expression data and clinical information were fetched from the cBioportal repository (DKFZ, TCGA) or Gene Expression Omnibus (GEO). 
+Selection criteria of the data sets were: availability of Gleason scoring, information of biochemical relapse and relapse-free survival, and availability of expression data for `r nrow(globals$genes_interest)` collagen-related genes covered by proteomic analysis as well (__Supplementary Table S2__).
+Donor-matched benign and malignant tissue samples were available for the 
+GSE70768 (n = `r norm_tumor$n_numbers$gse70768` tissue pairs) and TCGA cohort (n = `r norm_tumor$n_numbers$tcga` tissue pairs). 
+Expression values were $log_2$-transformed prior to analyses. 
+Non-immune cell content in cancer samples was estimated with the _MCP Counter_ and _xCell_ algorithms [@Aran2017; @Becht2016].
+single sample gene set enrichment scores (ssGSEA) for the Reactome pathway gene signatures retrieved from the MSig database version 7.5.1 were computed with the _GSVA_ algorithm [@Hanzelmann2013].
+
+The collagen-related genes were retrieved from the published Collagen Signature [@Kocher2021] and the Reactome collagen formation pathway (R-HSA-1474290) and restricted to genes covered by the proteomic analyses (__Supplementary Table S2__).
+
+## Differential expression in benign and malignant prostate tissue and cancers stratified by Gleason score
+
+Expression of the collagen-related genes of interest between patient-matched benign and malignant specimens was compared by the false discovery rate method (FDR) corrected paired two-tailed T test with Cohen's d effect size statistic.
+Differentially expressed genes were defined by pFDR < 0.05 and d $\geq$ 0.2 criteria indicative of significant, at least weak expression regulation (__Supplementary Table S3__).
+
+Differential gene expression in cancer samples stratified by Gleason scoring (5 - 6, 7 and $\geq$ 8) was assessed by FDR-corrected one-way ANOVA with $\eta^2$ effect size statistic. 
+Genes with pFDR < 0.05 and $\eta^2 \geq$ 0.02 were deemed differentially regulated (__Supplementary Table S4__).
+
+## Transcriptomic Collagen Score and prediction of biochemical relapse-free survival
+
+Biochemical relapse-free survival was modeled in the TCGA training cohort as a function of first- and second-order normalized, $\log_2$-transformed expression levels of the collagen-related genes of interest by Elastic Net multi-parameter Cox regression [@Simon2011; @Zou2005].
+The Transcriptomic Collagen Score was computed as the linear predictor score of the Elastic Net Cox model with non-zero model coefficients presented in __Figure \@ref(fig:fig-coll-score)A__ and __Supplementary Table S15__.
+
+Performance of the Transcriptomic Collagen Score at prediction of biochemical relapse-free survival in the training collective (TCGA) and validation data sets (remaining cohorts) was assessed by univariable Cox regression with Harrell's concordance index [@Harrell1996] and integrated Brier score [@Graf1999] as metrics of goodness of fit and calibration (__Supplementary Table 16__).
+
+## Data and code availability
+
+Publicly available data sets were analyzed. 
+Formatted data sets used for analyses will be made available upon request to the corresponding author. 
+The transcriptome R analysis pipeline is available from GitHub (https://github.com/PiotrTymoszuk/collagen_pca).
 
@@
 
-```{r fig-norm-tumor, fig.width = figur::convert(paper_fig$norm_tumor, to = 'in')$w, fig.height = figur::convert(paper_fig$norm_tumor, to = 'in')$h, fig.cap = 'Differences in tissue diffusion capacity and expression of the collagen pathway genes between the prostate cancer and benign tissue.'}
+```{r fig-norm-tumor, fig.width = figur::convert(paper_fig$norm_tumor, to = 'in')$w, fig.height = figur::convert(paper_fig$norm_tumor, to = 'in')$h, fig.cap = 'Differences in tissue diffusion capacity and expression of collagen-related genes between the prostate cancer and benign tissue.
+Expression of collagen-related genes in cancers stratified by Gleason scores.'}
 
@@
 
-__Figure \@ref(fig:fig-norm-tumor). Differences in tissue diffusion capacity and expression of the collagen pathway genes between the prostate cancer and benign tissue.__
+__Figure \@ref(fig:fig-norm-tumor).
+Differences in tissue diffusion capacity and expression of collagen-related genes between the prostate cancer and benign tissue.
+Expression of collagen-related genes in cancers stratified by Gleason scores.__
 
@@
 Single ADC values are visualized as points, grey lines connect measurements of the same donors.
-The number of measurement pairs, effect size and p value are displayed in the plot caption._
+The number of analyzed benign - tumor pairs, effect size and p value are displayed in the plot caption._
 
-_(B) Differences in $log_2$-transformed expression levels of `r nrow(globals$genes_interest)` genes related to collagen metabolism between donor-matched tumor and benign prostate tissue were investigated by paired T test with Cohen's d effect size statistic in three published cohorts.
-P values were corrected for multiple testing with the false discovery method.
-Results for gene significantly regulated in at least two cohorts are presented.
-Lines represent means, 2 $\times$ SEM (standard error of the mean) intervals are visualized as tinted ribbons.
-Gene symbols, effect sizes and p values are shown in the Y axes.
-Significant effects are highlighted in bold.
-Numbers of tissue pairs are displayed in the plot captions._
+_(B) Differences in $log_2$-transformed expression of `r nrow(globals$genes_interest)` collagen-related genes between donor-matched tumor and benign prostate tissue samples were investigated by paired T test with Cohen's d effect size statistic in the GSE70768 (n = `r norm_tumor$n_numbers$gse70768`) and TCGA (n = `r norm_tumor$n_numbers$tcga` tissue pairs) cohorts.
+P values were corrected for multiple testing with the false discovery rate method.
+Mean normalized $log_2$-transformed expression levels in benign and cancer tissue for `r length(unique(unlist(norm_tumor$common)))` genes significantly regulated in both cohorts with at least weak effect size (pFDR < 0.05, d $\geq$ 0.2) are presented in a heat map._
+
+_(C) Differences in $log_2$-transformed expression levels of `r nrow(globals$genes_interest)` collagen-related genes between cancer samples stratified by Gleason scores were assessed by one-way ANOVA with $\eta^2$ effect size statistic in the GSE54460 (`r format_strata_n(gs_uni$n_numbers$gse54460)`), GSE70768 (`r format_strata_n(gs_uni$n_numbers$gse70768)`), GSE70769 (`r format_strata_n(gs_uni$n_numbers$gse70769)`), GSE220095 (`r format_strata_n(gs_uni$n_numbers$gse220095)`), DKFZ (`r format_strata_n(gs_uni$n_numbers$dkfz)`), and TCGA cohort (`r format_strata_n(gs_uni$n_numbers$tcga)`).
+Mean normalized $log_2$-transformed expression levels of genes significantly regulated between the Gleason score strata with at least weak effect size (pFDR < 0.05, $\eta^2 \geq$ 0.02) are presented in a heat map._
 
@@
 
-```{r fig-coll-clusters, fig.width = figur::convert(paper_fig$coll_clusters, to = 'in')$w, fig.height = figur::convert(paper_fig$coll_clusters, to = 'in')$h, fig.cap = 'Collagen clusters of prostate carcinoma and their clinical characteristic.'}
+```{r fig-tme, fig.width = figur::convert(paper_fig$tme, to = 'in')$w, fig.height = figur::convert(paper_fig$tme, to = 'in')$h, fig.cap = 'Expression of collagen-related genes in components of the tumor microenvironment.'}
 
-paper_fig$coll_clusters$plot
+paper_fig$tme$plot
 
@@
 
-__Figure \@ref(fig:fig-coll-clusters). Collagen clusters of prostate carcinoma and their clinical characteristic.__
+__Figure \@ref(fig:fig-tme). Expression of collagen-related genes in components of the tumor microenvironment.__
 
-_Prostate cancers in the training TCGA cohort were clustered in respect to normalized $log_2$-transformed expression levels of the collagen pathway genes of interest by the PAM algorithm with Manhattan distance between the observations.
-Three clusters, collagen^low^ (low), collagen^intermediate^ (int) and collagen^high^ (hi) were defined.
-Cancer samples from the test cohorts (`r globals$study_labels[globals$study_labels != 'TCGA'] %>% paste(collapse = ', ')`) were assigned to the collagen clusters with an inverse distance weighted k-nearest neighbor classifier._
+_(A) Cancer samples in six published cohorts were classified as collagen low and collagen high tumors in respect to `r nrow(globals$genes_interest)` collagen-related transcripts by semi-supervised PAM clustering (GSE54460: `r format_strata_n(ana_globals$n_numbers$gse54460)`, GSE70768: `r format_strata_n(ana_globals$n_numbers$gse70768)`, GSE70769: `r format_strata_n(ana_globals$n_numbers$gse70769)`, GSE220095: `r format_strata_n(ana_globals$n_numbers$gse220095)`, DKFZ: `r format_strata_n(ana_globals$n_numbers$dkfz)`, TCGA: `r format_strata_n(ana_globals$n_numbers$tcga)`).
+Counts of cancer-associated fibroblasts and endothelial cells were predicted by the MCP Counter algorithm and compared between the collagen clusters by Mann-Whitney test with r effect size statistic.
+Median normalized cell counts with interquartile ranges are presented as boxes, whiskers span over 150% of the interquartile ranges.
+Values for single cancer samples are depicted as points.
+All presented effects were significant (pFDR < 0.05) with at least weak (endothelial cells, r > 0.21) or moderate effect size (fibroblasts, r > 0.55)._
 
-_(A) Mean normalized $log_2$-transformed expression levels of the collagen pathway genes in the collagen clusters and cohorts presented in a heat map. ECM: extracellular matrix._
-
-_(B) Distribution of the collagen clusters expressed as percentage of the cohort cancer samples visualized in a stack plot. Total numbers of cancer samples are displayed in the X axis. _
-
-_(C, D) Distribution of Gleason scores (GS, C) and pathological tumor stages (pT, D) in the collagen clusters.
-Statistical significance was determined by $\chi^2$ test with Cramer's V effect size statistic.
-P values were corrected for multiple testing with the false discovery rate method.
-Gleason score and tumor stage distributions are shown in stack plots.
-Effect sizes and p values are displayed in the plot captions.
-Numbers of observations in the clusters are indicated in the X axes._
+_(B) Placeholder for the single cell RNA seq data._
 
@@
 
-```{r fig-cluster-biology, fig.width = figur::convert(paper_fig$cluster_biology, to = 'in')$w, fig.height = figur::convert(paper_fig$cluster_biology, to = 'in')$h, fig.cap = 'Infiltration, biological processes, signaling and metabolic pathways differentially regulated in the collagen clusters.'}
+```{r fig-coll-score, fig.width = figur::convert(paper_fig$coll_score, to = 'in')$w, fig.height = figur::convert(paper_fig$coll_score, to = 'in')$h, fig.cap = 'Transcriptional Collagen Score and prediction of biochemical relapse-free survival.'}
 
-paper_fig$cluster_biology$plot
+paper_fig$coll_score$plot
 
@@
 
-__Figure \@ref(fig:fig-cluster-biology). Infiltration, biological processes, signaling and metabolic pathways differentially regulated in the collagen clusters.__
+__Figure \@ref(fig:fig-coll-score). Transcriptional Collagen Score and prediction of biochemical relapse-free survival.__
 
-_(A) Counts of cancer-associated fibroblasts (CAF) and endothelial cells (EC) estimated by the MCP counter algorithm were found to differ significantly between the collagen clusters in all investigated collectives.
-Statistical significance was assessed by Kruskal-Wallis test with $\eta^2$ effect size statistic.
-P values were corrected for multiple comparison with the false discovery rate method.
-Counts of fibroblasts and endothelial cells in the collagen clusters in the TCGA cohort are shown in violin plots.
-Red diamonds with whiskers represented medians with interquartile ranges.
-Single tumor samples are visualized as points.
-Effect sizes and p values are displayed in the plot captions.
-Numbers of observations in the clusters are indicated in the X axes.
-Full analysis results are presented in Supplementary Table S4._
+_Transcriptional Collagen Score was established in the TCGA training cohort by multi-parameter Elastic Net Cox modeling of biochemical relapse-free survival as a function of first- and second-order normalized $log_2$-transformed expression levels of `r nrow(globals$genes_interest)` collagen-related genes._
 
-_(B) Gene set variation analysis (GSVA) with Reactome pathway gene signatures.
-Statistical significance for differences in signature single sample gene set enrichment analysis scores (ssGSEA) between the collagen clusters was determined by one-way ANOVA with $\eta^2$ effect size statistic.
-P values were corrected for multiple testing with the false discovery rate method.
-Mean ssGSEA scores in the clusters and cohorts for signatures significantly regulated in at least four cohorts are presented in a heat map.
-Full analysis results are presented in Supplementary Table S6.
-ECM: extracellular matrix, GF: growth factor._
+_(A) Non-zero Elastic Net Cox model coefficients presented as hazard ratios (HR). Point color codes for marker status (blue: favorable, red: unfavorable), point size represents absolute value of HR._
 
-_(C) Modulation of KEGG-listed signaling pathways in the collagen^high^ (hi) or collagen^intermediate^ (hi) as compared with collagen^low^ tumors was predicted by the SPIA algorithm based on differential gene expression estimates.
-Predicted magnitude of pathway modulation (tA parameter) for pathways significantly modulated in at least four cohorts was visualized in bubble plots (tA < 0: inhibition, blue; tA > 0, activation, red).
-Points are labeled with their tA values.
-Significant effects are highlighted in bold.
-Full analysis results are presented in Supplementary Table S8.
-ECM: extracellular matrix, ACT: actin, SCLC: small cell lung carcinoma, S.
-aureus: Staphylococcus aureus, SLE: systemic lupus erythematosus._
+_(B) Performance of the Elastic Net Cox model at prediction of biochemical relapse-free survival in the training TCGA cohort and test collectives assessed by concordance index (C-index, high values indicate high concordance between the predicted and observed survival) and integrated Brier score (IBS, low values indicate good model fit and proper calibration).
+Values of C-index and IBS expected for random survival prediction are depicted as dashed lines._
 
-_(D) Modulation of Recon2 model metabolic reactions in the collagen^high^ (hi) or collagen^intermediate^ (hi) as compared with collagen^low^ tumors was predicted by the BiGGR and biggrExtra algorithms based on differential gene expression estimates.
-Enrichment of significantly activated and significantly inhibited reactions in the collagen^high^ or collagen^intermediate^ clusters within the Recon model metabolic subsystems was investigated by Fisher's exact test.
-Enrichment p values were corrected for multiple testing with the false discovery rate method.
-Enrichment odds ratios for activated (red) and inhibited reaction (blue) for significantly enriched metabolic subsystems shared by at least four cohorts are presented in bubble plots.
-Points are labeled with their odds ration values.
-Full analysis results are presented in Supplementary Table S10.
-Val: valine, Leu: leucine, Ile: isoleucine, FAOX: fatty acid oxidation._
+_(C) Fractions of biochemical relapse-free patients in low, intermediate and high tertiles of Transcriptional Collagen Score in the training TCGA cohort, and the GSE70768 and DKFZ validation collectives presented in Kaplan-Meier plots.
+Differences in survival between the tertiles were investigated by Peto-Peto test.
+Total numbers of patients and numbers of biochemical relapses are displayed in the plot captions.
+P values are shown in the plots._
 

### 2024-01-01 report/markdown/manuscript_supplement.Rmd
Report update
Main figures as discussed with the study team, abbreviated Supplementary Material

--- 2023-07-27 report/markdown/manuscript_supplement.Rmd
@@
 ---
-title: "Prognostic and biologic relevance of collagen biosynthesis pathway in prostate cancer"
+title: "Prognostic and biological relevance of collagen biosynthesis pathway in prostate cancer"
 subtitle: "Supplementary material, transcriptome part"
@@
 
+# Supplementary Methods
+
+## Software
+
+The analysis was done with R version 4.2.3 (R Foundation).
+
+Tabular data were handled with the packages _tidyverse_ [@Wickham2019], _rlang_ [@Henry2022] and [_trafo_](https://github.com/PiotrTymoszuk/trafo). 
+Text data were handled with _stringi_ [@Gagolewski2021].
+
+Import of the TCGA and DKFZ data sets from the cBioportal repository was accomplished with in-house-developed R scripts. 
+Transcriptome datasets from the Gene Expression Omnibus were fetched with the _GEOquery_ package [@Sean2007]. 
+Gene and probe annotation was accomplished with the _AnnotationDbi_ [@Pages2022] and _org.Hs.eg.db_ packages [@Carlson2022]. 
+For prediction of non-malignant cell counts and fractions in cancer samples, the R implementation of the _MCP Counter_ and _xCell_ algorithms provided by the _immunedeconv_ package was used [@Sturm2020; @Aran2017; @Becht2016].
+Single sample gene set enrichment analysis scores (ssGSEA) were computed with the _GSVA_ algorithm [@Hanzelmann2013] implemented by the package [_gseaTools_](https://github.com/PiotrTymoszuk/gseaTools).
+
+Semi-supervised clustering was done with the package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) employing algorithms from the _philentropy_, _cluster_, _factoextra_ and _umap_ packages [@Kassambara2020; @Schubert2019; @Drost2018; @Konopka2022].
+
+For statistical hypothesis testing, effect size calculation, gene set variation analysis (GSVA) and differential gene expression analysis, the packages _rstatix_ [@Kassambara2021], [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) and [_microViz_](https://github.com/PiotrTymoszuk/microViz) were employed.
+Biological process gene ontology (GO) enrichment analysis was performed with the _goana_ algorithm implemented by the _limma_ [@Young2010] and [_microViz_](https://github.com/PiotrTymoszuk/microViz) packages.
+Clustering of GO terms by their semantic similarity was accomplished with _GOSemSim_ [@Yu2010], [_microViz_](https://github.com/PiotrTymoszuk/microViz) and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools).
+Differential modulation of transcriptional regulons and signaling pathways was analyzed with the _collecTRI_ [@Muller-Dott2023] and _PROGENy_ [@Schubert2018] databases by linear modeling tools of the _decoupleR_ package [@Badia-I-Mompel2022].
+Metabolic reaction activity modeling and metabolic subsystem enrichment analysis were performed with the packages _BiGGR_ [@Gavai2015; @King2016] and [_biggrExtra_](https://github.com/PiotrTymoszuk/biggrExtra).
+
+For multi-parameter Elastic Net Cox modeling, the packages _glmnet_ [@Friedman2010], _survival_ [@Therneau2000], _surviminer_ [@Kassambara2016] and [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions) were used.
+
+For visualization of the results, the packages 
+_ggplot2_ [@Wickham2019] (scatter, stack, bubble and bar plots, heat maps), 
+[_ExDA_](https://github.com/PiotrTymoszuk/ExDA) (violin, stack and ribbon plots), _survminer_ and [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions) (Kaplan-Meier plots), and 
+[_microViz_](https://github.com/PiotrTymoszuk/microViz) (heat maps) were used. 
+Figures and tables were created with the packages _cowplot_ [@Wilke2019] and _flextable_ [@Gohel2022].
+
+## Data sources and data import
+
+Transcriptome cohorts for analysis in the current report were selected from studies deposited at cBioportal and Gene Expression Omnibus (GEO) with the following criteria: availability of information on Gleason scoring, biochemical relapse and biochemical relapse-free survival, and expression data for `r nrow(globals$genes_interest)` collagen-related genes investigated also in the proteomic part of the project (__Supplementary Table S\@ref(tab:tab-cohorts)__ and __S\@ref(tab:tab-genes)__).
+
+The TCGA prostate cancer data set [@Liu2018; @Abeshouse2015] and DKFZ data set [@Gerhauser2018] consisted of normalized RNA sequencing data for `r tcga$expression %>% filter(tissue_type == 'tumor') %>% nrow` and `r dkfz$expression %>% filter(tissue_type == 'tumor') %>% nrow` cancer samples, respectively, with accompanying clinical information and were obtained from the cBioportal repository with in-house-developed R scripts.
+The TCGA cohort included also 
+`r norm_tumor$n_numbers$tcga` donor-matched samples of the benign prostate and prostate cancer tissue.  
+The GSE54460 [@Long2014] (RNA sequencing, n = `r gse54460$expression %>% filter(tissue_type == 'tumor') %>% nrow` cancer samples), 
+GSE70768 [@Ross-Adams2015] (Illumina HumanHT-12 V4.0 expression beadchip microarray, n = `r gse70768$expression %>% filter(tissue_type == 'tumor') %>% nrow`), 
+GSE70769 [@Ross-Adams2015] (Illumina HumanHT-12 V4.0 expression beadchip microarray, n = `r gse70769$expression %>% filter(tissue_type == 'tumor') %>% nrow`), and 
+GSE220095 [@Schimmelpfennig2023] (RNA sequencing, n = `r gse220095$expression %>% filter(tissue_type == 'tumor') %>% nrow`) data sets included normalized whole-transcriptome and basic clinical information and were fetched from GEO with the `getGEO()` function from the _GEOquery_ package.
+The GSE70768 data set [@Ross-Adams2015] included also `r norm_tumor$n_numbers$gse70768` donor-matched normal prostate and cancer tissue specimens. 
+Demographic and clinical characteristic of the investigated transcriptome data sets is provided in  __Supplementary Table S\@ref(tab:tab-cohorts)__.
+
+Gene expression levels were transformed with the $log_2(expression)$ and $log_2(count + 1)$ function for the microarray and RNA sequencing data sets prior to further analyses. 
+The _MCP Counter_ and _xCell_ estimates of non-malignant cell content in cancer samples were computed with the `deconvolute()` function from the _immunedeconv_ package. 
+Gene signatures of the [Reactome pathways](https://reactome.org/) were obtained from the [MSig database](https://www.gsea-msigdb.org/gsea/msigdb/index.jsp), version 7.5.1. 
+ssGSEA scores were computed with the `calculate()` function from the [_gseaTools_](https://github.com/PiotrTymoszuk/gseaTools) package.
+
+The investigated collagen-related genes included the published Collagen Signature [@Kocher2021] and the Reactome Collagen formation pathway genes (R-HSA-1474290) and were constrained to features covered by proteomic analyses of the current report (__Supplementary Table S\@ref(tab:tab-genes)__).
+
+## Statistical hypothesis testing, effect size and statistical significance
+
+Differences in numeric variables between two groups were investigated by paired and unpaired two-tailed T tests with Cohen's d effect size statistic or Mann-Whitney tests with r effect size statistic.
+Differences in numeric variables between three or more groups were assessed by one-way ANOVA with $\eta^2$ effect size statistic. 
+Differences in distribution of categorical variables were assessed by $\chi^2$ test with Cramer's V effect size statistic. 
+Odds ratio served as an effect size statistic in enrichment analyses. 
+Effect sizes were interpreted as follows [@Cohen2013]:
+
+* Cohen's d, weak: 0.2 - 0.5, moderate: 0.5 - 0.8, large: $\geq$ 0.8
+
+* r, weak: 0.1 - 0.3, moderate: 0.3 - 0.5, large: $\geq$ 0.5
+
+* $\eta^2$: weak: 0.02 - 0.13, moderate: 0.13 - 0.26, large: $\geq$ 0.26
+
+* Cramer's V, weak: 0.1 - 0.3, moderate: 0.3 - 0.5, large: $\geq$ 0.5
+
+* OR, weak: 1.44 - 2.48, moderate: 2.48 - 4.27, large: $\geq$ 4.27
+
+P values were corrected for multiple testing with the false discovery rate method (FDR) [@Benjamini1995] within each analysis step and cohort. 
+Effects were considered statistically significant for FDR-corrected p values < 0.05.
+
+## Comparison of gene expression between the benign and cancer tissue, and cancer samples stratified by Gleason scores
+
+Differences in $log_2$-transformed expression levels of the collagen pathway genes between donor-matched cancer and benign prostate specimens were investigated with paired two-tailed T test with Cohen's effect size statistic in the GSE70768 and TCGA cohorts (__Figure 1B__, __Supplementary Figure S\@ref(fig:fig-norm-tumor)__ and __Supplementary Table S\@ref(tab:tab-norm-tumor)__).
+Differences in gene expression between cancer samples stratified by Gleason score (5 - 6, 7, $\geq$ 8) were explored by one-way ANOVA with $\eta^2$ effect size statistic (__Figure 1C__, __Supplementary Table S\@ref(tab:tab-gleason)__).
+Differentially regulated genes were defined by pFDR < 0.05 and at least weak effect size of the difference (d $\geq$ 0.2 or $\eta^2 \geq$ 0.02). 
+The analyses were done with the function `compare_variables()` from the [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) package.
+
+## Collagen clusters of prostate cancer samples
+
+Prostate cancer samples in the TCGA cohort were clustered by normalized $log_2$-transformed expression levels of the collagen-related genes of interest with the PAM (partition around medoids) algorithm with cosine distance between the samples (function `kcluster()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)).
+The choice of the clustering algorithm was motivated by the following criteria (__Supplementary Figure S\@ref(fig:fig-clust-dev)A__):
+
+* cluster separation measured by mean silhouette statistic [@Rousseeuw1987]
+
+* fraction of potentially misclassified observations (observations with negative silhouette widths) [@Rousseeuw1987]
+
+* explanatory performance gauged by fraction of explained clustering variance (ratio of the total between-cluster sum of squares to the total sum of squares)
+
+* preservation of the nearest neighborhood (mean fraction of the five nearest neighbors placed in the same cluster) [@Venna2001]
+
+* cluster re-assignment accuracy in 5-fold cross-validation [@Lange2004; @Leng2014]. 
+Computation of those numeric statistics of clustering quality in a cross-validation setting was done with the methods `summary()` and `cv()` provided by the [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) package.
+The cluster number choice was based on the bend of the peak of the average silhouette statistic [@Rousseeuw1987] (method `plot()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools), __Supplementary Figure S\@ref(fig:fig-clust-dev)B__).
+By this means, two clusters of cancer samples were identified: 'collagen high' and 'collagen low' cancers.
+
+Assignment of cancers samples of the GSE54460, GSE70768, GSE70769, GSE220095, and DKFZ cohorts to the collagen clusters based on normalized $log_2$-transformed expression levels of the collagen-related genes was done with an inverse distance weighted k-nearest neighbor classifier (method `predict()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) [@Leng2014].
+The quality of the cluster assignment was assessed by comparing the collagen cluster distribution and numeric statistics of clustering quality described above the training TCGA data set and the test cohorts (__Supplementary Figure S\@ref(fig:fig-clust-dev)CD__).
+In addition, separation of the clusters was assessed by a visual inspection of the UMAP layout plots (method `plot()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) and heat maps of the mean levels of the clustering factors in the collagen clusters (__Supplementary Figure S\@ref(fig:fig-clusters)__).
+
+Differences in the $log_2$-transformed collagen pathway gene expression between the collagen clusters were investigated by two-tailed T test with Cohen's d effect size statistic (__Supplementary Table S\@ref(tab:tab-cluster)__).
+
+## Clinical characteristic of the collagen clusters
+
+Differences in numeric clinical variables (age and PSA) between the collagen clusters were assessed by Mann-Whitney test with r effect size statistic (`test_two_groups()`, [_microViz_](https://github.com/PiotrTymoszuk/microViz)).
+Differences in qualitative clinical variables between the collagen clusters were investigated by $\chi^2$ test with Cramer's V effect size statistic (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)).
+The demographic and clinical characteristic of the collagen clusters is presented in __Supplementary Figure S\@ref(fig:fig-clust-tumor)__ and __Supplementary Table S\@ref(tab:tab-clinic)__.
+
+## Non-malignant cell infiltration in the collagen clusters
+
+Non-malignant cell counts and non-malignant cell fractions in the cancer samples were predicted by the _MCP Counter_ [@Becht2016] and _xCell_ [@Aran2017] algorithms, respectively.
+Differences in the predicted infiltration levels between the collagen clusters were investigated by Mann-Whitney test with r effect size statistic (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)).
+Populations differing between the clusters with pFDR < 0.05 and at least weak effect size (r $\geq$ 0.1) were considered biologically relevant. 
+The analysis results are shown in __Supplementary Figure S\@ref(fig:fig-intiltration)__ and __Supplementary Table S\@ref(tab:tab-infiltration)__.
+
+## Gene set variance analysis of Reactome pathway gene signatures
+
+ssGSEA scores [@Hanzelmann2013] of the Reactome pathway gene signatures were compared between the collagen clusters by two-tailed T test with Cohen's d effect size statistic. 
+Signatures found to be significantly regulated (pFDR < 0.05) with at least weak effect size (d $\geq$ 0.2) in at least five cohorts were further investigated. 
+To get additional functional insight and for purposes of visualization, these common regulated signatures were classified by their co-expression patterns in the TCGA data set by unsupervised KMEANS clustering (function `kcluster()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)).
+The analysis results are presented in __Supplementary Figure S\@ref(fig:fig-biology)__ and __Supplementary Table S\@ref(tab:tab-gsva)__.
+
+## Differential gene expression and GO enrichment analysis
+
+Differences in whole-transcriptome gene expression between the collagen clusters were investigated by FDR-corrected two-tailed T test with Cohen's d effect size statistic (function `test_two_groups()`, package [_microViz_](https://github.com/PiotrTymoszuk/microViz)).
+Genes significantly differentially expressed between the clusters were identified by the significance cutoff (pFDR < 0.05) and at least weak effect size of the expression differences (d $\geq$ 0.2). 
+Genes found to be differentially regulated between the clusters are listed in __Supplementary Table S\@ref(tab:tab-dge)__.
+
+Biological process GO enrichment within genes upregulated in each of the collagen high and collagen low clusters was analyzed with the _limma_ package _goana_ algorithm [@Young2010], re-implemented by the development package [_microViz_](https://github.com/PiotrTymoszuk/microViz) (function `GOana()`).
+Enrichment p values were adjusted for multiple testing with the FDR method. 
+Odds ratio (OR) calculated for a GO term with the following formula served as a measure of effect size:
+
+$$OR_{GO} = \frac{N_{GO|DGE} \times N_{total}}{N_{GO|total} \times N_{DGE}}$$
+where $N_{GO|DGE}$ stands for the number of differentially regulated genes assigned to the GO term, 
+$N_{total}$ is the total number of investigated genes, 
+$N_{GO|total}$ is the number of investigated genes assigned to the GO term, and 
+$N_{DGE}$ stands for the number of differentially regulated genes. 
+GO terms found to be significantly enriched with at least weak effect size (pFDR < 0.05, OR $\geq$ 1.44) in at least five cohorts were further investigated. 
+For additional functional insight, such common enriched GO terms were subjected to multi-dimensional scaling and unsupervised KMEANS clustering with pairwise Wang distance as a measure of semantic similarity (functions `go_sem()` and `kcluster()`, packages _GOSemSim_, [_microViz_](https://github.com/PiotrTymoszuk/microViz) and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) [@Yu2010].
+Results of the GO enrichment analysis are shown in __Supplementary Figure S\@ref(fig:fig-biology)__ and __Supplementary Table S\@ref(tab:tab-go-enrichment)__.
+
+## Transcriptional regulons and signaling pathway activity in the collagen clusters
+
+Modulation of transcriptional regulons, i.e.
+sets of genes whose expression is controlled by a common transcriptional factor, and activity of selected signaling pathways in collagen high cancers as compared with collagen low tumors were explored by linear modeling with the _collecTRI_ [@Muller-Dott2023] and _PROGENy_ [@Schubert2018] databases, respectively.
+Liner models were constructed for whole-transcriptome effect sizes of gene regulation (Cohen's d, see: __Differential gene expression and GO enrichment analysis__) with the `run_ulm()` and `run_mlm()` tools from the _decoupleR_ package [@Badia-I-Mompel2022].
+In each case, linear model score (LM score) served as a metric of activity magnitude. 
+P values (LM score $\neq$ 0) were corrected for multiple testing with the FDR method. 
+Regulons and signaling pathways found to be significantly activated or inhibited in at least five cohorts were further discussed. 
+The analysis results are presented in __Supplementary Figure S\@ref(fig:fig-signaling)__,and __Supplementary Tables S\@ref(tab:tab-regulons)__ and __S\@ref(tab:tab-signaling)__.
+
+## Activity of metabolic reactions in the collagen clusters
+
+Rules of assignment of genes to biochemical reactions were retrieved from the Recon2 human metabolism model available via the BiGG database [@King2016] and the R package _BiGGR_ [@Gavai2015]. 
+Estimates and standard errors of $log_2$ fold-regulation of expression for all available genes between the collagen clusters were calculated by two-tailed T test as described in __Differential gene expression and GO enrichment analysis__.
+Estimates of biochemical pathway fold-regulation were computed by evaluation of the gene assignment rules in the Recon2 model. 
+The 'gene~A~ OR gene~B~' operator was interpreted as arithmetic mean of expression regulation estimates for the genes A and B. 
+The 'gene~A~ AND gene~B~' operator was interpreted as minimum of expression regulation estimates for the gene A and gene B [@Gavai2015]. 
+Standard deviation, 95% confidence intervals and p values for the predicted reaction regulation estimates were obtained by a Monte Carlo simulation with n = 3000 draws from normal distribution of gene expression regulation estimates (mean: gene expression regulation estimate, standard deviation: standard error of the gene expression regulation estimate) [@Gavai2015; @Pichler2023].
+P values were corrected for multiple testing with the FDR method. 
+The analysis was done with the packages _BiGGR_ [@Gavai2015] and [_biggrExtra_](https://github.com/PiotrTymoszuk/biggrExtra) [@Pichler2023] (function `build_geneSBML()`). 
+Biochemical reactions predicted to be significantly activated or inhibited in the collagen high cluster as compared with the collagen low cluster are listed in __Supplementary Table S\@ref(tab:tab-reactions)__.
+
+Enrichment of significantly activated or inhibited biochemical reactions within the Recon model metabolic subsystems [@King2016] was investigated by comparing the frequency of significantly activated or inhibited reactions in the given subsystem with 10000 random samples from the total reaction pool (function `suba()`, package [_biggrExtra_](https://github.com/PiotrTymoszuk/biggrExtra)).
+The magnitude of enrichment was assessed with OR defined for the $i$-th metabolic subsystem as follows:
+
+$$OR_i = \frac{n_{regulated, i} \times n_{total}}{n_i \times n_{regulated}}$$
+
+where $n_{regulated, i}$ denotes the number of significantly activated or inhibited biochemical reactions within the $i$-th metabolic subsystem, 
+$n_{total}$ denotes the total number of investigated reactions, 
+$n_i$ denotes the number of metabolic reactions within the $i$-th metabolic subsystem, and 
+$n_{regulated}$ is the total number of significantly activated or inhibited biochemical reactions. 
+Common significantly activated or inhibited metabolic subsystems were identified as metabolic subsystems significantly activated or inhibited in at least four cohorts, with at least weak effect size defined by OR $\geq$ 1.44 (__Supplementary Figure S\@ref(fig:fig-metabolism)__, __Supplementary Table S\@ref(tab:tab-subsystems)__).
+
+## Survival analysis and transcriptional collagen score
+
+Biochemical relapse-free survival was modeled in the TCGA training cohort as a function of normalized $log_2$-transformed expression of the collagen-related genes of interest by multi-parameter Elastic Net Cox modeling [@Simon2011; @Zou2005].
+Of note, both first and second order terms were included in the Elastic Net model to account for possible non-linear association of gene expression with survival probability. 
+The optimal values of the shrinkage parameter $\lambda$ was found by tuning in 200 repeats 10-fold cross-validation and corresponded to the model with the lowest deviance (function `cv.glmnet()`, package _glmnet_) [@Friedman2010].
+The Elastic Net model was trained for the TCGA data set for the tuned $\lambda$ parameter with the `glmnet()` function from the _glmnet_ package. 
+The Transcriptional Collagen Score was defined as the linear predictor score of the Elastic Net model and was computed for the training TCGA cohort and the validation collectives GSE54460, GSE70768, GSE70769, GSE220095 and DKFZ with the `predict()` method.
+Gene expression values used for calculation of the Transcriptional Collagen Score, i.e.
+non-zero coefficients of the Elastic Net model, are presented as hazard ratios (HR) in __Figure 3A__ and __Supplementary Table S\@ref(tab:tab-coll-score)__.
+
+Performance of the Transcriptional Collagen Score at prediction of biochemical relapse-free survival was assessed by uni-variable Cox regression (functions `coxph()` and `as_coxex()`, packages _survival_ and [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions)) [@Therneau2000].
+The proportional hazard assumption of the Cox models was checked with the `summary(type = 'assumptions')` method, which implements the genuine `cox.zph()` algorithm [@Grambsch1994]. 
+Harrell's concordance index [@Harrell1996] and integrated Brier score [@Graf1999] were employed to investigate the overall concordance between the predicted and observed survival and to assess overall calibration (method `summary(type = 'fit')`, package [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions), __Figure 3B__ and __Supplementary Table S\@ref(tab:tab-coll-survival)__).
+Statistical significance of differences in biochemical relapse-free survival between patients stratified by tertiles of the Transcriptional Collagen Score was determined by Peto-Peto test (functions `surv_fit()` and `surv_pvalue()`, package _survminer_, __Figure 3C__).
+
+## Data and code availability
+
+Publicly available data sets were analyzed. 
+Formatted data sets used for analyses will be made available upon request to the corresponding author. 
+The transcriptome R analysis pipeline is available from GitHub (https://github.com/PiotrTymoszuk/collagen_pca).
+
+\newpage
+
 # Supplementary Tables
+
+```{r tab-cohorts, tab.cap = 'Characteristic of the analyzed cohorts.
+Numeric variables are presented as medians with interquartile ranges (IQR) and ranges.
+Qualitative variables are presented as percentages of categories within the complete observation set.'}
+
+flextable::flextable(suppl_paper_tbl$cohorts) %>% 
+  set_widths(c(3, rep(2.2, 6))) %>% 
+  merge_v(1) %>% 
+  footnote(1, 1, 
+           value = as_paragraph('PSA: prostate-specific antigen.'), 
+           part = 'header', 
+           ref_symbols = 'a') %>%
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-genes, tab.cap = 'Collagen-related genes and their classification.'}
+
+flextable::flextable(suppl_paper_tbl$genes) %>% 
+  set_widths(widths = c(4.5, 3.5, 3)) %>% 
+  italic(j = 2, part = 'body') %>% 
+  merge_v(1) %>% 
+  theme_vanilla
+
+```
+
+\newpage
 
@@
 
-```{r tab-cluster, tab.cap = 'Expression of the cluster-defining collagen pathway genes in the collagen clusters of prostate cancer.
-Statistical significance was assessed by one-way ANOVA with eta-squared effect size statistic.
+\newpage
+
+```{r tab-gleason, tab.cap = 'Expression of the collagen pathway genes in cancer samples stratified by Gleason scores compared by one-way ANOVA with eta-square effect size statistic.
 P values were corrected for multiple testing with the false discovery rate method.
@@
 The table is available as a supplementary Excel file.'}
+
+suppl_paper_tbl$gleason[1, 1] %>% 
+  mutate(Cohort = '') %>% 
+  set_names(' ') %>% 
+  flextable %>% 
+  set_widths(16)
+
+```
+
+\newpage
+
+```{r tab-cluster, tab.cap = "Expression of the cluster-defining collagen pathway genes in the collagen clusters of prostate cancer.
+Statistical significance was assessed by two-tailed T test with Cohen's d effect size statistic.
+P values were corrected for multiple testing with the false discovery rate method.
+log2-transformed expression values are presented as medians with interquartile ranges (IQR) and ranges.
+The table is available as a supplementary Excel file."}
 
@@
 flextable::flextable(suppl_paper_tbl$clinic) %>% 
-  set_widths(c(2.2, 3, rep(3.5, 3), 2.8, 2.8)) %>% 
-  merge_v(1:2) %>% 
+  set_widths(c(2.8, 3.5, 4, 4, 2.8, 2.8)) %>% 
+  merge_v(1) %>% 
   footnote(1, 2, 
@@
            ref_symbols = 'a') %>% 
-  footnote(1, 6:7, 
-           value = as_paragraph("Categorical variables: \u03C7\u00B2 test with Cramer's V effec size statistic.
-Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic.
-P values corrected for multiple testing with the false discovery rate method."),
+  footnote(1, 5:6, 
+           value = as_paragraph('Qualitative variables: \u03C7\u00B2 test with Cramer V effect size statistic.
+Numeric variables: Mann-Whitney test with r effect size statistic.
+P values corrected for multiple testing with the false discovery rate.'),
            part = 'header', 
@@
 
-```{r tab-mcp, tab.cap = 'Non-malignant cell numbers predicted for the collagen Non-maignant cell fractions predicted for the collagen Statistical significance was assessed by Kruskal-Wallis test with eta-squared effect size statistic.
+```{r tab-infiltration, tab.cap = 'Non-malignant cell numbers predicted for the collagen clusters by the MCP Counter and xCell algorithms.
+Statistical significance was assessed by Mann-Whitney test with r effect size statistic.
 P values were corrected for multiple testing with the false discovery method.
@@
 
-suppl_paper_tbl$mcp[1, 1] %>% 
+suppl_paper_tbl$infiltration[1, 1] %>% 
+  mutate(Algorithm = '') %>% 
+  set_names(' ') %>% 
+  flextable %>% 
+  set_widths(16)
+
+```
+
+\newpage
+
+```{r tab-gsva, tab.cap = "Gene set variation analysis with the Reactome pathway gene signatures.
+Differences in ssGSEA scores between collagen high and collagen low cancers were investigated by two-tailed T test with Cohen's d effect size statistic.
+Results for signatures significantly regulated with at least weak effect size (d at least 0.2) in at least five cohorts are presented.
+P values were corrected for multiple testing with the false discovery rate method (FDR).
+The table is available as a supplementary Excel file."}
+
+suppl_paper_tbl$reactome[1, 1] %>% 
   mutate(Cohort = '') %>% 
@@
 
-```{r tab-xcell, tab.cap = 'clusters by the MCP counter algorithm.
-clusters by the xCell algorithm.
-Statistical significance was assessed by Kruskal-Wallis test with eta-squared effect size statistic.
-P values were corrected for multiple testing with the false discovery method.
-The table is available as a supplementary Excel file.'}
+\newpage
 
-suppl_paper_tbl$xcell[1, 1] %>% 
-  mutate(Cohort = '') %>% 
-  set_names(' ') %>% 
-  flextable %>% 
-  set_widths(16)
-
-```
-
-```{r tab-gsva, tab.cap = 'Gene set variation analysis with the Reactome pathway gene signatures.
-Differences between the collagen intermediate or high clusters versus collagen low cancers were investigated by one-way ANOVA with eta-squared effect size statistic and linear modeling.
-Results for signatures significantly regulated with moderate-to-large effect size (eta-squared at leat 0.06) in at least four cohorts are presented.
-P values were corrected for multiple testing with the false discovery rate method (FDR).
-The table is available as a supplementary Excel file.'}
-
-suppl_paper_tbl$gsva[1, 1] %>% 
-  mutate(Cohort = '') %>% 
-  set_names(' ') %>% 
-  flextable %>% 
-  set_widths(16)
-
-```
-
-```{r tab-dge, tab.cap = 'Genes differentially expressed in the collagen intermediate or high cluster as compared with collagen low cancers were identified by one-way ANOVA and linear modeling with the 1.25-fold regulation cutoff P values were corrected for multiple testing with the false discovery rate method (FDR).
-The table is available as a supplementary Excel file.'}
+```{r tab-dge, tab.cap = "Genes differentially expressed in the collagen high cluster as compared with collagen low cancers were identified by two-tailed T test with the 1.25-fold regulation cutoff and the Cohen's d effect size statistic of 0.2 P values were corrected for multiple testing with the false discovery rate method (FDR).
+The table is available as a supplementary Excel file."}
 
@@
 
-```{r tab-signaling, tab.cap = 'Signaling pathway activity in the collagen clusters investigated by the SPIA algorithm.
-Resulat for signaling pathways significantly activated or inhibited in at least four cohorts are shown.
+\newpage
+
+```{r tab-go-enrichment, tab.cap = 'Biological process gene ontology (GO) term enrichment within genes differentially regulated in the collagen clusters.
+The enrichment analysis was performed with goana tool, enrichment p values were corrected for multiple testing with the false discovery rate (FDR) method.
+Significant enrichment was defined by pFDR < 0.05 and odds ratio (OR) for enrichment within differentially regulated genes of at least 1.44.
+OR for enrichment within genes upregulated in the collagen high and collagen low clusters are presented for significant GO terms shared by at least five cohorts.
+The table is available as a supplementary Excel file.'}
+
+suppl_paper_tbl$go_enrichment[1, 1] %>% 
+  mutate(`Collagen cluster` = '') %>% 
+  set_names(' ') %>% 
+  flextable %>% 
+  set_widths(16)
+
+```
+
+\newpage
+
+```{r tab-regulons, tab.cap = 'Activity of transcriptional regulons in the collagen high cluster as compared with the collagen low cluster predicted by the collecTRI model.
+Regulon activity was estimated with uni-parameter linear modeling with whole-transcriptome effect sizes of differential gene expression, p values were corrected for multiple testing with the false discovery rate (FDR) method.
+Linear model scores are presented for regulons significantly activated or inhibited in at least five cohorts.
+The table is available as a supplementary Excel file.'}
+
+suppl_paper_tbl$regulons[1, 1] %>% 
+  mutate(Cohort = '') %>% 
+  set_names(' ') %>% 
+  flextable %>% 
+  set_widths(16)
+
+```
+
+\newpage
+
+```{r tab-signaling, tab.cap = 'Activity of signaling pathways in the collagen high cluster as compared with the collagen low cluster predicted by the PROGENy model.
+Pathway activity was estimated with multi-parameter linear modeling with whole-transcriptome effect sizes of differential gene expression.
+P values were corrected for multiple testing with the false discovery rate (FDR) method,  linear model scores serve as measures of pathway activity.
 The table is available as a supplementary Excel file.'}
@@
 
-```{r tab-reactions, tab.cap = 'Biochemical reactions predicted to be significantly activated in the collagen high or collagen low cluster as compared with collagen low cancers.
-Statistical significance was determined by Monte Carlo simulation.
-P values were corrected for multiple testing with the false discovery rate method.
+\newpage
+
+```{r tab-reactions, tab.cap = 'Biochemical reactions predicted to be significantly activated in collagen high as compared with collagen low cancers.
+Statistical significance was determined by a Monte Carlo simulation.
+P values were corrected for multiple testing with the false discovery rate (FDR) method.
 The table is available as a supplementary Excel file.'}
 
-suppl_paper_tbl$reactions[1, 1] %>% 
+suppl_paper_tbl$metabolism[1, 1] %>% 
   mutate(Cohort = '') %>% 
@@
 
-```{r tab-subsystems, tab.cap = "Results of enrichment analysis for significantly activated and inhibited biochemical reaction within the Recon metabolism subsystem.
-Statistical significance was determined by Fisher's exact test corrected for multiple testing with the false discovery rate method (FDR).
-The table is available as a supplementary Excel file."}
+\newpage
+
+```{r tab-subsystems, tab.cap = 'Results of enrichment analysis for significantly activated and inhibited biochemical reactions within the Recon metabolism subsystem.
+Statistical significance was determined by random sampling from the entire reaction pool p values were corrected for multiple testing with the false discovery rate (FDR) method.
+Effect size of enrichment of the subsystem in significantly activated or inhibited reactions was measured by odds ratio (OR) statistic.
+The table is available as a supplementary Excel file.'}
 
@@
 
+```{r tab-coll-score, tab.cap = 'Transcriptomic collagen score was established by Elastic Net Cox modeling of biochemical relapse-free survival in the TCGA cohort.
+Hazard ratios (HR) of the member genes are presented.'}
+
+flextable::flextable(suppl_paper_tbl$coll_score) %>% 
+  set_widths(c(3, 3)) %>% 
+  italic(1) %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-coll-survival, tab.cap = 'Performance of the transcriptomic collagen score at prediction of biochemical relapse-free survival in the TCGA training cohort and validation collectives.'}
+
+flextable::flextable(suppl_paper_tbl$coll_survival) %>% 
+  set_widths(c(2.8, 3, 2.7, 2.7, 3, 3)) %>% 
+  merge_v(1) %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
 # Supplementary Figures
+
+```{r fig-norm-tumor, fig.width = figur::convert(suppl_paper_fig$norm_tumor, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$norm_tumor, to = 'in')$h, fig.cap = 'Expression of collagen-related genes in the normal prostate and prostate cancer tissue.'}
+
+suppl_paper_fig$norm_tumor$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-norm-tumor). Expression of collagen-related genes in the normal prostate and prostate cancer tissue.__
+
+_Differences in $log_2$-transformed expression of `r nrow(globals$genes_interest)` collagen-related genes between donor-matched pairs of the prostate cancer and benign tissue were assessed by paired T test with Cohen's effect size statistic in the GSE70768 and TCGA cohorts.
+P values were corrected for multiple testing with the false discovery rate (FDR) method.
+Full analysis results are listed in Supplementary Table S\@ref(tab:tab-norm-tumor)_
+
+_(A) Numbers of significantly up- and downregulated genes in the tumor tissue as compared with the benign tissue in the investigated cohorts presented in Venn plots._
+
+_(B) Mean normalized $log_2$ expression of the collagen-related genes presented as lines.
+Tinted regions represent the 2 $\times$ SEM (standard error of the mean) intervals.
+Significant effects are highlighted with bold font in the Y axis.
+Numbers of tissue pairs are displayed in the plot captions._
+
+\newpage
+
+```{r fig-clust-dev, fig.width = figur::convert(suppl_paper_fig$clust_dev, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$clust_dev, to = 'in')$h, fig.cap = 'Semi-supervised clustering of prostate cancer samples in respect to expression of collagen-related genes.'}
+
+suppl_paper_fig$clust_dev$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-clust-dev). Semi-supervised clustering of prostate cancer samples in respect to expression of collagen-related genes.__
+
+_Cancer samples in the TCGA training cohort (n = `r nrow(clust_globals$data$tcga)`) were clustered in respect to normalized, $log_2$-transformed expression levels of the collagen-related genes of interest with the PAM (partition around medoids) algorithm with cosine distance metric.
+Two clusters were defined: collagen low and collagen high.
+Assignment of the tumor samples in the training GSE54460 (n = `r nrow(clust_globals$data$gse54460)`), GSE70768 (n = `r nrow(clust_globals$data$gse70768)`), GSE70769 (n = `r nrow(clust_globals$data$gse70769)`), GSE220095 (n = `r nrow(clust_globals$data$gse220095)`), and DKFZ collective (n = `r nrow(clust_globals$data$dkfz)`) to the collagen clusters was accomplished by an inverse distance-weighted k-nearest neighbor classifier._
+
+_(A) Comparison of performance of several clustering algorithms (PAM: partition around medoids, HCL: hierarchical clustering/Ward D2 and KMEANS) and distance metrics (Euclidean, Manhattan, squared Euclidean and cosine) in 5-fold cross-validation of the TCGA data set gauged by mean silhouette width as a measure of cluster separation, fraction of observations with negative silhouette widths indicative of possible misclassification, explained clustering variance and fraction of misclassified 5-nearest neighbors, cluster assignment accuracy in the cross-validation folds.
+Note superior performance of the PAM/cosine algorithm used for definition of the collagen clusters._
+
+_(B) Choice of cluster number (k = 2) for PAM/cosine clustering of cancer samples in the TCGA cohort motivated by the peak mean silhouette width statistic._
+
+_(C) Distribution of sizes of the collagen clusters in the TCGA training cohort and the validation collectives. Percentages of samples assigned to the collagen clusters are presented in a stack plot._
+
+_(D) Quality of collagen clusters in the training cohort (TCGA) and validation collectives.
+Cluster separation was assessed by mean silhouette width, misclassification rate is expressed as fraction of observations with negative silhouette widths.
+Explanatory performance was gauged by fraction of explained clustering variance.
+Neighborhood preservation was assessed by mean fraction of 5-nearest neighbors placed in different clusters._
+
+\newpage
+
+```{r fig-clusters, fig.width = figur::convert(suppl_paper_fig$clusters, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$clusters, to = 'in')$h, fig.cap = 'Expression of the collagen-related genes in the collagen clusters of prostate cancers.'}
+
+suppl_paper_fig$clusters$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-clusters). Expression of the collagen-related genes in the collagen clusters of prostate cancers.__
+
+_The collagen clusters of prostate cancer samples were developed by semi-supervised PAM clustering as presented in Supplementary Figure S\@ref(fig:fig-clust-dev).
+Results of comparison of levels of the cluster-defining factors between the collagen clusters are listed in Supplementary Table S\@ref(tab:tab-cluster)._
+
+_(A) Uniform manifold approximation and projection (UMAP) layout of normalized $log_2$-transformed expression values of the collagen-related genes utilized for definition of the collagen clusters.
+Each point represents a single cancer sample, cluster assignment is color-coded.
+Numbers of cancer samples in the collagen clusters are indicated in the figure legends._
+
+_(B) Mean normalized $log_2$-transformed expression values of the cluster-defining collagen-related genes visualized in a heat map.
+The cluster-defining genes were classified by their biological function (Pro: proline metabolism, ECM: extracellular matrix, Adh.: adhesion)._
+
+\newpage
+
+```{r fig-clust-tumor, fig.width = figur::convert(suppl_paper_fig$clust_tumor, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$clust_tumor, to = 'in')$h, fig.cap = 'Gleason score and pathological tumor stage in the collagen clusters.'}
+
+suppl_paper_fig$clust_tumor$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-clust-tumor). Gleason score and pathological tumor stage in the collagen clusters.__
+
+_Differences in distribution of Gleason scores (A, 5 - 6, 7 and $\geq$ 8) and pathological tumor stages (B) between the collagen clusters were investigated by $\chi^2$ test with Cramer V effect size statistic.
+P values were corrected for multiple testing with the false discovery rate method.
+Percentages of samples assigned to the Gleason score strata or tumor stages within each cluster are shown in stack plots.
+Effect sizes and p values are displayed in the plot facets.
+Numbers of complete observations in the clusters are indicated in the Y axes.
+Full clinical and pathological characteristic of the collagen clusters is shown in Supplementary Table S\@ref(tab:tab-clinic)._
+
+\newpage
+
+```{r fig-intiltration, fig.width = figur::convert(suppl_paper_fig$infiltration, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$infiltration, to = 'in')$h, fig.cap = 'Cancer-associated fibroblasts and endothelial cell infiltration in the collagen clusters.'}
+
+suppl_paper_fig$infiltration$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-intiltration). Cancer-associated fibroblasts and endothelial cell infiltration in the collagen clusters.__
+
+_Levels of cancer-associated fibroblasts (A) and endothelial cells (B) in the collagen clusters were predicted by the MCP Counter and xCell algorithms.
+Statistical significance of differences between the clusters was assessed by Mann-Whitney test with r effect size statistic.
+P values were corrected for multiple testing with the false discovery rate method.
+Median infiltration levels with interquartile ranges are visualized as boxes, whiskers span over 150% of the interquartile ranges.
+Points represent single cancer samples.
+Effect sizes and p values are displayed in the Y axes, significant effects are highlighted in bold.
+GSE54460: `r format_strata_n(ana_globals$n_numbers$gse54460)`, GSE70768: `r format_strata_n(ana_globals$n_numbers$gse70768)`, GSE70769: `r format_strata_n(ana_globals$n_numbers$gse70769)`, GSE220095: `r format_strata_n(ana_globals$n_numbers$gse220095)`, DKFZ: `r format_strata_n(ana_globals$n_numbers$dkfz)`, TCGA: `r format_strata_n(ana_globals$n_numbers$tcga)`.
+Results of the analysis for other cell populations are listed in Supplementary Table S\@ref(tab:tab-infiltration)._
+
+\newpage
+
+```{r fig-biology, fig.width = figur::convert(suppl_paper_fig$biology, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$biology, to = 'in')$h, fig.cap = 'Gene set variation analysis of Reactome pathway gene signatures and gene ontology term enrichment analysis for the collagen clusters.'}
+
+suppl_paper_fig$biology$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-biology). Gene set variation analysis of Reactome pathway gene signatures and gene ontology term enrichment analysis for the collagen clusters.__
+
+_(A) Gene set variation analysis (GSVA) with Reactome pathway gene signatures.
+Statistical significance for differences in signature single sample gene set enrichment analysis scores (ssGSEA) between the collagen clusters was determined by two-tailed T test with Cohen's d effect size statistic.
+P values were corrected for multiple testing with the false discovery rate method.
+Mean ssGSEA scores in the clusters and cohorts for signatures significantly regulated in at least five cohorts with at least weak effect size (d $\geq$ 0.2) are presented in a heat map.
+Full analysis results are presented in Supplementary Table S\@ref(tab:tab-gsva).
+The common regulated gene signatures were classified in respect their co-expression patterns by KMEANS clustering; signature classification is indicated in the Y axis facets of the heat map (ECM: extracellular matrix, GF: growth factor, GPCR: G protein-coupled receptor)._
+
+_(B) Genes differentially expressed between the collagen clusters were identified by false discovery rate (FDR) corrected two-tailed T test with Cohen's d effect size statistic.
+Differentially expressed genes were defined by the pFDR < 0.05 and d $\geq$ 0.2 cutoffs (Supplementary Table \@ref(tab:tab-dge)).
+Biological process gene ontology (GO) enrichment within genes significantly unregulated in the collagen high and the collagen low clusters was investigated with the goana algorithm.
+Enrichment p values were adjusted for multiple testing with the FDR method, odds ratio of enrichment over the whole-transcriptome GO frequency served as an effect size statistic.
+GO terms significantly enriched with at least weak effect size (OR $\geq$ 1.44) in at least five cohorts were subjected to multi-dimensional scaling and unsupervised KMEANS clustering in respect to their semantic similarity.
+Multi-dimensional scaling layouts of Wang distances between the common enriched GO terms are presented.
+Each point represents a single GO term, GO cluster assignment is color coded.
+Complete results of the GO enrichment analysis are listed in Supplementary Table S\@ref(tab:tab-go-enrichment)._
+
+\newpage
+
+```{r fig-signaling, fig.width = figur::convert(suppl_paper_fig$signaling, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$signaling, to = 'in')$h, fig.cap = 'Transcriptional regulons and signaling pathway activity in the collagen clusters.'}
+
+suppl_paper_fig$signaling$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-signaling). Transcriptional regulons and signaling pathway activity in the collagen clusters.__
+
+_Differences in activity of transcriptional regulons (gene sets controlled by a common transcription factor, A) and signaling pathways (B) in the collagen high cluster as compared with the collagen low cluster were assessed by linear modeling with the collecTRI and PROGENy databases implemented by the decoupler algorithm.
+Linear model score (LM score) served as a measure of regulon or pathway activity, p values for significant modulation of activity (LM score $\neq$ 0) were corrected for multiple testing with the false discovery rate method.
+Full analysis results are listed in Supplementary Tables S\@ref(tab:tab-regulons) and S\@ref(tab:tab-signaling)._
+
+_(A) LM scores of top modulated transcriptional regulons found to be significantly activated or inhibited in at least five cohorts visualized as bubble plots.
+Point color codes for the regulation sign, point size codes for absolute value of the LM score.
+LM scores for regulon activity in particular cohorts are presented next to the data points._
+
+_(B) LM scores of signaling pathways found to be significantly activated or inhibited in at least five cohorts visualized as bubble plots.
+Point color codes for the regulation sign, point size codes for absolute value of the LM score.
+LM scores for regulon activity in particular cohorts are presented next to the data points._
+
+\newpage
+
+```{r fig-metabolism, fig.width = figur::convert(suppl_paper_fig$metabolism, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$metabolism, to = 'in')$h, fig.cap = 'Metabolism in the collagen clusters.'}
+
+suppl_paper_fig$metabolism$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-metabolism). Metabolism in the collagen clusters.__
+
+_Modulation of Recon2 model metabolic reactions in collagen high tumors as compared with collagen low cancers was predicted by Monte Carlo simulation implemented by the BiGGR and biggrExtra algorithms based on whole-genome differential gene expression estimates (Supplementary Table \@ref(tab:tab-reactions)).
+Enrichment of significantly activated and significantly inhibited reactions in the Recon metabolic subsystems was assessed by comparing the observed frequency of the subsystem's significantly regulated reactions with 10000 random draws from the entire reaction pool.
+Enrichment p values were corrected for multiple testing with the false discovery rate method.
+Enrichment odds ratios (OR) for significantly enriched metabolic subsystems with at least weak effect size (OR $\geq$ 1.44)  shared by at least five cohorts are presented in a bubble plot.
+Point color codes for reaction modulation status, point sizes represent absolute values of OR.
+Points are labeled with their OR values.
+Full analysis results are presented in Supplementary Table S\@ref(tab:tab-subsystems)._
 

### 2024-01-18 report/markdown/manuscript_figures_tables.Rmd
Figures, tables and methods for survival analysis

@@
 
-## Transcriptomic Collagen Score and prediction of biochemical relapse-free survival
+## Prediction of biochemical relapse-free survival by gradient boosted machine modeling
 
-Biochemical relapse-free survival was modeled in the TCGA training cohort as a function of first- and second-order normalized, $\log_2$-transformed expression levels of the collagen-related genes of interest by Elastic Net multi-parameter Cox regression [@Simon2011; @Zou2005].
-The Transcriptomic Collagen Score was computed as the linear predictor score of the Elastic Net Cox model with non-zero model coefficients presented in __Figure \@ref(fig:fig-coll-score)A__ and __Supplementary Table S15__.
+For survival analyses, $log_2$-transformed expression values of the collagen-related genes (__Supplementary Table S2__) were subjected to batch-adjustment with the ComBat algorithm [@Leek2012]. 
+Subsequently, the ComBat-adjusted expression data sets of the GSE54460, GSE70768, GSE70769, GSE220095 cohorts were merged (further referred to as 'pooled GEO' cohort).
 
-Performance of the Transcriptomic Collagen Score at prediction of biochemical relapse-free survival in the training collective (TCGA) and validation data sets (remaining cohorts) was assessed by univariable Cox regression with Harrell's concordance index [@Harrell1996] and integrated Brier score [@Graf1999] as metrics of goodness of fit and calibration (__Supplementary Table 16__).
+Biochemical relapse-free survival in the pooled GEO training cohort was modeled with normalized (Z-scores) expression values of the collagen-related genes as explanatory variables by the Gradient Boosted Machine (GBM) algorithm [@Friedman2001; @Greenwell2022; @Natekin2013; @Friedman2002].
+Selection of the optimal set of modeling parameters was accomplished by tuning with 10-fold cross-validation and out-of-fold model deviance as performance statistic (__Supplementary Tables S15__). 
+The Transcriptomic Collagen Score was computed as the linear predictor score of the GBM model. 
+Importance of the explanatory variables for the predictive performance of the GBM model were measured with the relative influence method [@Friedman2001] and expressed as difference in the sum of squared errors ($\Delta SSE$, __Figure \@ref(fig:fig-coll-score)A__).
+
+Performance of the Transcriptomic Collagen Score at prediction of biochemical relapse-free survival in the training pooled GEO collective and validation data sets (TCGA and DKFZ) was assessed by univariable Cox regression with Harrell's concordance index [@Harrell1996] and integrated Brier score [@Graf1999] as metrics of goodness of fit and calibration (__Figure \@ref(fig:fig-coll-score)B__, __Supplementary Table 16__).
+Of note, predictive performance of the GBM algorithm was excellent and reproducible in the investigated cohorts as compared with several other machine learning algorithms (e.g.
+Ridge Cox regression, Elastic Net or Support Vector Machines, __Supplementary Figure S9__).
+Differences in biochemical relapse-free survival in cancer patients stratified by tertiles of the Transcriptomic Collagen Scores were assessed by false discovery rate (FDR) corrected Peto-Peto test (__Figure \@ref(fig:fig-coll-score)C__).
 
@@
 
-```{r fig-coll-score, fig.width = figur::convert(paper_fig$coll_score, to = 'in')$w, fig.height = figur::convert(paper_fig$coll_score, to = 'in')$h, fig.cap = 'Transcriptional Collagen Score and prediction of biochemical relapse-free survival.'}
+```{r fig-coll-score, fig.width = figur::convert(paper_fig$coll_score, to = 'in')$w, fig.height = figur::convert(paper_fig$coll_score, to = 'in')$h, fig.cap = 'Transcriptional Collagen Model of biochemical relapse-free survival.'}
 
@@
 
-__Figure \@ref(fig:fig-coll-score). Transcriptional Collagen Score and prediction of biochemical relapse-free survival.__
+__Figure \@ref(fig:fig-coll-score). Transcriptional Collagen Model of biochemical relapse-free survival.__
 
-_Transcriptional Collagen Score was established in the TCGA training cohort by multi-parameter Elastic Net Cox modeling of biochemical relapse-free survival as a function of first- and second-order normalized $log_2$-transformed expression levels of `r nrow(globals$genes_interest)` collagen-related genes._
+_Transcriptional Collagen Score was established in the pooled GEO training cohort by multi-parameter Gradient Boosted Machine (GBM) modeling of biochemical relapse-free survival as a function of normalized $log_2$-transformed and ComBat-adjusted expression levels of `r nrow(globals$genes_interest)` collagen-related genes._
 
-_(A) Non-zero Elastic Net Cox model coefficients presented as hazard ratios (HR). Point color codes for marker status (blue: favorable, red: unfavorable), point size represents absolute value of HR._
+_(A) Importance of explanatory variables measured by the gradient of sum of squared errors ($\Delta SSE$) during the training process attributed to particular collagen-related transcript levels._
 

### 2024-01-18 report/markdown/manuscript_supplement.Rmd
Figures, tables and methods for survival analysis

@@
 
-For multi-parameter Elastic Net Cox modeling, the packages _glmnet_ [@Friedman2010], _survival_ [@Therneau2000], _surviminer_ [@Kassambara2016] and [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions) were used.
+Multi-parameter modeling of biochemical relapse-free survival was done with the packages 
+_glmnet_ [@Friedman2010] (Ridge, Elastic Net and LASSO cox regression), 
+_survivalsvm_ [@Fouodo2018] (Support Vector Machines [SVM]), 
+_rfsrc_ [@Ishwaran2008; @Ishwaran2022a] (Random Forest), 
+_gbm_ [@Greenwell2022] (Gradient Boosted Machines [GBM]), 
+_survival_ [@Therneau2000], 
+_surviminer_ [@Kassambara2016] and 
+[_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions) were used. 
+Univariable analysis of relapse-free survival was accomplished with [_kmOptimizer_](https://github.com/PiotrTymoszuk/kmOptimizer').
 
@@
 
-## Survival analysis and transcriptional collagen score
+## Multi-parameter survival modeling and transcriptional collagen score
 
-Biochemical relapse-free survival was modeled in the TCGA training cohort as a function of normalized $log_2$-transformed expression of the collagen-related genes of interest by multi-parameter Elastic Net Cox modeling [@Simon2011; @Zou2005].
-Of note, both first and second order terms were included in the Elastic Net model to account for possible non-linear association of gene expression with survival probability. 
-The optimal values of the shrinkage parameter $\lambda$ was found by tuning in 200 repeats 10-fold cross-validation and corresponded to the model with the lowest deviance (function `cv.glmnet()`, package _glmnet_) [@Friedman2010].
-The Elastic Net model was trained for the TCGA data set for the tuned $\lambda$ parameter with the `glmnet()` function from the _glmnet_ package. 
-The Transcriptional Collagen Score was defined as the linear predictor score of the Elastic Net model and was computed for the training TCGA cohort and the validation collectives GSE54460, GSE70768, GSE70769, GSE220095 and DKFZ with the `predict()` method.
-Gene expression values used for calculation of the Transcriptional Collagen Score, i.e.
-non-zero coefficients of the Elastic Net model, are presented as hazard ratios (HR) in __Figure 3A__ and __Supplementary Table S\@ref(tab:tab-coll-score)__.
+For survival analyses, $log_2$-transformed expression values of the collagen-related genes (__Supplementary Table S2__) were subjected to batch-adjustment with the ComBat algorithm [@Leek2012]. 
+Subsequently, the ComBat-adjusted expression data sets of the GSE54460, GSE70768, GSE70769, GSE220095 cohorts were merged (further referred to as 'pooled GEO' cohort).
+
+Biochemical relapse-free survival in the pooled GEO training cohort was modeled with normalized (Z-scores) expression values of the the collagen-related genes as explanatory variables by the GBM algorithm [@Friedman2001; @Greenwell2022; @Natekin2013; @Friedman2002].
+Selection of the optimal set of modeling parameters (number of trees, shrinkage factor, interaction depth, and minimal number of observations in the tree node) was accomplished by tuning with 10-fold cross-validation and out-of-fold model deviance as performance statistic (__Supplementary Table S\@ref(tab:tab-tuning)__).
+The tuning was performed with an in-house developed script. 
+The GBM model was trained in the pooled GEO cohort with the optimal tuning parameter set with the `gbm()` function (package _gbm_). 
+The Transcriptomic Collagen Score was defined as the linear predictor score of the GBM model and was computed for the best model iteration (function `gbm.perf(method = 'cv')`) with the `predict()` method and expression data from the training pooled GEO cohort and the TCGA and DKFZ test collectives.
+Importance of the explanatory variables for the predictive performance of the GBM model in the training pooled GEO cohort were measured with the relative influence method [@Friedman2001] and expressed as gradient of the sum of squared errors attributed to particular variables ($\Delta SSE$, __Figure 3A__).
 
@@
 The proportional hazard assumption of the Cox models was checked with the `summary(type = 'assumptions')` method, which implements the genuine `cox.zph()` algorithm [@Grambsch1994]. 
-Harrell's concordance index [@Harrell1996] and integrated Brier score [@Graf1999] were employed to investigate the overall concordance between the predicted and observed survival and to assess overall calibration (method `summary(type = 'fit')`, package [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions), __Figure 3B__ and __Supplementary Table S\@ref(tab:tab-coll-survival)__).
+Harrell's concordance index [@Harrell1996] and integrated Brier score [@Graf1999] were employed to investigate the overall concordance between the predicted and observed survival and to assess overall calibration (method `summary(type = 'fit')`, package [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions), __Figure 3B__ and __Supplementary Table S\@ref(tab:tab-model-survival)__).
+Of note, as compared with several other machine learning algorithms (Ridge, Elastic Net, and LASSO Cox regression, SVM and Random Forest) [@Friedman2010; @Fouodo2018; @Ishwaran2008; @Ishwaran2022a], the GBM model demonstrated an excellent performance and the highest reproducibility at prediction of biochemical relapse-free survival in the investigated cohorts (__Supplementary Figure S\@ref(fig:fig-surv-algos)__, __Supplementary Table S\@ref(tab:tab-model-survival)__).
 Statistical significance of differences in biochemical relapse-free survival between patients stratified by tertiles of the Transcriptional Collagen Score was determined by Peto-Peto test (functions `surv_fit()` and `surv_pvalue()`, package _survminer_, __Figure 3C__).
+
+## Survival analysis for single collagen-related transcripts
+
+The univariable analysis of biochemical relapse-free survival for single collagen-related genes was done in the pooled GEO, TCGA and DKFZ data sets described above for the multi-parameter survival modeling.
+For each of the collagen-related genes, cancer patients were classified as high and low expressors by an expression cutoff corresponding to the largest difference in biochemical relapse-free survival measured by FDR-corrected Mentel-Henszel test (function `find_cutoff()`, package [_kmOptimizer_](https://github.com/PiotrTymoszuk/kmOptimizer), minimal expression strata size set to 25% of the observations).
+Hazard ratio (HR) with 95% confidence intervals for relapse risk in high expressors as compared with low expressors was calculated with the canonical _survival_ package functions _coxph()_, _coef_ and _confint()_.
+Genes associated with unfavorable prognosis were defined by pFDR < 0.05 and HR > 1 for the high vs low expressor comparison. 
+Genes associated with favorable prognosis were defined by pFDR < 0.05 and HR < 1 (__Supplementary Figure S\@ref(fig:fig-surv-uni-volcano)__). 
+Common survival markers were defined as genes significantly associated with survival in all cohorts (__Supplementary Figure S\@ref(fig:fig-surv-uni-forest)__ - __S\@ref(fig:fig-surv-uni-detal3)__, __Supplementary Table S\@ref(tab:tab-rfs-univariable)__).
 
@@
 
-```{r tab-coll-score, tab.cap = 'Transcriptomic collagen score was established by Elastic Net Cox modeling of biochemical relapse-free survival in the TCGA cohort.
-Hazard ratios (HR) of the member genes are presented.'}
+```{r tab-tuning, tab.cap = 'Selection of the optimal parameters of machine learining survival models of biochemical relapse-free survival with expression of the collagen-related genes.
+The selection process was accomplished by cross-validation tuning in the pooled GEO cohort.'}
 
-flextable::flextable(suppl_paper_tbl$coll_score) %>% 
-  set_widths(c(3, 3)) %>% 
-  italic(1) %>% 
+flextable::flextable(suppl_paper_tbl$tuning) %>% 
+  set_widths(c(3.5, 4, 3, 3)) %>% 
+  merge_v(1:2, combine = TRUE) %>% 
+  footnote(1, 1, 
+           value = as_paragraph('SVM: Support Vector Machines; GBM: Gradient Boosted Machines.'),
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  footnote(1, 2, 
+           value = as_paragraph('CV: cross-validation.'), 
+           part = 'header', 
+           ref_symbols = 'b') %>% 
   theme_vanilla
@@
 
-```{r tab-coll-survival, tab.cap = 'Performance of the transcriptomic collagen score at prediction of biochemical relapse-free survival in the TCGA training cohort and validation collectives.'}
+```{r tab-model-survival, tab.cap = 'Performance of machine learning models at prediction of biochemical relapse-free survival with expression of the collagen-related genes.'}
 
-flextable::flextable(suppl_paper_tbl$coll_survival) %>% 
-  set_widths(c(2.8, 3, 2.7, 2.7, 3, 3)) %>% 
-  merge_v(1) %>% 
+flextable::flextable(suppl_paper_tbl$model_survival) %>% 
+  set_widths(c(3.5, 2.7, 3, 2.8, 2.8)) %>% 
+  merge_v(1:2) %>% 
+  footnote(1, 1, 
+           value = as_paragraph('SVM: Support Vector Machines; GBM: Gradient Boosted Machines.'), 
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  footnote(1, 3, 
+           value = as_paragraph('pooled GEO: GSE54460, GSE70768, GSE70769, and GSE220095'), 
+           part = 'header', 
+           ref_symbols = 'b') %>% 
   theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-rfs-univariable, tab.cap = 'Results of univariable analysis of biochemical relapse-free survival with expression of the collagen-related genes.
+Prostate cancer patients were stratified by expression cutoffs corresponding to the largest difference in survival assessed by Mentel-Henszel test.
+Genes found to be significantly associated with the survival in all analyzed cohorts (pooled GEO, TCGA and DKFZ) are presented.
+The table is available in a supplementary Excel file.'}
+
+suppl_paper_tbl$rfs_univariable[1, 1] %>% 
+  mutate( `Association with survival` = '') %>% 
+  set_names(' ') %>% 
+  flextable %>% 
+  set_widths(18)
 
@@
 P values were corrected for multiple testing with the false discovery rate (FDR) method.
-Full analysis results are listed in Supplementary Table S\@ref(tab:tab-norm-tumor)_
+Full analysis results are listed in Supplementary Table S\@ref(tab:tab-norm-tumor)._
 
@@
 
+```{r fig-surv-algos, fig.width = figur::convert(suppl_paper_fig$surv_algos, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$surv_algos, to = 'in')$h, fig.cap = 'Comparison of performance of various machine learning algorithms at prediction of biochemical relapse-free survival with expression of the collagen-related genes.'}
+
+suppl_paper_fig$surv_algos$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-surv-algos).
+Comparison of performance of various machine learning algorithms at prediction of biochemical relapse-free survival with expression of the collagen-related genes.__
+
+_Biochemical relapse-free survival was modeled with normalized, $log_2$-transformed and ComBat-adjusted expression levels of the collagen-related genes by various machine learning algorithms (Ridge Cox regression, Elastic Net Cox regression, LASSO Cox regression, Support Vector Machines [SVM], Random Forest, and Gradient Boosted Machines [GBM]).
+Their performance at prediction of biochemical relapse-free survival in the training pooled GEO cohort, and the test TCGA and DKFZ collectives was assessed by Harrell's concordance index (C-index) and integrated Brier scores (IBS).
+High C-index values indicate good concordance between the observed and predicted survival.
+Low values of IBS indicate good model calibration._
+_C-index and IBS values are presented in scatter plots.
+Each point represents a single algorithm.
+Values of C-index and IBS expected for a model predicting survival at random are visualized as dashed lines._
+
+\newpage
+
+```{r fig-surv-uni-volcano, fig.width = figur::convert(suppl_paper_fig$surv_uni_volcano, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$surv_uni_volcano, to = 'in')$h, fig.cap = 'Differences in biochemical relapse-free survival between high and low expressors of the collagen-related genes.'}
+
+suppl_paper_fig$surv_uni_volcano$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-surv-uni-volcano). Differences in biochemical relapse-free survival between high and low expressors of the collagen-related genes.__
+
+_Cancer patients were stratified by cutoffs of normalized expression for each collagen-related genes which corresponded to the largest differences in biochemical relapse-free survival assessed by false discovery rate (FDR) corrected Mentel-Henszel test into high and low expressors for particular genes.
+Relative risk of biochemical relapse in high expressors as compared with low expressors was modeled by Cox proportional hazard regression and expressed as hazard ratio (HR).
+Unfavorable survival markers were defined by pFDR < 0.05 and HR > 1.
+Favorable survival markers were defined by pFDR < 0.05 and HR < 1.
+HR and pFDR values are presented in Volcano plots.
+Each point represents a single collagen-related gene.
+Significance and sign of association with the relapse risk are color coded.
+Significant genes are labeled with their symbols.
+Dashed lines indicate the significance and regulation cutoffs._
+
+\newpage
+
+```{r fig-surv-uni-forest, fig.width = figur::convert(suppl_paper_fig$surv_uni_forest, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$surv_uni_forest, to = 'in')$h, fig.cap = 'Collagen-related transcriptional markers of biochemical relapse risk shared by all investigated cohorts.'}
+
+suppl_paper_fig$surv_uni_forest$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-surv-uni-forest). Collagen-related transcriptional markers of biochemical relapse risk shared by all investigated cohorts.__
+
+_Differences in biochemical relapse-free survival in high expressors versus low expressors of each collagen-related gene were compared by Cox modeling and false discovery rate (FDR) corrected Mentel-Henszel test as presented in Supplementary Figure S\@ref(fig:fig-surv-uni-volcano).
+Unfavorable survival markers were defined by pFDR < 0.05 and hazard ratio (HR) > 1.
+Favorable survival markers were defined by pFDR < 0.05 and HR < 1.
+HR with 95% confidence intervals (CI) for unfavorable and favorable markers shared by all investigated cohorts are presented in a Forest plot._
+
+\newpage
+
+```{r fig-surv-uni-detail1, fig.width = figur::convert(suppl_paper_fig$surv_uni_detail1, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$surv_uni_detail1, to = 'in')$h, fig.cap = 'Differences in biochemical relapse-free survival between high and low gene expressors: _COL1A1_, _COL3A1_, and _COL5A2_.'}
+
+suppl_paper_fig$surv_uni_detail1$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-surv-uni-detail1). Differences in biochemical relapse-free survival between high and low gene expressors: _COL1A1_, _COL3A1_, and _COL5A2_.__
+
+_Differences in biochemical relapse-free survival in high expressors versus low expressors of each collagen-related gene were compared by Cox modeling and false discovery rate (FDR) corrected Mentel-Henszel test as presented in Supplementary Figure S\@ref(fig:fig-surv-uni-volcano).
+Unfavorable survival markers were defined by pFDR < 0.05 and hazard ratio (HR) > 1.
+Differences in biochemical relapse-free survival for the COL1A1, COL3A1, and COL5A2 genes identified as unfavorable markers in all investigated cohorts are shown in Kaplan-Meier plots.
+Numbers of complete cases and relapses, expression cutoff values, and numbers of high and low expressors are displayed in the plot captions.
+HR with 95% confidence intervals and pFDR obtained in Mentel-Henszel test are indicated in the plots._
+
+\newpage
+
+```{r fig-surv-uni-detail2, fig.width = figur::convert(suppl_paper_fig$surv_uni_detail2, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$surv_uni_detail2, to = 'in')$h, fig.cap = 'Differences in biochemical relapse-free survival between high and low gene expressors: _COL11A1_, _COL11A2_, _LOXL1_, and _SERPINH1_.'}
+
+suppl_paper_fig$surv_uni_detail2$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-surv-uni-detail2). Differences in biochemical relapse-free survival between high and low gene expressors: _COL11A1_, _COL11A2_, _LOXL1_, and _SERPINH1_.__
+
+_Differences in biochemical relapse-free survival in high expressors versus low expressors of each collagen-related gene were compared by Cox modeling and false discovery rate (FDR) corrected Mentel-Henszel test as presented in Supplementary Figure S\@ref(fig:fig-surv-uni-volcano).
+Unfavorable survival markers were defined by pFDR < 0.05 and hazard ratio (HR) > 1.
+Differences in biochemical relapse-free survival for the COL11A1, COL11A2, LOXL1, and SERPINH1 genes identified as unfavorable markers in all investigated cohorts are shown in Kaplan-Meier plots.
+Numbers of complete cases and relapses, expression cutoff values, and numbers of high and low expressors are displayed in the plot captions.
+HR with 95% confidence intervals and pFDR obtained in Mentel-Henszel test are indicated in the plots._
+
+\newpage
+
+```{r fig-surv-uni-detal3, fig.width = figur::convert(suppl_paper_fig$surv_uni_detal3, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$surv_uni_detal3, to = 'in')$h, fig.cap = 'Differences in biochemical relapse-free survival between high and low gene expressors: _COL4A6_, _DST_, _LAMB3_, and _PCOLCE2_.'}
+
+suppl_paper_fig$surv_uni_detal3$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-surv-uni-detal3). Differences in biochemical relapse-free survival between high and low gene expressors: _COL4A6_, _DST_, _LAMB3_, and _PCOLCE2_.__
+
+_Differences in biochemical relapse-free survival in high expressors versus low expressors of each collagen-related gene were compared by Cox modeling and false discovery rate (FDR) corrected Mentel-Henszel test as presented in Supplementary Figure S\@ref(fig:fig-surv-uni-volcano).
+Unfavorable survival markers were defined by pFDR < 0.05 and hazard ratio (HR) > 1.
+Differences in biochemical relapse-free survival for the COL4A6, DST, LAMB3, and PCOLCE2 genes identified as favorable markers in all investigated cohorts are shown in Kaplan-Meier plots.
+Numbers of complete cases and relapses, expression cutoff values, and numbers of high and low expressors are displayed in the plot captions.
+HR with 95% confidence intervals and pFDR obtained in Mentel-Henszel test are indicated in the plots._
+
+\newpage
+
 # References

### 2024-01-23 report/markdown/manuscript_figures_tables.Rmd
Minor update of the figures and tables

@@
 ---
-title: "Prognostic and biological relevance of collagen biosynthesis pathway in prostate cancer"
+title: "Prognostic and biological relevance of collagen-related genes in prostate cancer"
 subtitle: "Methods, figures and tables for the transcriptome part"

### 2024-01-23 report/markdown/manuscript_supplement.Rmd
Minor update of the figures and tables

@@
 ---
-title: "Prognostic and biological relevance of collagen biosynthesis pathway in prostate cancer"
+title: "Prognostic and biological relevance of collagen-related genes in prostate cancer"
 subtitle: "Supplementary material, transcriptome part"

### 2024-01-25 report/markdown/manuscript_figures_tables.Rmd
ISUP grading in Figure 1

@@
 
-Differential gene expression in cancer samples stratified by Gleason scoring (5 - 6, 7 and $\geq$ 8) was assessed by FDR-corrected one-way ANOVA with $\eta^2$ effect size statistic. 
+Differential gene expression in cancer samples stratified by the ISUP risk strata (ISUP1: Gleason score 5 - 6, ISUP2: Gleason score 7 and ISUP3: Gleason score $\geq$ 8) was assessed by FDR-corrected one-way ANOVA with $\eta^2$ effect size statistic.
 Genes with pFDR < 0.05 and $\eta^2 \geq$ 0.02 were deemed differentially regulated (__Supplementary Table S4__).
@@
 
-_(C) Differences in $log_2$-transformed expression levels of `r nrow(globals$genes_interest)` collagen-related genes between cancer samples stratified by Gleason scores were assessed by one-way ANOVA with $\eta^2$ effect size statistic in the GSE54460 (`r format_strata_n(gs_uni$n_numbers$gse54460)`), GSE70768 (`r format_strata_n(gs_uni$n_numbers$gse70768)`), GSE70769 (`r format_strata_n(gs_uni$n_numbers$gse70769)`), GSE220095 (`r format_strata_n(gs_uni$n_numbers$gse220095)`), DKFZ (`r format_strata_n(gs_uni$n_numbers$dkfz)`), and TCGA cohort (`r format_strata_n(gs_uni$n_numbers$tcga)`).
+_(C) Differences in $log_2$-transformed expression levels of `r nrow(globals$genes_interest)` collagen-related genes between cancer samples stratified according to the ISUP risk system were assessed by one-way ANOVA with $\eta^2$ effect size statistic in the GSE54460 (`r format_strata_n(gs_uni$n_numbers$gse54460)`), GSE70768 (`r format_strata_n(gs_uni$n_numbers$gse70768)`), GSE70769 (`r format_strata_n(gs_uni$n_numbers$gse70769)`), GSE220095 (`r format_strata_n(gs_uni$n_numbers$gse220095)`), DKFZ (`r format_strata_n(gs_uni$n_numbers$dkfz)`), and TCGA cohort (`r format_strata_n(gs_uni$n_numbers$tcga)`).
 Mean normalized $log_2$-transformed expression levels of genes significantly regulated between the Gleason score strata with at least weak effect size (pFDR < 0.05, $\eta^2 \geq$ 0.02) are presented in a heat map._

### 2024-01-25 report/markdown/manuscript_supplement.Rmd
ISUP grading in Figure 1

@@
 
-## Comparison of gene expression between the benign and cancer tissue, and cancer samples stratified by Gleason scores
+## Comparison of gene expression between the benign and cancer tissue, and cancer samples stratified by the ISUP risk scheme
 
 Differences in $log_2$-transformed expression levels of the collagen pathway genes between donor-matched cancer and benign prostate specimens were investigated with paired two-tailed T test with Cohen's effect size statistic in the GSE70768 and TCGA cohorts (__Figure 1B__, __Supplementary Figure S\@ref(fig:fig-norm-tumor)__ and __Supplementary Table S\@ref(tab:tab-norm-tumor)__).
-Differences in gene expression between cancer samples stratified by Gleason score (5 - 6, 7, $\geq$ 8) were explored by one-way ANOVA with $\eta^2$ effect size statistic (__Figure 1C__, __Supplementary Table S\@ref(tab:tab-gleason)__).
+Differences in gene expression between cancer samples stratified according to the ISUP risk scheme (ISUP1: Gleason Score 5 - 6, ISUP2: Gleason score 7, ISUP3+: Gleason Score $\geq$ 8) were explored by one-way ANOVA with $\eta^2$ effect size statistic (__Figure 1C__, __Supplementary Table S\@ref(tab:tab-gleason)__).
 Differentially regulated genes were defined by pFDR < 0.05 and at least weak effect size of the difference (d $\geq$ 0.2 or $\eta^2 \geq$ 0.02). 
@@
 
-Biochemical relapse-free survival in the pooled GEO training cohort was modeled with normalized (Z-scores) expression values of the the collagen-related genes as explanatory variables by the GBM algorithm [@Friedman2001; @Greenwell2022; @Natekin2013; @Friedman2002].
+Biochemical relapse-free survival in the pooled GEO training cohort was modeled with normalized (Z-scores) expression values of the collagen-related genes as explanatory variables by the GBM algorithm [@Friedman2001; @Greenwell2022; @Natekin2013; @Friedman2002].
 Selection of the optimal set of modeling parameters (number of trees, shrinkage factor, interaction depth, and minimal number of observations in the tree node) was accomplished by tuning with 10-fold cross-validation and out-of-fold model deviance as performance statistic (__Supplementary Table S\@ref(tab:tab-tuning)__).
@@
 
-```{r tab-gleason, tab.cap = 'Expression of the collagen pathway genes in cancer samples stratified by Gleason scores compared by one-way ANOVA with eta-square effect size statistic.
+```{r tab-gleason, tab.cap = 'Expression of the collagen pathway genes in cancer samples stratified by the ISUP risk groups compared by one-way ANOVA with eta-square effect size statistic.
 P values were corrected for multiple testing with the false discovery rate method.

### 2024-01-31 report/markdown/manuscript_supplement.Rmd
Confusion matrix heat maps, PIRADS3

@@
 
-Transcriptome cohorts for analysis in the current report were selected from studies deposited at cBioportal and Gene Expression Omnibus (GEO) with the following criteria: availability of information on Gleason scoring, biochemical relapse and biochemical relapse-free survival, and expression data for `r nrow(globals$genes_interest)` collagen-related genes investigated also in the proteomic part of the project (__Supplementary Table S\@ref(tab:tab-cohorts)__ and __S\@ref(tab:tab-genes)__).
+Transcriptome cohorts for analysis in the current report were selected from studies deposited at cBioportal and Gene Expression Omnibus (GEO) with the following criteria: availability of information on Gleason scoring or ISUP risk strata (International Society of Urological Pathology), biochemical relapse and biochemical relapse-free survival, and expression data for `r nrow(globals$genes_interest)` collagen-related genes investigated also in the proteomic part of the project (__Supplementary Table S\@ref(tab:tab-cohorts)__ and __S\@ref(tab:tab-genes)__).
 
@@
 ssGSEA scores were computed with the `calculate()` function from the [_gseaTools_](https://github.com/PiotrTymoszuk/gseaTools) package. 
+ISUP strata were defined based on Gleason Score information as follows: ISUP1: Gleason score 5 - 6, ISUP2: Gleason score 7, ISUP3+: Gleason score $\geq$ 8.
 
@@
   footnote(1, 2, 
-           value = as_paragraph('PSA: prostate-specific antigen.'), 
+           value = as_paragraph('PSA: prostate-specific antigen; ISUP: International Society of Urological Pathology risk stratification system.'), 
            part = 'header', 
@@
 
-```{r fig-clust-tumor, fig.width = figur::convert(suppl_paper_fig$clust_tumor, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$clust_tumor, to = 'in')$h, fig.cap = 'Gleason score and pathological tumor stage in the collagen clusters.'}
+```{r fig-clust-tumor, fig.width = figur::convert(suppl_paper_fig$clust_tumor, to = 'in')$w, fig.height = figur::convert(suppl_paper_fig$clust_tumor, to = 'in')$h, fig.cap = 'ISUP and pathological tumor stage in the collagen clusters.'}
 
@@
 
-__Supplementary Figure S\@ref(fig:fig-clust-tumor). Gleason score and pathological tumor stage in the collagen clusters.__
+__Supplementary Figure S\@ref(fig:fig-clust-tumor). ISUP and pathological tumor stage in the collagen clusters.__
 
-_Differences in distribution of Gleason scores (A, 5 - 6, 7 and $\geq$ 8) and pathological tumor stages (B) between the collagen clusters were investigated by $\chi^2$ test with Cramer V effect size statistic.
+_Differences in distribution of ISUP risk strata (International Society of Urological Pathology: ISUP1, ISUP2 and ISUP3+) and pathological tumor stages (B) between the collagen clusters were investigated by $\chi^2$ test with Cramer V effect size statistic.
 P values were corrected for multiple testing with the false discovery rate method.
-Percentages of samples assigned to the Gleason score strata or tumor stages within each cluster are shown in stack plots.
+Percentages of samples assigned to the ISUP strata or tumor stages within each cluster are shown in stack plots.
 Effect sizes and p values are displayed in the plot facets.

### 2024-02-03 report scripts/paper_tables.R
95% CI for C-indexes in survival models

@@ -522,14 +522,27 @@
    
 # Table S16: performance of the transcriptomic models at prediction of RFS --------
   
-  insert_msg('Table S16: RS prediction by the transcriptomic models')
+  insert_msg('Table S16: RFS prediction by the transcriptomic models')
   
   suppl_paper_tbl$model_survival <- surv_summary$stats %>%
+    map(function(x) if('lower_ci' %in% names(x)) {
+      
+      mutate(x, 
+             c_index = paste0(signif(c_index, 2), ' [95% CI: ', 
+                              signif(lower_ci, 2), ' to ', 
+                              signif(upper_ci, 2), ']'))
+      
+    } else {
+      
+      mutate(x, c_index = as.character(signif(c_index, 2)))
+      
+    }) %>% 
+    map(select, dataset, cohort, c_index, ibs_model) %>% 
     compress(names_to = 'algorithm') %>% 
-    select(algorithm, dataset, cohort, c_index, ibs_model) %>% 
+    relocate(algorithm) %>% 
     mutate(cohort = surv_globals$study_labels[cohort], 
            algorithm = surv_globals$algo_labels[algorithm]) %>% 
-    map_dfc(function(x) if(is.numeric(x)) signif(x, 3) else x) %>% 
+    map_dfc(function(x) if(is.numeric(x)) signif(x, 2) else x) %>% 
     set_names(c('Algorithm', 
                 'Data set type', 
                 'Cohort', 

### 2024-02-03 survival scripts/summary.R

@@ -23,7 +23,9 @@
                              gbm = gbm_surv) %>% 
     map(~.x$stats) %>% 
     map(select, 
-        dataset, cohort, c_index, ibs_model, ibs_reference)
+        dataset, cohort, 
+        c_index, any_of(c('lower_ci', 'upper_ci')), 
+        ibs_model, ibs_reference)
   
   ## variable importance
   

### 2024-02-03 report/markdown/manuscript_supplement.Rmd
95% CI included in Table S16

@@
 flextable::flextable(suppl_paper_tbl$model_survival) %>% 
-  set_widths(c(3.5, 2.7, 3, 2.8, 2.8)) %>% 
+  set_widths(c(2.8, 2.7, 2.7, 4.6, 2.5)) %>% 
   merge_v(1:2) %>% 

### 2024-02-10 report/markdown/manuscript_supplement.Rmd
Corrections, Figure S4, Table S6
Correct ISUP labeling

@@
 
-## Multi-parameter survival modeling and transcriptional collagen score
+## Multi-parameter survival modeling with transcriptome data
 
@@
 
-Performance of the Transcriptional Collagen Score at prediction of biochemical relapse-free survival was assessed by uni-variable Cox regression (functions `coxph()` and `as_coxex()`, packages _survival_ and [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions)) [@Therneau2000].
+Performance of the Transcriptomic Collagen Score at prediction of biochemical relapse-free survival was assessed by uni-variable Cox regression (functions `coxph()` and `as_coxex()`, packages _survival_ and [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions)) [@Therneau2000].
 The proportional hazard assumption of the Cox models was checked with the `summary(type = 'assumptions')` method, which implements the genuine `cox.zph()` algorithm [@Grambsch1994]. 
@@
 _(A) Comparison of performance of several clustering algorithms (PAM: partition around medoids, HCL: hierarchical clustering/Ward D2 and KMEANS) and distance metrics (Euclidean, Manhattan, squared Euclidean and cosine) in 5-fold cross-validation of the TCGA data set gauged by mean silhouette width as a measure of cluster separation, fraction of observations with negative silhouette widths indicative of possible misclassification, explained clustering variance and fraction of misclassified 5-nearest neighbors, cluster assignment accuracy in the cross-validation folds.
-Note superior performance of the PAM/cosine algorithm used for definition of the collagen clusters._
+Note the superior performance of the PAM/cosine algorithm used for definition of the collagen clusters._
 
@@
 Explanatory performance was gauged by fraction of explained clustering variance.
-Neighborhood preservation was assessed by mean fraction of 5-nearest neighbors placed in different clusters._
+Neighborhood preservation was assessed by mean fraction of 5-nearest neighbors placed in the same cluster._
 
@@
 Differentially expressed genes were defined by the pFDR < 0.05 and d $\geq$ 0.2 cutoffs (Supplementary Table \@ref(tab:tab-dge)).
-Biological process gene ontology (GO) enrichment within genes significantly unregulated in the collagen high and the collagen low clusters was investigated with the goana algorithm.
+Biological process gene ontology (GO) term enrichment within genes significantly unregulated in the collagen high and the collagen low clusters was investigated with the goana algorithm.
 Enrichment p values were adjusted for multiple testing with the FDR method, odds ratio of enrichment over the whole-transcriptome GO frequency served as an effect size statistic.
